---
title: "Slicejam Evaluation Workflow"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
    df_print: kable
    self_contained: false
---
   
About this document:

* prepared on **`r format(Sys.time(), "%a %b %d %X %Y")`**
* using **`r R.version.string`**
* R slicejam package version **`r packageVersion("slicejam")`**.

### R environment setup

This document can be reproduced using the R code displayed in
this document. The initial steps involve loading R packages
into memory.

#### Rmarkdown options

```{r, setup, include=TRUE}
#    css: liao_style.css
## To render this page:
## In the farrisSeq folder run:
## rmarkdown::render("Adam19_Rnaseq_Analysis_vM21.Rmd", "html_document", knit_root_dir=getwd(), output_dir=getwd());
##    css: /Users/wardjm/Projects/R-scripts/liao_style.css

## Define the cache so it is dependent upon:
## - the versions of a few R packages,
## - the environmental variable FC_FILE
knitr::opts_chunk$set(cache=TRUE,
   autodep=TRUE,
   dev=c("png", "cairo_pdf"),
   dev.args=list(
      pointsize=14
   ),
   fig.align="center",
   pdf.options(encoding="ISOLatin9.enc"),
   cache.path=Sys.getenv("CACHEDIR", "cache"),
   cache.extra=list(
      packageVersion("jamba"),
      packageVersion("jamma"),
      packageVersion("splicejam"),
      packageVersion("slicejam"),
      paste0("FC_FILE:", Sys.getenv("FC_FILE")),
      paste0("CURATION_TXT:", Sys.getenv("CURATION_TXT")),
      paste0("GTF:", Sys.getenv("GTF"))
   )
);

options("kable_styling_position"="left",
   "kable_styling_full_width"=FALSE,
   "kable_styling_bootstrap_options"=c("striped",
      "hover",
      "condensed",
      "responsive")
);
## Set font to Arial which helps compatibility in Adobe Illustrator,
## and helps display of up and down arrow characters in Venn diagrams
par("family"="Arial",
   "lend"="square",
   "ljoin"="mitre");
```

#### SliceJam analysis parameters

```{r, slicejam_args, include=TRUE, cache=FALSE}
sliceargs <- slicejam::get_slicejam_args(verbose=TRUE,
   name="eval",
   type="data.frame");
if ("sliceargs_df" %in% names(attributes(sliceargs))) {
   sliceargs_df <- attr(sliceargs, "sliceargs_df");
   kable_coloring(
      sliceargs_df,
      caption=paste0(
         "slicejam arguments:"));
}

DRYRUN <- sliceargs$DRYRUN;
ATAC <- sliceargs$ATAC;
fc_file <- sliceargs$fc_file;
fc_filepath <- sliceargs$fc_filepath;
CURATION_TXT <- sliceargs$CURATION_TXT;
GTF <- sliceargs$GTF;
GTFNAME <- sliceargs$GTFNAME;
DRYRUN <- sliceargs$DRYRUN;
OUTDIR <- sliceargs$OUTDIR;
CACHEDIR <- sliceargs$CACHEDIR;
ATAC <- sliceargs$ATAC;
MGM <- sliceargs$MGM;
NORM <- sliceargs$NORM;
NORMMIN <- sliceargs$NORMMIN;
NORMSHORT <- sliceargs$NORMSHORT;
GROUPCHECK <- sliceargs$GROUPCHECK;
SAVE_RDATA <- sliceargs$SAVE_RDATA;

detectedTx <- sliceargs$DETECTED_TX;
detectedGenes <- sliceargs$DETECTED_GENES;

upstream_promoter <- sliceargs$UPSTREAM_PROMOTER;
downstream_promoter <- sliceargs$DOWNSTREAM_PROMOTER;
upstream_tts <- sliceargs$UPSTREAM_TTS;
downstream_tts <- sliceargs$DOWNSTREAM_TTS;

fc_base <- sliceargs$fc_base;
fc_basedir <- sliceargs$fc_basedir;
fc_html <- sliceargs$fc_html;
knit_root_dir <- sliceargs$knit_root_dir;

```

#### R package initialization

```{r, cbiorutils, cache=TRUE}
# Load custom CBioRUtils that will be removed
# once functions are ported into Jam packages
suppressPackageStartupMessages(library(jamba));
if (!"CBioRUtils" %in% search()) {
   cbioPath <- file.path(
      dirname(Sys.getenv("HOME")),
      "wardjm/Builds/NIEHS_IntegrativeBIx-svn/Bpipe/bpipeScripts/CBioRUtils.R"
   );
   CBioRUtils <- new.env(parent=.GlobalEnv);
   suppressMessages(
      sys.source(cbioPath,
         envir=CBioRUtils,
         keep.source=TRUE));
   attach(CBioRUtils);
} else {
   ## Refresh CBioRUtils.R
   refreshCBioRUtils();
}
```

```{r, r_initialization, cache=FALSE}
## Prepend a common R package install location to .libPaths()
#.libPaths(
#   file.path(dirname(Sys.getenv("HOME")),
#      "wardjm/R/x86_64-pc-linux-gnu-library/3.5")
#);
knitr::kable(
   data.frame(check.names=FALSE,
      stringsAsFactors=FALSE,
      `.libPaths()`=.libPaths()),
   caption=paste0(
      "R library paths in use."));

suppressPackageStartupMessages(library(SummarizedExperiment));
suppressPackageStartupMessages(library(colorjam));
suppressPackageStartupMessages(library(jamma));
suppressPackageStartupMessages(library(splicejam));
suppressPackageStartupMessages(library(slicejam));
suppressPackageStartupMessages(library(data.table));
suppressPackageStartupMessages(library(matrixStats));
suppressPackageStartupMessages(library(extrafont));
#suppressPackageStartupMessages(library(GenomicFeatures));
suppressPackageStartupMessages(library(rtracklayer));
suppressPackageStartupMessages(library(methods));
suppressPackageStartupMessages(library(multienrichjam));
suppressPackageStartupMessages(library(kableExtra));

```

## Description of workflow

* Import the first `.fc` file found in the current working directory.

    * Optionally use "FC_FILE" environmental variable.

* Curate filename columns by one of two processes:

    * A file defined by `"CURATION_TXT"` is used to parse the colnames
    in the featureCounts file into three columns: `"Sample"`, `"Group"`, `"Run"`

* An R object `SummarizedExperiment` is created, using the columns
from featureCounts to populate rows, such as
`"Geneid", "Chr", "Start", "End", "Strand", "Length"`.
* Assign categorical colors to the unique values in the `"Group"` column.

* Normalization and batch adjustment

    * Normalize peak counts across all samples using the method defined by `NORM`.
    * When `"Run"` contains more than one unique value,
    apply batch adjustment using the `"Run"` column to define batch,
    and `"Group"` to define groups, using normalized data.

* Create MA-plots:

    * un-normalized data, global-centered
    * un-normalized data, centered by "Run"
    * normalized data, global-centered
    * normalized data, centered by "Run"
    * normalized data, centered by "Group","Run"
    * normalizeddata, global-centered, batch-adjusted by "Run"

* Create sample correlation heatmaps:

   * normalized data, centered by the first factor level
   in "Group", intended to act as a control group.
   * normalized data, batch-adjusted by "Run", global-centered across
   all samples.
   * normalized data, batch-adjusted by "Run", centered by the
   first factor level in "Group".

* Plot peak width by peak counts


## Evaluation workflow


### Import .fc file

Currently this workflow will use a `".fc"` file defined
via the environmental variable `"FC_FILE"`, or if that
variable is not defined, it searches for the first `".fc"`
file in the current working directory.

#### Define sample groups

```{r, import_fc}
## Import
fc_data <- import_featurecounts(fc_file);

## Add zero-padding to peak labels
fc_data[,1] <- paste0("peak_",
   jamba::padInteger(seq_len(nrow(fc_data))));
rownames(fc_data) <- fc_data[,1];

# featureCounts colnames
fc_colnames <- attr(fc_data, "fc_colnames");
filenames <- attr(fc_data, "filenames");
if (length(fc_colnames) == 0) {
   stop(paste0("None of the featureCounts colnames were detected: ",
      jamba::cPaste(fc_colnames)));
}


## Use curation_txt if that file exists
curation_txt <- CURATION_TXT;
curation_df <- data.table::fread(curation_txt,
   data.table=FALSE,
   sep="\t");
filename_colname <- "Filename";
sample_df <- slicejam::curate_to_df_by_pattern(
   x=filenames,
   input_colname=filename_colname,
   df=curation_df);
if (nrow(sample_df) == 0) {
   jamba::printDebug("Error: ",
      c("The curation_txt file '",
         curation_txt,
         "' did not match filenames."),
      sep="",
      fgText=c("red", "dodgerblue"));
   jamba::printDebug("filenames: ",
      paste0("\n'", filenames, "'"));
   jamba::printDebug("curation_txt contents:");
   print(curation_df);
   stop("curation_txt file did not match filenames.");
}

## Check for samples not curated, and therefore dropped from the analysis
if (length(sample_df[[filename_colname]]) < length(filenames)) {
   samples_dropped <- setdiff(filenames,
      sample_df[[filename_colname]]);
   samples_retained <- sample_df[[filename_colname]];
   samples_retmax <- max(c(
      length(samples_dropped),
      length(samples_retained)));
   samples_dropped <- unlist(rep(list(samples_dropped, ""),
      c(1, samples_retmax - length(samples_dropped))));
   samples_retained <- unlist(rep(list(samples_retained, ""),
      c(1, samples_retmax - length(samples_retained))));
   
   knitr::kable(
      data.frame(
         samples_retained=samples_retained,
         samples_dropped=samples_dropped),
      caption=paste0(
         "Samples retained and dropped."));
}

sample_colname <- head(jamba::provigrep(
   c("^label$",
      "^sample$",
      "^sample.*(name|id)$"),
   colnames(sample_df)), 1);
if (length(sample_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Label","' nor '", "Sample", "' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Sample column was recognized.");
}
sample_df[[sample_colname]] <- jamba::makeNames(sample_df[[sample_colname]]);
rownames(sample_df) <- sample_df[[sample_colname]];

run_colname <- head(jamba::provigrep(
   c("^batch$",
      "^run$"),
   colnames(sample_df)), 1);
if (length(run_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Run","' or '","Batch","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
}

group_colname <- head(jamba::provigrep(
   c("^Group$",
      "^group.*name$",
      "group$"),
   colnames(sample_df)), 1);
if (length(group_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Group","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Group column was recognized.");
}
## Ensure group names have no invalid characters like "-"
sample_df[[group_colname]] <- gsub("[%]", "pct",
   gsub("[- ]+",
      ".",
      sample_df[[group_colname]]));
sample_df[[group_colname]] <- factor(sample_df[[group_colname]],
   levels=unique(sample_df[[group_colname]]));

normgroup_colname <- head(jamba::provigrep(
   c("^normgroup$",
      "norm.*group"),
   colnames(sample_df)), 1);
if (length(normgroup_colname) == 0) {
   normgroup_colname <- "normgroup";
   sample_df[[normgroup_colname]] <- sample_df[[group_colname]];
}
   
## Sort sample_df by group
sample_df <- jamba::mixedSortDF(sample_df,
   byCol=c(group_colname,
      sample_colname));

## Define useful sample annotation vectors
iSamples <- sample_df[[sample_colname]];
Run <- jamba::nameVector(sample_df[[run_colname]],
   iSamples);
iGroup <- jamba::nameVector(sample_df[[group_colname]],
   iSamples);
iNormGroup <- jamba::nameVector(sample_df[[normgroup_colname]],
   iSamples);

## Make sure the order in filenames is propagated
filenames <- jamba::nameVector(sample_df[[filename_colname]],
   iSamples);
fc_data <- fc_data[,c(fc_colnames, filenames),drop=FALSE];

## Print sample annotation table
kable_coloring(sample_df,
   caption=paste0(
      "Sample Groups defined by curation_txt: '",
      basename(curation_txt),
      "'"));
if (any(!sapply(c("iSamples", "Run", "iGroup", "iNormGroup"), exists))) {
   stop("One or more of iSamples,Run,Group could not be defined.");
}

## Rename featureCounts columns to match iSamples
knitr::kable(
   data.frame(
      Rename_from=filenames,
      Rename_to=iSamples),
   caption=paste0(
      "Renamed featureCounts colnames."));
fc_data <- jamba::renameColumn(fc_data,
   from=filenames,
   to=iSamples);

## Make SummarizedExperiment
fc_se <- SummarizedExperiment(
   assays=list(counts=log2(1+as.matrix(fc_data[,iSamples,drop=FALSE]))),
   rowData=fc_data[,fc_colnames,drop=FALSE],
   colData=sample_df);

## Re-order samples by Group
iSamples <- mixedSortDF(colData(fc_se),
   byCols=group_colname)[[sample_colname]];
fc_se <- fc_se[,iSamples];
Run <- nameVector(colData(fc_se)[[run_colname]],
   colnames(fc_se));
iGroup <- nameVector(colData(fc_se)[[group_colname]],
   colnames(fc_se));

## Define sample colors
g_levels <- levels(colData(fc_se)[[group_colname]]);
n_levels <- length(g_levels);
group_colors <- jamba::nameVector(
   head(colorjam::rainbowJam(n_levels), n_levels),
   g_levels);
color_sub1 <- nameVector(group_colors[colData(fc_se)[[group_colname]]],
   colData(fc_se)[[sample_colname]]);
run_colors <- group2colors(unique(Run));

run_colors1 <- nameVector(run_colors[colData(fc_se[,iSamples])[[run_colname]]],
   iSamples);
```

#### Define statistical contrasts

```{r, define_contrasts}
sample_txt_file <- file.path(OUTDIR,
   "curated_samples.txt");
data.table::fwrite(
   x=sample_df,
   file=sample_txt_file,
   sep="\t",
   col.names=TRUE);

contrast_txt_file <- file.path(OUTDIR,
   "contrasts.txt");
if (file.exists(contrast_txt_file)) {
   contrastNames <- gsub("^[ \t\r\n]+|[ \t\r\n]+$",
      "",
      readLines(contrast_txt_file));
   jamba::printDebug("Loaded contrastNames from pre-existing file: '",
      contrast_txt_file,
      "'");
} else {
   contrastNames <- splicejam::groups2contrasts(
      sample_df[,group_colname,drop=FALSE],
      maxDepth=1)$contrastName;
   if (length(contrastNames) > 0) {
      writeLines(text=contrastNames,
         con=contrast_txt_file);
   } else {
      jamba::printDebug("Note there are no contrasts defined.",
         fgText="red");
   }
}

## Print sample annotation table
if (length(contrastNames) > 0) {
   contrasts_df <- data.frame(Contrasts=contrastNames);
   kable_coloring(contrasts_df,
      caption=paste0(
         "Contrasts stored in contrast_txt_file: '",
         basename(contrast_txt_file),
         "'"));
}

## Optional GROUPCHECK which exits here to allow reviewing the sample grouping
if (GROUPCHECK) {
   #txt_ext <- paste0(".curated_samples.txt");
   #sample_txt_file <- file.path(OUTDIR,
   #   gsub("[.]fc$|[.]fc[.]txt$",
   #      txt_ext,
   #      basename(fc_file)));
   stop(paste0("Curated sample data for review: '",
      sample_txt_file,
      "' and contrastNames file '",
      contrast_txt_file,
      "'"));
}

```

### Plot distribution of peak sizes

If the featureCounts output included column name `"Length"`,
the plots below will show the distribution of lengths.

```{r, peak_sizes, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes",
      xlab="Peak width",
      breaks=100);
   
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes (log-scales x-axis",
      log="x",
      xlab="Peak width",
      breaks=100);
}
```

### Define Max Group Mean (MGM) Threshold

The maxGroupMean threshold is defined by the environmental
variable `MGM`, used to customize the threshold during the stats
tests, applied to signal log2-transformed with `log2(1 + x)`,
and therefore the default `MGM=4` defines a threshold of `15 counts`
in normal space, using `( 2^MGM ) - 1` to reverse the
transform: `log2(1 + x)`


```{r, mgm_arg}
## Get maxGroupMean threshold argument
MGM <- Sys.getenv("MGM", 4);
MGM <- jamba::rmNA(as.numeric(unlist(strsplit(MGM, "[ ,]+"))));
if (length(MGM) == 0) {
   MGM <- c(4);
}
if (length(MGM) > 1) {
   jamba::printDebug("maxGroupMean thresholds: ", MGM);
} else {
   jamba::printDebug("maxGroupMean threshold: ", MGM);
}
```

The max group mean threshold is: **`r print(MGM)`**


### Apply Normalization

The normalization method defined is **`r print(NORM)`** normalization.

```{r, norm_quantile, dependson="import_fc"}
## Raw data
iMatrix <- assays(fc_se[,iSamples])[["counts"]];

## Perform normalization
if ("median" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("mediangroup" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            centerGroups=iNormGroup,
            controlSamples=iSamples,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("quantile" %in% NORM) {
   ## Evaluate quantile normalization
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("quantile"));
   normname <- "quantile_counts";
} else if ("none" %in% NORM) {
   normname <- "counts";
}

## Normalized data
iMatrixQ <- assays(fc_se[,iSamples])[[normname]];

```

### Apply batch-adjustment by "Run", grouped by "Group"

This section performs batch adjustment when there are multiple
distinct values in the `"Run"` column.

```{r, batch_adjust_run, dependson="norm_quantile"}
## Batch adjustment
batch_adjusted <- FALSE;
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   fc_batch <- nameVector(colData(fc_se[,iSamples])[[run_colname]],
      iSamples);
   fc_group <- nameVector(colData(fc_se[,iSamples])[[group_colname]],
      iSamples);
   fc_se_b <- tryCatch({
      jamses::se_normalize(fc_se[,iSamples],
         assay_names=unique(c("counts", normname)),
         method=c("limma_batch_adjust"),
         params=list(
            limma_batch_adjust=list(batch=fc_batch,
               group=fc_group))
      );
   }, error=function(e){
      printDebug("Error during batch adjustment by se_normalize():");
      print(e);
      NULL;
   });
   bnormname <- paste0("limma_batch_adjust_", normname);
   if (length(fc_se_b) > 0) {
      batch_adjusted <- TRUE;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se_b[,iSamples])[["limma_batch_adjust_counts"]];
      ## quantile-normalized data, batch-adjusted
      iMatrixQB <- assays(fc_se_b[,iSamples])[[bnormname]];
   } else {
      ## Do not perform batch adjustment
      fc_se_b <- fc_se;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
      ## quantile-normalized data, batch-adjusted
      bnormname <- normname;
      iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
   }
} else {
   ## Do not perform batch adjustment
   fc_se_b <- fc_se;
   ## batch-adjusted raw data
   iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
   ## quantile-normalized data, batch-adjusted
   bnormname <- normname;
   iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
}


```

### Create MA-plots

#### Un-normalized data, global-centered

```{r, ma_layout, dependson="import_fc"}
## Determine blank plot positions to organize Group and Run
Group_ctm <- as.matrix(table(colData(fc_se)[,c(group_colname, run_colname)]));
sub_n <- colMaxs(Group_ctm);
ctm_n <- matrix(cumsum(rep(sub_n, nrow(Group_ctm))),
   ncol=ncol(Group_ctm),
   byrow=TRUE);
#ctm_n;
ma_ncol <- sum(sub_n);
blank_pos <- unlist(lapply(1:nrow(Group_ctm), function(i){
   rmNULL(lapply(1:ncol(Group_ctm), function(j){
      iseq <- seq(from=ctm_n[i,j], by=-1, length.out=sub_n[j]);
      head(iseq, sub_n[j] - Group_ctm[i,j])
   }))
}));

```

```{r, ma_raw_global, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
jp_raw <- jammaplot(
   noiseFloor(iMatrix[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("data (counts)",
      "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Un-normalized data, centered by "Run"

```{r, ma_raw_run, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
# Only create when there are multiple Run values
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
jp_raw_run <- jammaplot(
   noiseFloor(iMatrix[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("data (counts)",
      "centered by 'Run'"),
   centerGroups=colData(fc_se[,iSamples])[[run_colname]],
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
)
}
```

#### Normalized data, global-centered

```{r, ma_q_global, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
jp_q <- jammaplot(
   noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c(paste0("data (", normname, ")"),
      "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Normalized data, centered by "Run"

```{r, ma_q_run, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
# Only create when there are multiple Run values
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
jp_q_run <- jammaplot(
   noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c(paste0("data (", normname, ")"),
      "centered by 'Run'"),
   centerGroups=colData(fc_se[,iSamples])[[run_colname]],
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
)
}
```

#### Normalized data, centered by "Group", "Run"

If there are replicates per Group_Run, then MA-plots will
be displayed below, centered by Group and Run.

```{r, ma_q_group_run, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
grouprun_ct <- tcount(pasteByRow(colData(fc_se[,iSamples])[,c(group_colname, run_colname)]));
if (any(grouprun_ct > 1)) {
   subtitle_color_set <- as.list(as.data.frame(
      t(cbind(color_sub1[iSamples], run_colors1[iSamples]))));
   subtitle_colors <- colorjam::blend_colors(subtitle_color_set);
   
   jp_q_group_run <- jammaplot(
      noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
         minimum=1,
         newValue=NA),
      ncol=ma_ncol,
      blankPlotPos=blank_pos,
      maintitle=c(paste0("data (", normname, ")"),
         paste0("centered by ",
            cPaste(paste0("'", c(group_colname, run_colname), "'"))
         )),
      centerGroups=pasteByRow(colData(fc_se[,iSamples])[,c(group_colname, run_colname)]),
      groupSuffix="",
      subtitlePreset="bottomleft",
      subtitle=paste0("by ", iGroup[iSamples], ",", Run[iSamples]),
      subtitleBoxColor=subtitle_colors,
      useMean=TRUE,
      displayMAD=TRUE,
      outlierMAD=2,
      ablineV=MGM,
      titleBoxColor=color_sub1[iSamples]
   )
}
```

#### Normalized data, global-centered, batch-adjusted by "Run"

This set of MA-plots represents the data used for statistical comparisons
(where applicable).

Note that panels display the MAD factor, which is a multiple of
variability of each sample to the median variability across all samples.
The threshold "2x MAD" is defined as one replicate having at least
twice the median variability across all samples.

> **At this point, outlier samples are not removed from
downstream statistical comparisons.**

```{r, ma_q_b_run, fig.height=10, fig.width=14, dependson=c("batch_adjust_run", "ma_layout")}
jp_q_b_run <- jammaplot(
   noiseFloor(iMatrixQB[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c(paste0("data (", bnormname, ")"),
      "global-centered",
      paste0("batch-adjusted by '", run_colname, "'")
   ),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("adj. by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Optional subset of sample outliers

Initially, all samples are used, but this section is a placeholder
for outlier detection and removal.

```{r, sample_outliers}
## Optional subset of samples (e.g. MA-plot outliers)
iSamplesUse <- iSamples;
printDebug("Using samples:", iSamplesUse);
```

### Sample Correlation Heatmaps


#### Normalized data, centered by Group control

```{r, cor_hm, fig.height=12, fig.width=12, dependson=c("norm_quantile", "sample_outliers")}
## Define centering control samples
iControl <- head(levels(iGroup), 1);
iControlSamples <- intersect(
   subset(colData(fc_se), colData(fc_se)[[group_colname]] %in% iControl)[[sample_colname]],
   iSamplesUse);
printDebug("Using control: ", iControl);
printDebug("Control Samples: ", iControlSamples);

## Center iMatrixQ
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQctr <- centerGeneData(iMatrixQ[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorM <- cor(use="pairwise.complete.obs",
   iMatrixQctr[,iSamplesUse,drop=FALSE]);

## hcluster
iCorM_hc <- tryCatch({
   amap::hcluster(iCorM);
}, error=function(e){
   printDebug("Error during hcluster(iCorM):");
   print(e);
   FALSE;
});

## ComplexHeatmap
hm_cor <- ComplexHeatmap::Heatmap(
   iCorM[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=iCorM_hc,
   cluster_columns=iCorM_hc,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_cor,
   column_title=paste0("Counts (", normname, ")"))

```


#### Normalized data, batch-adjusted, global-centered

```{r, cor_hm_b, fig.height=12, fig.width=12, dependson=c("batch_adjust_run", "sample_outliers")}
## Center iMatrixQB
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQBctrGlobal <- centerGeneData(iMatrixQB[,iSamples,drop=FALSE],
   mean=TRUE);

## Correlation matrix
iCorMBglobal <- cor(use="pairwise.complete.obs",
   iMatrixQBctrGlobal[,iSamplesUse,drop=FALSE])

## hcluster
iCorMBglobal_hc <- tryCatch({
   amap::hcluster(iCorMBglobal);
}, error=function(e){
   printDebug("Error during hcluster(iCorM):");
   print(e);
   FALSE;
});

## ComplexHeatmap
hm_corbg <- ComplexHeatmap::Heatmap(
   iCorMBglobal[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=iCorMBglobal_hc,
   cluster_columns=iCorMBglobal_hc,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_corbg,
   column_title=paste0("Counts (", bnormname, ")"))

```

#### Normalized data, batch-adjusted, centered by Group control

```{r, cor_hm_bg, fig.height=12, fig.width=12, dependson=c("batch_adjust_run", "sample_outliers")}
## Center iMatrixQB
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQBctr <- centerGeneData(iMatrixQB[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorMB <- cor(use="pairwise.complete.obs",
   iMatrixQBctr[,iSamplesUse,drop=FALSE])

## hcluster
iCorMB_hc <- tryCatch({
   amap::hcluster(iCorMB);
}, error=function(e){
   printDebug("Error during hcluster(iCorM):");
   print(e);
   FALSE;
});

## ComplexHeatmap
hm_corb <- ComplexHeatmap::Heatmap(
   iCorMB[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=iCorMB_hc,
   cluster_columns=iCorMB_hc,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_corb,
   column_title=paste0("Counts (", bnormname, ")"))

```

### Plot Peak Width by Peak Signal

This section is intended to help evaluate peak width, and peak slice width
when using `mergeSplitCountBed`.

#### Raw data

```{r, peak_width_signal_raw, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 4, 2) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQ[, iCol],
      minimum=1,
      newValue=NA);
   iy <- log2(1 + rowData(fc_se[iRows, ])$Length);
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main=iCol,
      yaxt="n");
   minorLogTicksAxis(2,
      logBase=2,
      offset=1,
      displayBase=10);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="bottomleft",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```

#### Normalized data


```{r, peak_width_signal_norm, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 3, 1) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQB[, iCol],
         minimum=1,
         newValue=NA);
   iy <- log2(1 + rowData(fc_se[iRows, ])$Length);
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main=iCol,
      yaxt="n");
   minorLogTicksAxis(2,
      logBase=2,
      offset=1,
      displayBase=10);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="bottomleft",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```


**Normal scale y-axis peak width:**

```{r, peak_width_signal_norm2, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 3, 1) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQB[, iCol],
         minimum=1,
         newValue=NA);
   iy <- rowData(fc_se[iRows, ])$Length;
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main=iCol);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="bottomleft",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```

