---
title: "Slicejam Analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
    df_print: kable
    self_contained: true
    keep_md: true
---
   
About this document:

* prepared on **`r format(Sys.time(), "%a %b %d %X %Y")`**
* using **`r R.version.string`**
* R slicejam package version **`r packageVersion("slicejam")`**.

### R environment setup

This document can be reproduced using the R code displayed in
this document. The initial steps involve loading R packages
into memory.

#### Rmarkdown options

```{r, setup, include=TRUE}
#    css: liao_style.css
## To render this page:
## In the farrisSeq folder run:
## rmarkdown::render("Adam19_Rnaseq_Analysis_vM21.Rmd", "html_document", knit_root_dir=getwd(), output_dir=getwd());
##    css: /Users/wardjm/Projects/R-scripts/liao_style.css

## Define the cache so it is dependent upon:
## - the versions of a few R packages,
## - the environmental variable FC_FILE
knitr::opts_chunk$set(cache=TRUE,
   autodep=TRUE,
   dev=c("png", "cairo_pdf"),
   dev.args=list(
      pointsize=14
   ),
   fig.align="center",
   fig.height=9,
   fig.width=8,
   pdf.options(encoding="ISOLatin9.enc"),
   cache.path=Sys.getenv("CACHEDIR", "cache"),
   cache.lazy=FALSE,
   cache.extra=list(
      packageVersion("jamba"),
      packageVersion("jamma"),
      packageVersion("platjam"),
      packageVersion("slicejam"),
      packageVersion("jamses"),
      paste0("CURATION_TXT:", Sys.getenv("CURATION_TXT"))
   )
);

options("kable_styling_position"="left",
   "kable_styling_full_width"=FALSE,
   "kable_styling_bootstrap_options"=c("striped",
      "hover",
      "condensed",
      "responsive")
);
## Set font to Arial which helps compatibility in Adobe Illustrator,
## and helps display of up and down arrow characters in Venn diagrams
par("family"="Arial",
   "lend"="square",
   "ljoin"="mitre");
```

#### SliceJam analysis parameters

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseSlicejam1" role="button" aria-expanded="false" aria-controls="collapseSlicejam1">(Expand...)</a></p>
<div class="collapse" id="collapseSlicejam1"><div class="card card-body">

```{r, slicejam_minimum_version}
# check version of slicejam
min_slicejam_version <- "0.0.37.900";
if (compareVersion(as.character(packageVersion("slicejam")), "0.0.37.900") < 0) {
   jamba::kable_coloring(
   data.frame(check.names=FALSE,
      stringsAsFactors=FALSE,
      `.libPaths()`=.libPaths()),
   caption=paste0(
      "R library paths in use."));

   stop(paste0("Slicejam must be at least slicejam-", min_slicejam_version,
      ", please verify .libPaths():\n",
      jamba::cPaste(paste0("'", .libPaths(), "'"), sep="\n")));
}
```


```{r, slicejam_args, include=TRUE, cache=FALSE}
sliceargs <- slicejam::get_slicejam_args(verbose=TRUE,
   name="analysis",
   type="data.frame");
if ("sliceargs_df" %in% names(attributes(sliceargs))) {
   sliceargs_df <- attr(sliceargs, "sliceargs_df");
   suppressPackageStartupMessages(library(kableExtra));
   jamba::kable_coloring(
      sliceargs_df,
      caption=paste0(
         "slicejam arguments:"));
}

DRYRUN <- sliceargs$DRYRUN;
ATAC <- sliceargs$ATAC;
fc_file <- sliceargs$fc_file;
fc_filepath <- sliceargs$fc_filepath;
CURATION_TXT <- sliceargs$CURATION_TXT;
GTF <- sliceargs$GTF;
GTFNAME <- sliceargs$GTFNAME;
MASK <- sliceargs$MASK;
DRYRUN <- sliceargs$DRYRUN;
OUTDIR <- sliceargs$OUTDIR;
CACHEDIR <- sliceargs$CACHEDIR;
ATAC <- sliceargs$ATAC;
MGM <- sliceargs$MGM;
NORM <- sliceargs$NORM;
NORMMIN <- sliceargs$NORMMIN;
NORMSHORT <- sliceargs$NORMSHORT;
GROUPCHECK <- sliceargs$GROUPCHECK;
FACTOR_ORDER <- sliceargs$FACTOR_ORDER;
SAVE_RDATA <- sliceargs$SAVE_RDATA;

detectedTx <- sliceargs$DETECTED_TX;
detectedGenes <- sliceargs$DETECTED_GENES;

upstream_promoter <- sliceargs$UPSTREAM_PROMOTER;
downstream_promoter <- sliceargs$DOWNSTREAM_PROMOTER;
upstream_tts <- sliceargs$UPSTREAM_TTS;
downstream_tts <- sliceargs$DOWNSTREAM_TTS;

fc_base <- sliceargs$fc_base;
fc_basedir <- sliceargs$fc_basedir;
fc_html <- sliceargs$fc_html;
knit_root_dir <- sliceargs$knit_root_dir;

```


#### R package initialization


```{r, r_initialization, cache=FALSE}
library(pacman)
suppressPackageStartupMessages(p_load_gh("jmw86069/jamba"));

jamba::kable_coloring(
   data.frame(check.names=FALSE,
      stringsAsFactors=FALSE,
      `.libPaths()`=.libPaths()),
   caption=paste0(
      "R library paths in use."));

options("warn"=-1);
suppressPackageStartupMessages(p_load_gh("jmw86069/colorjam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/slicejam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/jamses"));
suppressPackageStartupMessages(p_load_gh("jmw86069/jamma"));
suppressPackageStartupMessages(p_load_gh("jmw86069/splicejam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/multienrichjam"));

suppressPackageStartupMessages(p_load("SummarizedExperiment"));
suppressPackageStartupMessages(p_load("data.table"));
suppressPackageStartupMessages(p_load("matrixStats"));
suppressPackageStartupMessages(p_load("extrafont"));
#suppressPackageStartupMessages(library(GenomicFeatures));
suppressPackageStartupMessages(p_load("rtracklayer"));
suppressPackageStartupMessages(p_load(methods));
suppressPackageStartupMessages(p_load("kableExtra"));

```

</div></div>

## Description of workflow

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseDescription1" role="button" aria-expanded="false" aria-controls="collapseDescription1">(Expand...)</a></p>
<div class="collapse" id="collapseDescription1"><div class="card card-body">

* Import the first `.fc` file found in the current working directory.

    * Optionally use "FC_FILE" environmental variable.

* Curate filename columns by one of two processes:

    * A file defined by `"CURATION_TXT"` is used to parse the colnames
    in the featureCounts file into three columns: `"Sample"`, `"Group"`, `"Run"`

* An R object `SummarizedExperiment` is created, using the columns
from featureCounts to populate rows, such as
`"Geneid", "Chr", "Start", "End", "Strand", "Length"`.
* Assign categorical colors to the unique values in the `"Group"` column.

* Normalization and batch adjustment

    * Normalize peak counts across all samples using the method defined by `NORM`.
    * When `"Run"` contains more than one unique value,
    apply batch adjustment using the `"Run"` column to define batch,
    and `"Group"` to define groups, using normalized data.

* Create MA-plots:

    * un-normalized data, global-centered
    * un-normalized data, centered by "Run"
    * normalized data, global-centered
    * normalized data, centered by "Run"
    * normalized data, centered by "Group","Run"
    * normalizeddata, global-centered, batch-adjusted by "Run"

* Create sample correlation heatmaps:

   * normalized data, centered by the first factor level
   in "Group", intended to act as a control group.
   * normalized data, batch-adjusted by "Run", centered by the
   first factor level in "Group".

* Statistical comparisons

   * Use the `"Group"` column to define groups, design, and contrasts.
   
       * Note that contrasts are limited to comparisons to the first
       `"Group"` level.
   
   * Use normalized, batch-adjusted counts.
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05
   * Apply secondary max group mean filtering with MGM=**`r MGM`**, to
   reduce the number of peaks used for statistical testing
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05

* Create volcano plots

    * Plot each contrast using the P-value, then using adjusted P-value.

* Create Venn diagrams:

    * Include each set of two contrasts with "all peak" hits.
    * Include each set of two contrasts with "MGM-filtered peak" hits.

* Annotate peaks

    * Associate each peak with one or more overlapping feature_types
    * Call one feature_type_winner per peak, based upon order: promoter, TTS, exon, intron, extragenic
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "all peaks" comparison
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "MGM-filtered peaks" comparison

* Re-annotate peaks into broad_peaks by merging adjacent slices

    * Count broad_peak hits from the "all peaks" comparison.
    * Count broad_peak hits from the "MGM-filtered peaks" comparison.

* Heatmap of peak hits across samples

    * Heatmap of peak hits from the "all peaks" comparison.
    * Heatmap of peak hits from the "MGM-filtered peaks" comparison.

* Optional ATAC-mode annotation to "active promoter genes"

    * Define active promoter genes by presence of ATAC-seq peak in promoter region.
    * Annotate peaks by nearest active promoter peak, along with distance.

* Save stats results to text files

    * Save "all peak" stats to tab-delimited text file.
    * Save "MGM-filtered peak" stats to tab-delimited text file.
    * This file is also used to create the "table of Venn intersections."

* Optionally save RData session

</div></div>


## Analysis workflow


### Import featureCounts data

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseImport1" role="button" aria-expanded="false" aria-controls="collapseImport1">(Expand...)</a></p>
<div class="collapse" id="collapseImport1"><div class="card card-body">

```{r, import_fc}
## Import
fc_data <- import_featurecounts(fc_file);

## Add zero-padding to peak labels
fc_data[,1] <- paste0("peak_",
   jamba::padInteger(seq_len(nrow(fc_data))));
rownames(fc_data) <- fc_data[,1];

# featureCounts colnames
fc_colnames <- attr(fc_data, "fc_colnames");
filenames <- attr(fc_data, "filenames");
if (length(fc_colnames) == 0) {
   stop(paste0("None of the featureCounts colnames were detected: ",
      jamba::cPaste(fc_colnames)));
}


## Use curation_txt if that file exists
curation_txt <- CURATION_TXT;
curation_df <- data.table::fread(curation_txt,
   data.table=FALSE,
   sep="\t");
filename_colname <- "Filename";
sample_df <- slicejam::curate_to_df_by_pattern(
   x=filenames,
   input_colname=filename_colname,
   df=curation_df);
if (nrow(sample_df) == 0) {
   jamba::printDebug("Error: ",
      c("The curation_txt file '",
         curation_txt,
         "' did not match filenames."),
      sep="",
      fgText=c("red", "dodgerblue"));
   jamba::printDebug("filenames: ",
      paste0("\n'", filenames, "'"));
   jamba::printDebug("curation_txt contents:");
   print(curation_df);
   stop("curation_txt file did not match filenames.");
}

## Check for samples not curated, and therefore dropped from the analysis
if (length(sample_df[[filename_colname]]) < length(filenames)) {
   samples_dropped <- setdiff(filenames,
      sample_df[[filename_colname]]);
   samples_retained <- sample_df[[filename_colname]];
   samples_retmax <- max(c(
      length(samples_dropped),
      length(samples_retained)));
   samples_dropped <- unlist(rep(list(samples_dropped, ""),
      c(1, samples_retmax - length(samples_dropped))));
   samples_retained <- unlist(rep(list(samples_retained, ""),
      c(1, samples_retmax - length(samples_retained))));
   
   kable_coloring(
      data.frame(
         samples_retained=samples_retained,
         samples_dropped=samples_dropped),
      row.names=FALSE,
      caption=paste0(
         "Samples retained and dropped."));
}

sample_colname <- head(jamba::provigrep(
   c("^label$",
      "^sample$",
      "^sample.*(name|id)$"),
   colnames(sample_df)), 1);
if (length(sample_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Label","' nor '", "Sample", "' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Sample column was recognized.");
}
sample_df[[sample_colname]] <- jamba::makeNames(sample_df[[sample_colname]]);
rownames(sample_df) <- sample_df[[sample_colname]];

run_colname <- head(jamba::provigrep(
   c("^batch$",
      "^run$"),
   colnames(sample_df)), 1);
if (length(run_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Run","' or '","Batch","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
}

group_colname <- head(jamba::provigrep(
   c("^Group$",
      "^group.*name$",
      "group$"),
   colnames(sample_df)), 1);
if (length(group_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Group","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Group column was recognized.");
}
## Ensure group names have no invalid characters like "-"
sample_df[[group_colname]] <- gsub("[%]", "pct",
   gsub("[- ]+",
      ".",
      sample_df[[group_colname]]));
sample_df[[group_colname]] <- factor(sample_df[[group_colname]],
   levels=unique(sample_df[[group_colname]]));

normgroup_colname <- head(jamba::provigrep(
   c("^normgroup$",
      "norm.*group"),
   colnames(sample_df)), 1);
if (length(normgroup_colname) == 0) {
   normgroup_colname <- "normgroup";
   sample_df[[normgroup_colname]] <- sample_df[[group_colname]];
}

## Optional color assignment
color_colname <- head(jamba::provigrep(
   c("^Color$", "Color$"),
   colnames(sample_df)), 1);

## Sort sample_df by group
sample_df <- jamba::mixedSortDF(sample_df,
   byCol=c(group_colname,
      sample_colname));

## Define useful sample annotation vectors
iSamples <- sample_df[[sample_colname]];
Run <- jamba::nameVector(sample_df[[run_colname]],
   iSamples);
iGroup <- jamba::nameVector(sample_df[[group_colname]],
   iSamples);
iNormGroup <- jamba::nameVector(sample_df[[normgroup_colname]],
   iSamples);

## Make sure the order in filenames is propagated
filenames <- jamba::nameVector(sample_df[[filename_colname]],
   iSamples);
fc_data <- fc_data[,c(fc_colnames, filenames),drop=FALSE];

## Print sample annotation table
curation_color_list <- platjam::design2colors(
   plot_type="none",
   sample_df)
kable_coloring(sample_df,
   colorSub=curation_color_list,
   caption=paste0(
      "Sample Groups defined by curation_txt: '",
      basename(curation_txt),
      "'"));
if (any(!sapply(c("iSamples", "Run", "iGroup", "iNormGroup"), exists))) {
   stop("One or more of iSamples,Run,Group could not be defined.");
}

## Rename featureCounts columns to match iSamples
kable_coloring(
   data.frame(
      Rename_from=filenames,
      Rename_to=iSamples),
   row.names=FALSE,
   caption=paste0(
      "Renamed featureCounts colnames."));
fc_data <- jamba::renameColumn(fc_data,
   from=filenames,
   to=iSamples);

## Make SummarizedExperiment
fc_se <- SummarizedExperiment(
   assays=list(counts=log2(1+as.matrix(fc_data[,iSamples,drop=FALSE]))),
   rowData=fc_data[,fc_colnames,drop=FALSE],
   colData=sample_df);

## Re-order samples by Group
iSamples <- mixedSortDF(colData(fc_se),
   byCols=group_colname)[[sample_colname]];
fc_se <- fc_se[,iSamples];
Run <- nameVector(colData(fc_se)[[run_colname]],
   colnames(fc_se));
iGroup <- nameVector(colData(fc_se)[[group_colname]],
   colnames(fc_se));
```

</div></div>

#### Assign colors

Base filename: `define_colors`

```{r define_colors, fig.width=16, fig.height=9}
## Define sample colors
color_sub <- NULL;
if (length(color_colname) == 1) {
   # assume group color assignment
   colordf <- unique(data.frame(check.names=FALSE,
      colData(fc_se)[,c(group_colname, color_colname), drop=FALSE]));
   if (length(jamba::tcount(colordf[[group_colname]], 2)) == 0) {
      # no duplicate groups, therefore these are group colors
      color_sub <- jamba::nameVector(
         colordf[,c(color_colname, group_colname)]);
   } else {
      # these must be sample colors
      colordf <- data.frame(check.names=FALSE,
         colData(fc_se)[,c(sample_colname, color_colname), drop=FALSE]);
      color_sub <- jamba::nameVector(
         colordf[,c(color_colname, sample_colname)]);
   }
}
sample_color_list <- platjam::design2colors(
   data.frame(check.names=FALSE,
      colData(fc_se)),
   group_colnames=group_colname,
   lightness_colnames=run_colname,
   color_sub=color_sub)

group_colors <- sample_color_list[[group_colname]];
run_colors <- sample_color_list[[run_colname]];

color_sub1 <- nameVector(group_colors[colData(fc_se)[[group_colname]]],
   colData(fc_se)[[sample_colname]]);

run_colors1 <- nameVector(run_colors[colData(fc_se[,iSamples])[[run_colname]]],
   iSamples);
```

#### Statistical Contrasts

Base filename: `define_contrasts`

```{r, define_contrasts, fig.height=12, fig.width=12}
sample_txt_file <- file.path(OUTDIR,
   "curated_samples.txt");
data.table::fwrite(
   x=sample_df,
   file=sample_txt_file,
   sep="\t",
   col.names=TRUE);

## define experimental design groups and contrasts
sedesign <- jamses::groups_to_sedesign(
   data.frame(check.names=FALSE,
      colData(fc_se[, iSamples])[, group_colname, drop=FALSE]),
   factor_order=FACTOR_ORDER,
   max_depth=1);
iDesign <- sedesign@design;
iContrasts <- sedesign@contrasts;

contrastNames <- NULL;
if (length(iContrasts) > 0) {
   contrastNames <- colnames(iContrasts);
}
contrast_txt_file <- file.path(OUTDIR,
   "contrasts.txt");
kc <- NULL;
if (file.exists(contrast_txt_file)) {
   contrastNames <- gsub("^[ \t\r\n]+|[ \t\r\n]+$",
      "",
      readLines(contrast_txt_file));
   iContrasts <- tryCatch({limma::makeContrasts(
         contrasts=contrastNames,
         levels=iDesign);
   }, error=function(e){
      jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
         contrast_txt_file);
      print(e);
      NULL;
   })
   if (length(iContrasts) > 0) {
      rownames(iContrasts) <- colnames(iDesign);
      sedesign@contrasts <- iContrasts;
      kc <- kable_coloring(
         data.frame(
            Note="contrastNames were loaded from file: ",
            File=paste0("'", contrast_txt_file, "'")),
         row.names=FALSE);
   } else {
      contrastNames <- NULL;
      iContrasts <- NULL;
   }
} else {
   if (length(contrastNames) > 0) {
      writeLines(text=contrastNames,
         con=contrast_txt_file);
      kc <- kable_coloring(
         data.frame(Note="contrastNames were stored in file: ",
            paste0("'", contrast_txt_file, "'")),
         row.names=FALSE);
   } else {
      kc <- kable_coloring(
         data.frame(Note=paste0("There are no contrasts defined."),
         row.names=FALSE));
   }
}
kc;

kable_coloring(
   data.frame(
      `Contrast Names`=contrastNames,
      `Abbreviated Comp`=jamses::contrast2comp(contrastNames)),
   row.names=FALSE,
   caption=paste0(
      "Contrasts defined:"));

# plot sedesign
plot_sedesign(sedesign,
   arrow_ex=0.7,
   axis1=2,
   axis2=1,
   axis3=3,
   axis4=4)

## Optional GROUPCHECK which exits here to allow reviewing the sample grouping
if (GROUPCHECK) {
   #txt_ext <- paste0(".curated_samples.txt");
   #sample_txt_file <- file.path(OUTDIR,
   #   gsub("[.]fc$|[.]fc[.]txt$",
   #      txt_ext,
   #      basename(fc_file)));
   printDebug("Curated sample data for review: '",
      sample_txt_file,
      "' and contrastNames file '",
      contrast_txt_file,
      "'");
   knitr::knit_exit(fully=TRUE);
}

```

#### Peak Sizes {.tabset}

If the featureCounts output included column name `"Length"`,
the plots below will show the distribution of lengths.

##### Size distribution

Base filename: `peak_sizes`

```{r, peak_sizes, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes",
      xaxt="n",
      yaxt="n",
      breaks=100);
   title(xlab="Peak width", line=4)
   axis1at <- pretty(par("usr")[1:2])
   axis(1,
      las=2,
      cex.axis=1.3,
      at=axis1at,
      labels=formatInt(axis1at))
   axis2at <- pretty(par("usr")[3:4])
   axis(2,
      las=2,
      cex.axis=1.3,
      at=axis2at,
      labels=formatInt(axis2at))
}
```

##### Size distribution (log scale)

Base filename: `peak_sizes_log`

```{r, peak_sizes_log, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes (log-scaled x-axis)",
      log="x",
      xlab="Peak width",
      breaks=100);
}
```

#### Max Group Mean (MGM) Threshold

The maxGroupMean threshold is defined by the environmental
variable `MGM`, used to customize the threshold during the stats
tests, applied to signal log2-transformed with `log2(1 + x)`,
and therefore the default `MGM=4` defines a threshold of `15 counts`
in normal space, using `( 2^MGM ) - 1` to reverse the
transform: `log2(1 + x)`


```{r, mgm_arg}
## Get maxGroupMean threshold argument
MGM <- Sys.getenv("MGM", 4);
MGM <- jamba::rmNA(as.numeric(unlist(strsplit(MGM, "[ ,]+"))));
if (length(MGM) == 0) {
   MGM <- c(4);
}
kable_coloring(
   data.frame(check.names=FALSE,
      `max group mean threshold (MGM)`=MGM),
   row.names=FALSE);
```


### Apply Normalization


```{r, apply_norm, dependson="import_fc"}
kable_coloring(
   data.frame(check.names=FALSE,
      `normalization method (NORM)`=NORM),
   row.names=FALSE);

## Raw data
iMatrix <- assays(fc_se[,iSamples])[["counts"]];

## Perform normalization
if ("median" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("mediangroup" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            centerGroups=iNormGroup,
            controlSamples=iSamples,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("quantile" %in% NORM) {
   ## Evaluate quantile normalization
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("quantile"));
   normname <- "quantile_counts";
} else if ("none" %in% NORM) {
   normname <- "counts";
}

## Normalized data
iMatrixQ <- assays(fc_se[,iSamples])[[normname]];

```

#### Batch Adjust by "Run", Grouped by "Group"

This section performs batch adjustment when there are multiple
distinct values in the `"Run"` column.

```{r, batch_adjust_run, dependson="apply_norm"}
## Batch adjustment
batch_adjusted <- FALSE;
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   fc_batch <- nameVector(colData(fc_se[,iSamples])[[run_colname]],
      iSamples);
   fc_group <- nameVector(colData(fc_se[,iSamples])[[group_colname]],
      iSamples);
   fc_se_b <- tryCatch({
      jamses::se_normalize(fc_se[,iSamples],
         assay_names=unique(c("counts", normname)),
         method=c("limma_batch_adjust"),
         params=list(
            limma_batch_adjust=list(
               batch=fc_batch[iSamples],
               group=fc_group[iSamples]))
      );
   }, error=function(e){
      printDebug("Error during batch adjustment by se_normalize():");
      print(e);
      NULL;
   });
   bnormname <- paste0("limma_batch_adjust_", normname);
   if (length(fc_se_b) > 0) {
      batch_adjusted <- TRUE;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se_b[,iSamples])[["limma_batch_adjust_counts"]];
      ## quantile-normalized data, batch-adjusted
      iMatrixQB <- assays(fc_se_b[,iSamples])[[bnormname]];
   } else {
      ## Do not perform batch adjustment
      fc_se_b <- fc_se;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
      ## quantile-normalized data, batch-adjusted
      bnormname <- normname;
      iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
   }
} else {
   ## Do not perform batch adjustment
   fc_se_b <- fc_se;
   ## batch-adjusted raw data
   iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
   ## quantile-normalized data, batch-adjusted
   bnormname <- normname;
   iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
}


```

### MA-Plots {.tabset}


```{r, ma_layout, dependson="apply_norm"}
## Determine blank plot positions to organize Group and Run
blank_pos <- NULL;
ma_ncol <- ceiling(sqrt(ncol(fc_se)));
ma_nrow <- ceiling(sqrt(ncol(fc_se)));
tryCatch({
   Group_ctm <- as.matrix(table(
      data.frame(check.names=FALSE,
         colData(fc_se)[,c(group_colname, run_colname)])));
   sub_n <- apply(Group_ctm, 2, function(i){
      max(i, na.rm=TRUE)});
   ctm_n <- matrix(cumsum(rep(sub_n, nrow(Group_ctm))),
      ncol=ncol(Group_ctm),
      byrow=TRUE);
   #ctm_n;
   ma_ncol <- sum(sub_n);
   ma_nrow <- nrow(Group_ctm);
   blank_pos <- unlist(lapply(seq_len(nrow(Group_ctm)), function(i){
      jamba::rmNULL(lapply(seq_len(ncol(Group_ctm)), function(j){
         iseq <- seq(from=ctm_n[i,j], by=-1, length.out=sub_n[j]);
         head(iseq, sub_n[j] - Group_ctm[i,j])
      }))
   }));
}, error=function(e){
   print(e);
})
ma_height <- 5 + sqrt(ma_nrow) * 3;
ma_width <- 5 + sqrt(ma_ncol) * 3;
```

#### Un-normalized data {.tabset}

##### global-centered

```{r, ma_raw_global, fig.height=ma_height, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
jp_raw_attr <- attributes(jammaplot(
   noiseFloor(iMatrix[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("data (counts)",
      "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
))
```

##### centered by Run

```{r, ma_raw_run, fig.height=ma_height, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
# Only create when there are multiple Run values
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   jp_raw_run_attr <- attributes(jammaplot(
      noiseFloor(iMatrix[,iSamples,drop=FALSE],
         minimum=1,
         newValue=NA),
      ncol=ma_ncol,
      blankPlotPos=blank_pos,
      maintitle=c("data (counts)",
         "centered by 'Run'"),
      centerGroups=colData(fc_se[,iSamples])[[run_colname]],
      groupSuffix="",
      subtitlePreset="bottomleft",
      subtitle=paste0("by ", Run[iSamples]),
      subtitleBoxColor=run_colors1[iSamples],
      useMean=TRUE,
      ablineV=MGM,
      titleBoxColor=color_sub1[iSamples]
   ))
}
```

#### Normalized data {.tabset}

##### global-centered

```{r, ma_norm_global, fig.height=ma_height, fig.width=ma_width, dependson=c("apply_norm", "ma_layout")}
jp_norm_attr <- attributes(jammaplot(
   noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c(paste0("data (", normname, ")"),
      "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
))
```

##### centered by Run

```{r, ma_norm_run, fig.height=ma_height, fig.width=ma_width, dependson=c("apply_norm", "ma_layout")}
# Only create when there are multiple Run values
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   jp_norm_run_attr <- attributes(jammaplot(
      noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
         minimum=1,
         newValue=NA),
      ncol=ma_ncol,
      blankPlotPos=blank_pos,
      maintitle=c(paste0("data (", normname, ")"),
         "centered by 'Run'"),
      centerGroups=colData(fc_se[,iSamples])[[run_colname]],
      groupSuffix="",
      subtitlePreset="bottomleft",
      subtitle=paste0("by ", Run[iSamples]),
      subtitleBoxColor=run_colors1[iSamples],
      useMean=TRUE,
      displayMAD=TRUE,
      outlierMAD=2,
      ablineV=MGM,
      titleBoxColor=color_sub1[iSamples]
   ))
}
```

##### centered by Group, Run

```{r, ma_norm_group_run, fig.height=ma_height, fig.width=ma_width, dependson=c("apply_norm", "ma_layout")}
grouprun_ct <- tcount(pasteByRow(colData(fc_se[,iSamples])[,c(group_colname, run_colname)]));
if (any(grouprun_ct > 1)) {
   subtitle_color_set <- as.list(as.data.frame(
      t(cbind(color_sub1[iSamples], run_colors1[iSamples]))));
   subtitle_colors <- colorjam::blend_colors(subtitle_color_set);
   
   jp_norm_group_run_attr <- attributes(jammaplot(
      noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
         minimum=1,
         newValue=NA),
      ncol=ma_ncol,
      blankPlotPos=blank_pos,
      maintitle=c(paste0("data (", normname, ")"),
         paste0("centered by ",
            cPaste(paste0("'", c(group_colname, run_colname), "'"))
         )),
      centerGroups=pasteByRow(colData(fc_se[,iSamples])[,c(group_colname, run_colname)]),
      groupSuffix="",
      subtitlePreset="bottomleft",
      subtitle=paste0("by ", iGroup[iSamples], ",", Run[iSamples]),
      subtitleBoxColor=subtitle_colors,
      useMean=TRUE,
      displayMAD=TRUE,
      outlierMAD=2,
      ablineV=MGM,
      titleBoxColor=color_sub1[iSamples]
   ))
}
```

This MA-plot visualization is intended to show variability across
biological replicates within the same batch `"Run"`. If there
is only one biological replicate in the same batch, the sample
will be shown as a horizontal line at y=0.

#### Normalize, batch-adjusted {.tabset}

##### global-centered, batch-adjusted by "Run"

```{r, ma_norm_b_run, fig.height=ma_height, fig.width=ma_width, dependson=c("batch_adjust_run", "ma_layout")}
jp_norm_b_run_attr <- attributes(jammaplot(
   noiseFloor(iMatrixQB[,iSamples,drop=FALSE],
      minimum=1,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c(paste0("data (", bnormname, ")"),
      "global-centered",
      paste0("batch-adjusted by '", run_colname, "'")
   ),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("adj. by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   ablineV=MGM,
   titleBoxColor=color_sub1[iSamples]
))
```

> **Note: At this point, outlier samples are not removed from
downstream statistical comparisons.**

This set of MA-plots represents the data used for statistical comparisons
(where applicable).

Note that panels display the MAD factor, which is a multiple of
variability of each sample to the median variability across all samples.
The threshold "2x MAD" is defined as one replicate having at least
twice the median variability across all samples.

#### Sample outliers

Initially, all samples are used, but this section is a placeholder
for outlier detection and removal.

```{r, sample_outliers}
## Optional subset of samples (e.g. MA-plot outliers)
iSamplesUse <- iSamples;
iSamplesUseList <- split(iSamplesUse, SummarizedExperiment::colData(fc_se[, iSamplesUse])$Group)
maxlen <- max(lengths(iSamplesUseList), na.rm=TRUE);
iSamplesUseList <- lapply(iSamplesUseList, function(i){
   c(i, rep("", length.out=maxlen - length(i)))
})
iSamplesUse_df <- as.data.frame(iSamplesUseList)

kable_coloring(
   caption="Samples used for downstream analysis, organized by group.",
   iSamplesUse_df,
   row.names=FALSE,
   colorSub=sample_color_list)
```

### Sample Correlation {.tabset}

#### Normalized data {.tabset}

##### centered by Group control

Base filename: `correlation_hm`

```{r, correlation_hm, fig.height=12, fig.width=12, dependson=c("apply_norm", "sample_outliers")}
## Define centering control samples
iControl <- head(levels(iGroup), 1);
iControlSamples <- intersect(
   subset(colData(fc_se), colData(fc_se)[[group_colname]] %in% iControl)[[sample_colname]],
   iSamplesUse);
# printDebug("Using control: ", iControl);
# printDebug("Control Samples: ", iControlSamples);

## Center iMatrixQ
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQctr <- centerGeneData(iMatrixQ[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorM <- cor(use="pairwise.complete.obs",
   iMatrixQctr[,iSamplesUse,drop=FALSE]);

## hcluster
# iCorM_hc <- tryCatch({
#    amap::hcluster(iCorM);
# }, error=function(e){
#    printDebug("Error during hcluster(iCorM):");
#    print(e);
#    FALSE;
# });

## ComplexHeatmap
hm_cor_u_se <- heatmap_se(
   assays(fc_se[,iSamples]),
   assay_name=normname,
   correlation=TRUE,
   top_colnames=c("Run", "Group"),
   # row_type=paste0("hits at: ",
   #    gsub("^hit ", "", hm_cutoff)),
   data_type="signal",
   color_max=1,
   controlSamples=iControlSamples,
   control_label="versus control samples",
   rowData_colnames=c("feature_type_winner"),
   show_row_names=FALSE,
   sample_color_list=sample_color_list,
   column_cex=0.4,
   column_split=c("Group"),
   row_split=c("Group"),
   column_title_rot=90)
ComplexHeatmap::draw(hm_cor_u_se,
   column_title=attr(hm_cor_u_se, "hm_title"),
   merge_legends=TRUE)

# old heatmap (to be removed)
hm_cor_u <- ComplexHeatmap::Heatmap(
   iCorM[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=FALSE,
   cluster_columns=FALSE,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_cor_u,
   merge_legends=TRUE,
   column_title=paste0("Counts (", normname, ")"))

```


##### batch-adjusted, global-centered

```{r, cor_hm_b, fig.height=12, fig.width=12, dependson=c("batch_adjust_run", "sample_outliers")}
## Center iMatrixQB
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQBctrGlobal <- centerGeneData(iMatrixQB[,iSamples,drop=FALSE],
   mean=TRUE);

## Correlation matrix
iCorMBglobal <- cor(use="pairwise.complete.obs",
   iMatrixQBctrGlobal[,iSamplesUse,drop=FALSE])

## hcluster
# iCorMBglobal_hc <- tryCatch({
#    amap::hcluster(iCorMBglobal);
# }, error=function(e){
#    printDebug("Error during hcluster(iCorM):");
#    print(e);
#    FALSE;
# });

## ComplexHeatmap
hm_corbg_u <- ComplexHeatmap::Heatmap(
   iCorMBglobal[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=FALSE,
   cluster_columns=FALSE,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_corbg_u,
   merge_legends=TRUE,
   column_title=paste0("Counts (", bnormname, ")"))


```

##### batch-adjusted, centered by Group control

```{r, cor_hm_bg, fig.height=12, fig.width=12, dependson=c("batch_adjust_run", "sample_outliers")}
## Center iMatrixQB
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQBctr <- centerGeneData(iMatrixQB[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorMB <- cor(use="pairwise.complete.obs",
   iMatrixQBctr[,iSamplesUse,drop=FALSE])

## hcluster
# iCorMB_hc <- tryCatch({
#    amap::hcluster(iCorMB);
# }, error=function(e){
#    printDebug("Error during hcluster(iCorM):");
#    print(e);
#    FALSE;
# });

## ComplexHeatmap
hm_corb_u <- ComplexHeatmap::Heatmap(
   iCorMB[iSamplesUse, iSamplesUse, drop=FALSE],
   name="correlation",
   border=TRUE,
   col=colorjam::col_div_xf(1, lens=5),
   cluster_rows=FALSE,
   cluster_columns=FALSE,
   heatmap_legend_param=list(
      color_bar="discrete",
      at=seq(from=-1, to=1, by=0.25),
      border=TRUE),
   top_annotation=ComplexHeatmap::HeatmapAnnotation(
      border=TRUE,
      annotation_legend_param=list(
         border=TRUE),
      df=data.frame(
         Run=Run[iSamplesUse],
         Group=iGroup[iSamplesUse]),
      col=list(Group=group_colors,
         Run=run_colors))
)
ComplexHeatmap::draw(hm_corb_u,
   column_title=paste0("Counts (", bnormname, ")"))

```

### Peak Width by Peak Signal {.tabset}

This section is intended to help evaluate peak width, and peak slice width
when using `mergeSplitCountBed`.

A vertical dotted line is drawn to indicate the MGM threshold.

#### Raw data

```{r, peak_width_by_signal_raw, fig.height=ma_height*1.5, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 4, 2) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQ[, iCol],
      minimum=1,
      newValue=NA);
   iy <- log2(1 + rowData(fc_se[iRows, ])$Length);
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main="",
      # main=iCol,
      yaxt="n");
   # place title atop the plot, using plot panel width
   jamba::drawLabels(preset="top",
      adjPreset="top",
      txt=iCol,
      panelWidth="minimum",
      boxColor=color_sub1[iCol],
      boxBorderColor=jamba::makeColorDarker(color_sub1[iCol]),
      labelCol=jamba::setTextContrastColor(color_sub1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   minorLogTicksAxis(2,
      logBase=2,
      offset=1,
      displayBase=10);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="topleft",
      adjPreset="bottomright",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.0,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```

#### Normalized data


```{r, peak_width_by_signal_norm, fig.height=ma_height*1.5, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 3, 1) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQB[, iCol],
         minimum=1,
         newValue=NA);
   iy <- log2(1 + rowData(fc_se[iRows, ])$Length);
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main="",
      # main=iCol,
      yaxt="n");
   # place title atop the plot, using plot panel width
   jamba::drawLabels(preset="top",
      adjPreset="top",
      txt=iCol,
      panelWidth="minimum",
      boxColor=color_sub1[iCol],
      boxBorderColor=jamba::makeColorDarker(color_sub1[iCol]),
      labelCol=jamba::setTextContrastColor(color_sub1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   minorLogTicksAxis(2,
      logBase=2,
      offset=1,
      displayBase=10);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="topleft",
      adjPreset="bottomright",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.0,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```


**Normal scale y-axis peak width:**

```{r, peak_width_signal_norm2, fig.height=ma_height*1.5, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 3, 1) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQB[, iCol],
         minimum=1,
         newValue=NA);
   iy <- rowData(fc_se[iRows, ])$Length;
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main="");
   # place title atop the plot, using plot panel width
   jamba::drawLabels(preset="top",
      adjPreset="top",
      txt=iCol,
      panelWidth="minimum",
      boxColor=color_sub1[iCol],
      boxBorderColor=jamba::makeColorDarker(color_sub1[iCol]),
      labelCol=jamba::setTextContrastColor(color_sub1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="topleft",
      adjPreset="bottomright",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.0,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
```

### Statistical Comparisons

#### Confirm Replicates

If there is only one replicate per sample group, then
the collection of samples are used to estimate
variability during statistical testing. This procedure
uses the moderated t-test from the `limma` R package,
which uses an improved estimation of variability
by including pooled variability from neighboring
measurements.


```{r, check_reps}
## check for any available contrasts
if (length(contrastNames) == 0) {
   printDebug("There were no statistical contrasts defined, exiting now.");
   knitr::knit_exit(fully=TRUE);
}

##
ctGroup <- table(iGroup[iSamplesUse]);
denovo_controls <<- FALSE;
if (any(ctGroup > 1)) {
   Outcome <- "At least one sample group contains replicates.";
} else {
   Outcome <- "No sample groups contain replicates. Pooled replicates will be created.";
   denovo_controls <<- TRUE;
   fc_se_ctrl <- fc_se_b[,iSamplesUse];
   colData(fc_se_ctrl)[,group_colname] <- "DenovoControl";
   colnames(fc_se_ctrl) <- jamba::makeNames(rep("DenovoControl", ncol(fc_se_ctrl)));
   colData(fc_se_ctrl)[,"Sample"] <- colnames(fc_se_ctrl);
   fc_se_b <- cbind(fc_se_ctrl, fc_se_b);
   iSamplesUse <- colnames(fc_se_b);
   
   ## update design and contrasts
   sedesign <- jamses::groups_to_sedesign(
      data.frame(check.names=FALSE,
         colData(fc_se[, iSamplesUse])[, group_colname, drop=FALSE]),
      factor_order=FACTOR_ORDER,
      max_depth=1);
   iDesign <- sedesign@design;
   if (length(contrastNames) > 0) {
      iContrasts <- tryCatch({limma::makeContrasts(
            contrasts=contrastNames,
            levels=iDesign);
      }, error=function(e){
         jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
            contrast_txt_file);
         print(e);
         sedesign@contrasts;
      })
   } else {
      iContrasts <- sedesign@contrasts;
   }
}
kable_coloring(
   data.frame(Outcome=Outcome),
   row.names=FALSE,
   caption="Samples were evaluated for presence of biological replicates.")
```

#### Design Matrices

```{r, stats_design, dependson=c("batch_adjust_run", "sample_outliers", "check_reps")}
###############################################
# Statistical comparisons

## subset by iSamplesUse
sedesign <- jamses::validate_sedesign(sedesign,
   min_reps=1,
   samples=iSamplesUse,
   verbose=FALSE);
iDesign <- sedesign@design;
if (length(contrastNames) > 0) {
   iContrasts <- tryCatch({limma::makeContrasts(
         contrasts=contrastNames,
         levels=iDesign);
   }, error=function(e){
      jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
         contrast_txt_file);
      print(e);
      sedesign@contrasts;
   });
   sedesign@contrasts <- iContrasts;
} else {
   iContrasts <- sedesign@contrasts;
}

## Optionally remove denovo controls from contrasts if needed
if (denovo_controls) {
   iRemove <- igrep("DenovoControl-|-DenovoControl", colnames(iContrasts));
   if (length(iRemove) > 0) {
      iContrasts <- iContrasts[, -iRemove, drop=FALSE];
      sedesign@contrasts <- iContrasts;
   }
}

## Print design matrix
color_sub_im <- c(`-1`="#2656CC",
   `0`="#DDDDDD",
   `1`="#CC2626");
kable_coloring(
   data.frame(check.names=FALSE,
      iDesign),
   colorSub=color_sub_im,
   caption="Group design matrix") %>%
   kableExtra::column_spec(column=2:(ncol(iDesign) + 1),
      border_left="1px solid #DDDDDD")
## Print contrast matrix
kable_coloring(
   data.frame(check.names=FALSE,
      iContrasts),
   colorSub=color_sub_im,
   caption="Statistical contrast matrix") %>%
   kableExtra::column_spec(column=2:(ncol(iContrasts) + 1),
      border_left="1px solid #DDDDDD")
```

#### All Peaks

Limma uses a "moderated t-test" for statistics, and the method
below enables "limma-voom" which is intended for
use with count-based data.

Note that Voom is disabled when each group contains n=1.

```{r, run_limma, dependson=c("stats_design")}
## Find best available signal
assay_names <- head(intersect(c(bnormname, normname),
   names(assays(fc_se_b))), 1);
## Use all peaks
iPeaks <- rownames(fc_se);

## Determine whether we can use Voom
use_voom <- TRUE;
if (all(colSums(iDesign) == 1)) {
   use_voom <- FALSE;
}

## Stats using all batches
mgm_cutoff <- MGM;
fold_cutoff <- 1.5;
adjp_cutoff <- 0.05;
AtacStats <- tryCatch({
   jamses::se_contrast_stats(
      se=fc_se_b[, iSamplesUse],
      handle_na="none",
      assay_names=assay_names,
      p_cutoff=NULL,
      adjp_cutoff=adjp_cutoff,
      fold_cutoff=fold_cutoff,
      mgm_cutoff=mgm_cutoff,
      use_voom=use_voom,
      igenes=iPeaks,
      isamples=iSamplesUse,
      idesign=iDesign[iSamplesUse, , drop=FALSE],
      icontrasts=iContrasts,
      verbose=FALSE);
}, error=function(e){
   jamba::printDebug("Error during se_contrast_stats():");
   print(e);
   NULL;
});
if (length(AtacStats) > 0) {
   hitdf <- sestats_to_df(AtacStats);
   colnames(hitdf)[1] <- "assay_name";
   kable_coloring(
      hitdf,
      caption="Statistical hits per contrast, for each hit criteria.",
      row.names=FALSE) 
}
```

#### MGM Peaks

The method below is identical to that used above, except that
peaks are filtered by minimum signal first, since we also apply
a minimum signal threshold for candidate peak hits.

Peak mean counts (normalized, batch-adjusted) are required
to be at least this same threshold in at least one sample group to be
included in the moderated t-test.

The default threshold is **4** after `log2(1+x)` transformation,
which corresponds to **15 counts** `(2^4 - 1)` per peak, averaged
across replicates in each sample group. This same threshold
is reflected in the MA-plots above, since the x-axis minimum
value is also **4**.

The max group mean threshold: **`r MGM`**

```{r, run_limma_mgm, dependson=c("stats_design","run_limma","mgm_arg")}
## Use the lmFit1 coefficients directly, which ensures
## we apply the threshold to exactly the same data 
## used in the limma test
if (length(AtacStats) > 0) {
   
   ## TODO: change to use the full stat df
   mgm_colnames <- jamba::vigrep("^mgm[_ .]|[_ .]mean$",
      colnames(AtacStats$stats_df[[1]]));
   if (length(mgm_colnames) > 0) {
      mgm_m <- as.matrix(AtacStats$stats_df[[1]][,mgm_colnames,drop=FALSE]);
      if (length(rownames(mgm_m)) == 0) {
         rownames(mgm_m) <- AtacStats$stats_df[[1]][,1];
      }
      mgm_values <- apply(mgm_m, 1, max, na.rm=TRUE);
      names(mgm_values) <- rownames(mgm_m);
      rm(mgm_m);
      iPeaksMgm <- names(mgm_values)[mgm_values >= head(MGM, 1)];
   } else {
      iPeaksMgm <- iPeaks;
      printDebug("Note that iPeaks are being re-used for some reason. Check upstream data.");
   }
   
   if (length(AtacStats) == 0 || length(iPeaksMgm) == 0) {
      AtacStatsMgm <- NULL;
   } else {
      mgm_n <- length(mgm_values);
      # jamba::printDebug(sep="",
      #    c("Filtered from: ",
      #    format(mgm_n, big.mark=","),
      #    " total peaks to ",
      #    format(length(iPeaksMgm), big.mark=","),
      #    " (", format(length(iPeaksMgm)*100/mgm_n, digits=1),
      #    "%) with log2(1+x) max group mean >= ",
      #    MGM,
      #    " in signal ",
      #    head(names(AtacStats$stats_df), 1)));
      filtered_df <- data.frame(check.names=FALSE,
         `Total peaks:`=format(mgm_n, big.mark=","),
         `Filtered to:`=format(length(iPeaksMgm), big.mark=","),
         `Filtered pct:`=paste0(
            format(length(iPeaksMgm)*100/mgm_n, digits=1), "%"));
      kable_coloring(
         filtered_df,
         row.names=FALSE,
         caption=paste0(
            "log2( 1 + x ) of max group mean >= ",
            head(MGM, 1),
            " in signal ",
            head(names(AtacStats$stats_df), 1))) %>%
      kableExtra::column_spec(column=2:(ncol(filtered_df)),
         border_left="1px solid #DDDDDD")
      
      ## Stats using all batches
      fold_cutoff <- c(1.5, 1.5);
      adjp_cutoff <- c(1, 0.05);
      p_cutoff <- c(0.05, 0.05);
      AtacStatsMgm <- tryCatch({
         se_contrast_stats(se=fc_se_b[iPeaksMgm, iSamplesUse],
            handle_na="none",
            assay_names=assay_names,
            p_cutoff=p_cutoff,
            adjp_cutoff=adjp_cutoff,
            fold_cutoff=fold_cutoff,
            mgm_cutoff=c(MGM),
            use_voom=TRUE,
            igenes=iPeaksMgm,
            isamples=iSamplesUse,
            idesign=iDesign[iSamplesUse, , drop=FALSE],
            icontrasts=iContrasts,
            verbose=FALSE);
      }, error=function(e){
         jamba::printDebug("Error during allNormStatsTests():");
         print(e);
         NULL;
      });
   }
   if (length(AtacStatsMgm) > 0) {
      hitdf <- sestats_to_df(AtacStatsMgm);
      colnames(hitdf)[1] <- "assay_name";
      kable_coloring(
         hitdf,
         caption=paste0("Statistical hits per contrast, for each hit criteria ",
            "(MGM >= ", MGM, ")"),
         row.names=FALSE) 
   }
}
```

### Volcano Plots {.tabset}

#### All Peaks {.tabset}

##### using adj.P.Val

```{r, volcano_plots, fig.height=12, fig.width=12, dependson=c("run_limma"), results='asis'}
options("warn"=-1);
if (length(AtacStats) > 0) {
   iSigs <- dimnames(AtacStats$hit_array)[[3]];
   iHits <- dimnames(AtacStats$hit_array)[[1]];
   iCons <- dimnames(AtacStats$hit_array)[[2]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iCon in iCons) {
         cat(paste0("\n###### ", iCon, "\n\n"));
         iDF <- AtacStats$stats_dfs[[iSig]][[iCon]];
         par("mar"=c(7, 7, 4, 3));
         vp1adjp <- volcano_plot(iDF,
            verbose=FALSE,
            sig_cutoff=0.05,
            fold_cutoff=1.5,
            expr_cutoff=MGM,
            main=iCon,
            submain="All peaks (adj.P.Val)",
            sig_colname="adj.P.Val");
         par("mar"=c(5, 4, 4, 2) + 0.1);
      }
   }
}
```

##### using raw P-Value

```{r, volcano_plots_raw, fig.height=12, fig.width=12, dependson=c("run_limma"), results='asis'}
options("warn"=-1);
if (length(AtacStats) > 0) {
   iSigs <- dimnames(AtacStats$hit_array)[[3]];
   iHits <- dimnames(AtacStats$hit_array)[[1]];
   iCons <- dimnames(AtacStats$hit_array)[[2]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iCon in iCons) {
         cat(paste0("\n###### ", iCon, "\n\n"));
         iDF <- AtacStats$stats_dfs[[iSig]][[iCon]];
         par("mar"=c(7, 7, 4, 3));
         vp1p <- volcano_plot(iDF,
            verbose=FALSE,
            sig_cutoff=0.05,
            fold_cutoff=1.5,
            expr_cutoff=MGM,
            main=iCon,
            submain="All peaks (P-Value)",
            sig_colname="P.Value");
         par("mar"=c(5, 4, 4, 2) + 0.1);
      }
   }
}
```


#### MGM Peaks {.tabset}

##### using adj.P.Val

Volcano plots below were created using the limma results
after subsetting peaks for max group mean >= 
**`r MGM`**.

```{r, volcano_plots_mgm, fig.height=12, fig.width=12, dependson=c("run_limma_mgm"), results='asis'}
options("warn"=-1);
if (length(AtacStatsMgm) > 0) {
   iSigs <- dimnames(AtacStatsMgm$hit_array)[[3]];
   iHits <- dimnames(AtacStatsMgm$hit_array)[[1]];
   iCons <- dimnames(AtacStatsMgm$hit_array)[[2]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iCon in iCons) {
         cat(paste0("\n###### ", iCon, "\n\n"));
         iDF <- AtacStatsMgm$stats_dfs[[iSig]][[iCon]];
         par("mar"=c(7, 7, 4, 3))
         tryCatch({
            vp2adjp <- volcano_plot(iDF,
               verbose=FALSE,
               sig_cutoff=0.05,
               fold_cutoff=1.5,
               expr_cutoff=MGM,
               main=iCon,
               submain=paste0("Subset peaks mgm >= ", MGM, " (adj.P.Val)"),
               sig_colname="adj.P.Val");
         }, error=function(e){
            jamba::printDebug("Error during volcano_plot(), skipping.");
            jamba::printDebug("iSig:", iSig);
            jamba::printDebug("iCon:", iCon);
            print(e);
         });
         par("mar"=c(5, 4, 4, 2) + 0.1);
      }
   }
}
```

##### using raw P-Value

```{r, volcano_plots_mgm_raw, fig.height=12, fig.width=12, dependson=c("run_limma_mgm"), results='asis'}
options("warn"=-1);
if (length(AtacStatsMgm) > 0) {
   iSigs <- dimnames(AtacStatsMgm$hit_array)[[3]];
   iHits <- dimnames(AtacStatsMgm$hit_array)[[1]];
   iCons <- dimnames(AtacStatsMgm$hit_array)[[2]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iCon in iCons) {
         cat(paste0("\n###### ", iCon, "\n\n"));
         iDF <- AtacStatsMgm$stats_dfs[[iSig]][[iCon]];
         par("mar"=c(7, 7, 4, 3))
         tryCatch({
            vp2p <- volcano_plot(iDF,
               verbose=FALSE,
               sig_cutoff=0.05,
               fold_cutoff=1.5,
               expr_cutoff=MGM,
               main=iCon,
               submain=paste0("Subset peaks mgm >= ", MGM, " (P-Value)"),
               sig_colname="P.Value");
         }, error=function(e){
            jamba::printDebug("Error during volcano_plot(), skipping.");
            jamba::printDebug("iSig:", iSig);
            jamba::printDebug("iCon:", iCon);
            print(e);
         });
         par("mar"=c(5, 4, 4, 2) + 0.1);
      }
   }
}
```

### MGM Distribution

These plots represent the distribution of max group mean values (MGM)
among those peaks which are considered statistical hits using
the adjust P-value and fold change criteria.

```{r, distr_mgm, fig.height=10, fig.width=10, dependson=c("run_limma"), results='asis'}
if (length(AtacStats) > 0) {
   iSigs <- dimnames(AtacStats$hit_array)[[3]];
   iHits <- dimnames(AtacStats$hit_array)[[1]];
   iCons <- dimnames(AtacStats$hit_array)[[2]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iCon in iCons) {
         cat(paste0("\n###### ", iCon, "\n\n"));
         iDF <- AtacStats$stats_dfs[[iSig]][[iCon]];
         mgmcol <- vigrep("^mgm ", colnames(iDF));
         sigcol <- vigrep("^P.Value ", colnames(iDF));
         foldcol <- vigrep("^fold ", colnames(iDF));
         sigp <- which(abs(iDF[[foldcol]]) >= 1.5 &
               iDF[[sigcol]] <= 0.05);
         if (length(sigp) > 2) {
            tryCatch({
               jamba::plotPolygonDensity(
                  main=paste0("Max Group Mean for ",
                     formatInt(length(sigp)),
                     " statistical hits\n",
                     "with P.Value <= 0.05 and abs(fold) >= 1.5\n",
                     gsub("^mgm ", "", mgmcol)),
                  xlab=mgmcol,
                  iDF[sigp,mgmcol])
               abline(v=MGM, col="black", lwd=2);
               abline(v=MGM, col="gold", lty=2, lwd=2);
            }, error=function(e){
               jamba::printDebug("plot skipped, iSig: ",
                  iSig,
                  ", iCon: ",
                  iCon,
                  ", error:");
               print(e);
            })
         }
         
         sigcol <- vigrep("^adj.P.Val ", colnames(iDF));
         sigadjp <- which(abs(iDF[[foldcol]]) >= 1.5 &
               iDF[[sigcol]] <= 0.05);
         if (length(sigadjp) > 2) {
            tryCatch({
               jamba::plotPolygonDensity(
                  main=paste0("Max Group Mean for ",
                     formatInt(length(sigadjp)),
                     " statistical hits\n",
                     "with adj.P.Val <= 0.05 and abs(fold) >= 1.5\n",
                     gsub("^mgm ", "", mgmcol)),
                  xlab=mgmcol,
                  iDF[sigadjp,mgmcol])
               abline(v=MGM, col="black", lwd=2);
               abline(v=MGM, col="gold", lty=2, lwd=2);
            }, error=function(e){
               jamba::printDebug("plot skipped, iSig: ",
                  iSig,
                  ", iCon: ",
                  iCon,
                  ", error:");
               print(e);
            })
         }
      }
   }
}
```



### Venn diagrams {.tabset}

#### All Peaks

`r if (length(AtacStats) > 0 && length(dimnames(AtacStats$hit_array)[[2]]) > 1) {("Venn diagrams are created for all pairs of results, and indicate the direction of shared hits.")} else {("Venn diagrams are skipped since there are not two or more comparisons.")}`


```{r, venn_pairs, fig.height=10, fig.width=10, dependson=c("run_limma"), results='asis'}
if (length(AtacStats) > 0 && length(dimnames(AtacStats$hit_array)[[2]]) > 1) {
   pair_l <- unlist(recursive=FALSE,
      apply(combn(dimnames(AtacStats$hit_array)[[2]], 2), 2, list));
   contrast_colors <- group2colors(dimnames(AtacStats$hit_array)[[2]]);
   
   par("mar"=c(5, 4, 4, 2) + 0.1);
   iSigs <- dimnames(AtacStats$hit_array)[[3]];
   iHits <- dimnames(AtacStats$hit_array)[[1]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iHit in iHits) {
         cat(paste0("\n###### ", iHit, " {.tabset}\n\n"));
         for (iCon in pair_l) {
            cat(paste0("\n####### ", iCon, "\n\n"));
            tryCatch({
               setlist <- AtacStats$hit_array[iHit, iCon, iSig];
               names(setlist) <- gsub("-", "-<br>\n", names(setlist));
               venndir::venndir(setlist,
                  overlap_type="each",
                  offset=c(0, -1),
                  show_segments=FALSE,
                  label_preset="main outside",
                  set_colors=contrast_colors[iCon],
                  main=paste0("signal:", iSig,
                     "\nthreshold:", iHit));
            }, error=function(e){
               jamba::printDebug("venndir::venndir() ",
                  comment=FALSE,
                  "skipping plot for iSig:", iSig,
                  ", iHit:", iHit,
                  ", iCon:", iCon);
            });
         }
      }
   }
}
```

#### MGM Peaks

`r if (length(AtacStatsMgm) > 0 && length(dimnames(AtacStatsMgm$hit_array)[[2]]) > 1) {("")} else {("Venn diagrams for subset peaks are skipped since there are not two or more comparisons.")}`

```{r, venn_pairs_mgm, fig.height=10, fig.width=10, dependson=c("run_limma_mgm"), results='asis'}
if (length(AtacStatsMgm) > 0 && length(dimnames(AtacStatsMgm$hit_array)[[2]]) > 1) {
   pair_l_mgm <- unlist(recursive=FALSE,
      apply(combn(dimnames(AtacStatsMgm$hit_array)[[2]], 2), 2, list));
   contrast_colors <- group2colors(dimnames(AtacStatsMgm$hit_array)[[2]]);
   
   par("mar"=c(5, 4, 4, 2) + 0.1);
   iSigs <- dimnames(AtacStats$hit_array)[[3]];
   iHits <- dimnames(AtacStats$hit_array)[[1]];
   for (iSig in iSigs) {
      cat(paste0("\n##### ", iSig, " {.tabset}\n\n"));
      for (iHit in iHits) {
         cat(paste0("\n###### ", iHit, " {.tabset}\n\n"));
         for (iCon in pair_l) {
            cat(paste0("\n####### ", iCon, "\n\n"));
            tryCatch({
               setlist <- AtacStatsMgm$hit_array[iHit, iCon, iSig];
               names(setlist) <- gsub("-", "-<br>\n", names(setlist));
               venndir::venndir(setlist,
                  overlap_type="each",
                  offset=c(0, -1),
                  show_segments=FALSE,
                  label_preset="main outside",
                  set_colors=contrast_colors[iCon],
                  main=paste0("signal:", iSig,
                     "\nthreshold:", iHit));
            }, error=function(e){
               jamba::printDebug("venndir::venndir() ",
                  comment=FALSE,
                  "skipping subset peaks plot for iSig:", iSig,
                  ", iHit:", iHit,
                  ", iCon:", iCon);
            });
         }
      }
   }
}
```


### Annotate Peaks

The GTF file was used to define genomic regions. Each peak is annotated
with only one of the following categories, assigned in order.
Therefore a peak that overlaps a promoter and an exon is only
annotated "promoter".

* **promoters**: `r upstream_promoter` upstream, and `r downstream_promoter`
downstream each transcript TSS
* **TTS**: `r upstream_tts` upstream, and `r downstream_tts` downstream
the last stranded nucleotide position in each transcript.
* **exon**: each stranded exon per transcript
* **intron**: non-exonic region within a stranded transcript region
* **extragenic**: all other regions.


#### Define Genome Regions

This section converts the given GTF file to a set of genome regions.

The GTF input file:

`r jamba::printDebug(unlist(strsplit(GTF, "/")), collapse="/", sep="", comment=FALSE)`


A cached RData file is stored for future re-use, using this file extension:

> `.genome_regions.RData`


```{r, define_genome_regions}
genome_regions <- genomic_regions_from_gtf(gtf=GTF,
   upstream_promoter=upstream_promoter,
   downstream_promoter=downstream_promoter,
   upstream_tts=upstream_tts,
   downstream_tts=downstream_tts,
   detectedTx=detectedTx,
   detectedGenes=detectedGenes,
   save_rdata=TRUE,
   save_txdb=TRUE);
```


#### Annotate by Genome Region


```{r, peak_genes_fn}
## Annotate ATAC peaks
fc_gr <- as(rowData(fc_se), "GRanges");
values(fc_gr) <- renameColumn(values(fc_gr),
   from="Geneid",
   to="Gene");

# Call function that wraps the logic together
fc_gr <- annotate_gr_by_genome_region(
   gr=fc_gr,
   genome_regions=genome_regions,
   mask_regions=MASK,
   verbose=FALSE);

# now propagate annotations into the fc_se and fc_se_b objects
region_colnames <- provigrep(c(
   "gene_(name|id|feature)|feature_type|nearest_gene",
   "broad_peak|num_slices"),
   colnames(values(fc_gr)))
gr_match <- match(rownames(fc_se), values(fc_gr)$Gene);
if (!any(is.na(gr_match))) {
   rowData(fc_se)[,region_colnames] <- data.frame(check.names=FALSE,
      values(fc_gr)[gr_match, region_colnames, drop=FALSE])
}
gr_match <- match(rownames(fc_se_b), values(fc_gr)$Gene);
if (!any(is.na(gr_match))) {
   rowData(fc_se_b)[,region_colnames] <- data.frame(check.names=FALSE,
      values(fc_gr)[gr_match, region_colnames, drop=FALSE])
}

feature_type_colors <- colorjam::group2colors(unique(
   as.character(values(fc_se)$feature_type_winner)));
sample_color_list$feature_type_winner <- feature_type_colors;

# show the full set of colors
# showColors(sample_color_list)

```

#### Incidence Matrix by feature_type

The plot below shows the association of each peak with one or more
`feature_type` values.

Along the diagonal, it represents a count of overlaps for all
peaks over each `feature_type` value.

The cells off the diagonal indicate the number of peaks
that overlap two `feature_type` values. For example, `"Promoter"`
often also overlaps `"Exon"` particularly when the peak covers
the TSS itself. Similarly, `"Exon"` often overlaps `"Intron"`
partly because the average exon is only 100 bases wide, which
is usually smaller than most peaks.

```{r, peak_featuretype_im}
## For fun make an logical matrix of feature_type
ftcols <- c("Promoter", "TTS", "exon", "intron", "extragenic")
fc_typem <- as.data.frame(lapply(nameVector(ftcols), function(i){
   j <- grepl(i, values(fc_gr)$feature_type);
   names(j) <- values(fc_gr)$Gene;
   j;
}));
m1 <- as.matrix(fc_typem);
m1_counts <- (t(m1) %*% (m1));
m1_df <- as.data.frame(m1_counts);
for (i1 in colnames(m1_df)) {
   m1_df[,i1] <- formatInt(m1_df[,i1])
}
m1_values <- sort(unique(as.vector(m1_counts)));
names(m1_values) <- formatInt(m1_values)
color_sub_imct <- colorjam::vals2colorLevels(
   m1_values,
   lens=10);
kable_coloring(as.data.frame(m1_counts),
   colorSub=color_sub_imct,
   color_cells=TRUE,
   format.args=list(big.mark = ",",
      scientific=FALSE),
   row.names=TRUE) %>%
   kable_styling(bootstrap_options=c("striped","bordered")) %>%
   row_spec(0, background="#DDDDDD", color="#333333", bold=TRUE) %>%
   column_spec(1, background="#DDDDDD", color="#333333", bold=TRUE)

```

### Pie Charts by Region {.tabset}

#### Hits, All Peaks

A pie chart is created for each statistical comparison, showing
the number of peaks overlapping each genome region. A peak
is only assigned one genome region, with preference in order:
promoter, TTS, exon, intron, extragenic.

For comparison, the pie chart using all peaks is also shown.

Base filename: `peak_pies_all`

```{r, peak_pies_all, fig.height=15, fig.width=15}
feature_type_winner <- values(fc_gr)$feature_type_winner;
names(feature_type_winner) <- values(fc_gr)$Gene;

gg2 <- NULL;
gg2_width <- 14;
gg2_height <- 8;
if (length(AtacStats) > 0 && max(lengths(AtacStats$hit_array)) > 0) {
   gg2 <- gg_pie_by_feature_type(hit_array=AtacStats$hit_array,
      colorSub=sample_color_list$feature_type_winner,
      add_peak_list=list(`MGM-Filtered Peaks`=iPeaksMgm,
         `All Peaks`=iPeaks),
      do_pie=TRUE,
      feature_type_winner=feature_type_winner);
   jdf <- gg_pie_by_feature_type(hit_array=AtacStats$hit_array,
      colorSub=sample_color_list$feature_type_winner,
      add_peak_list=list(`MGM-Filtered Peaks`=iPeaksMgm,
         `All Peaks`=iPeaks),
      do_pie=FALSE,
      feature_type_winner=feature_type_winner);
      #PeakSet="All Peaks Tested",
   gg2_num <- attr(gg2, "panel_count");
   gg2_width <- 10 + 7 * (gg2_num > 1);
   gg2_height <- ceiling(gg2_num / 2) * 7;
}
```

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapsePie1" role="button" aria-expanded="false" aria-controls="collapsePie1">(Table summary...)</a></p>
<div class="collapse" id="collapsePie1"><div class="card card-body">

```{r, peak_pies_gg_all, fig.height=gg2_height, fig.width=gg2_width}
if (length(gg2) > 0) {
   print(gg2);
   
   # print table format
   kable_coloring(jdf[, 1:10, drop=FALSE],
      row.names=FALSE,
      colorSub=sample_color_list$feature_type_winner,
      full_width=TRUE,
      caption="Data used for the pie charts above:") %>%
      kable_styling(bootstrap_options=c("striped","bordered")) %>%
      kableExtra::column_spec(column=seq_len(10),
         extra_css="white-space: nowrap;") %>%
      row_spec(0, background="#DDDDDD");

}
```

</div></div>

#### Hits, MGM Peaks

The pie charts below reflect the peaks after filtering
for the max group mean threshold: **`r MGM`**

Base filename: `peak_pies_mgm`

```{r, peak_pies_mgm, fig.height=15, fig.width=15}
gg2 <- NULL;
gg2_width <- 14;
gg2_height <- 8;
if (length(AtacStatsMgm) > 0 && max(lengths(AtacStatsMgm$hit_array)) > 0) {
   gg2 <- gg_pie_by_feature_type(hit_array=AtacStatsMgm$hit_array,
      add_peak_list=list(`MGM-Filtered Peaks`=iPeaksMgm,
         `All Peaks`=iPeaks),
      do_pie=TRUE,
      feature_type_winner=feature_type_winner);
      #PeakSet=paste0("Subset Peaks Mgm", MGM),
   jdf <- gg_pie_by_feature_type(hit_array=AtacStatsMgm$hit_array,
      add_peak_list=list(`MGM-Filtered Peaks`=iPeaksMgm,
         `All Peaks`=iPeaks),
      do_pie=FALSE,
      feature_type_winner=feature_type_winner);
   gg2_num <- attr(gg2, "panel_count");
   gg2_width <- 10 + 7 * (gg2_num > 1);
   gg2_height <- ceiling(gg2_num / 2) * 7;
}
```

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapsePie2" role="button" aria-expanded="false" aria-controls="collapsePie2">(Table summary...)</a></p>
<div class="collapse" id="collapsePie2"><div class="card card-body">

```{r, peak_pies_gg_mgm, fig.height=gg2_height, fig.width=gg2_width}
if (length(gg2) > 0) {
   print(gg2);
   
   # print table format
   kable_coloring(jdf[, 1:10, drop=FALSE],
      row.names=FALSE,
      colorSub=sample_color_list$feature_type_winner,
      full_width=TRUE,
      caption="Data used for the pie charts above:") %>%
      kable_styling(bootstrap_options=c("striped","bordered")) %>%
      kableExtra::column_spec(column=seq_len(10),
         extra_css="white-space: nowrap;") %>%
      row_spec(0, background="#DDDDDD");
}
```

</div></div>

### Broad Peaks

Peaks are re-numbered so that contiguous peaks are numbered
in column `"broak_peak"` to indicate when one broad peak
has been sliced by `mergeSplitCountBed` into multiple pieces.
Each peak slice is labeled in `"broak_peak_number"` with
a suffix to indicate the slice number.

This process makes it feasible to count the number of contiguous enriched
regions, and to evaluate the statistical hits across different slices
in the same region.

For example, in one `"broad_peak"`, one slice may have enriched signal,
while another slice may be reduced. The total signal in the `"broad_peak"`
may remain constant even while there is significant change in the
detected chromatin signal.

> *TODO: This section may optionally allow a gap distance between adjacent*
> *peaks, consistent with the settings used in `mergeSplitCountBed`.*

```{r, peak_renumbering}
fc_gr_red <- GenomicRanges::reduce(fc_gr);
values(fc_gr_red)$broad_peak <- makeNames(
   rep("broad", length(fc_gr_red)),
   doPadInteger=TRUE,
   renameOnes=TRUE,
   suffix="_");
fc_gr <- annotateGRfromGR(fc_gr,
   fc_gr_red);
values(fc_gr)$broad_peak_number <- makeNames(
   values(fc_gr)$broad_peak,
   suffix="_",
   doPadInteger=TRUE,
   renameOnes=TRUE);
fc_gr_bp_ct <- table(values(fc_gr)$broad_peak);
values(fc_gr)$num_slices <- as.vector(fc_gr_bp_ct[values(fc_gr)$broad_peak]);
rowData(fc_se)$broad_peak <- values(fc_gr)$broad_peak;
rowData(fc_se)$broad_peak_number <- values(fc_gr)$broad_peak_number;
rowData(fc_se)$broad_peak_slices <- fc_gr_bp_ct[values(fc_gr)$broad_peak];
```


#### Broad Peaks Statistical Hits

Statistical hits are tabulate by unique `"broad_peak"` values.

There are two potential outcomes of this process:

1. **Significant broad_peak hits are nearly identical to individual peak hits.**

   * This result suggests most peaks are not sliced, or that
   significantly changing peaks are not regions broader than the slice width.

2. **Significant broad_peak hits are much smaller than individual peak hits.**

   * This result suggests most peaks are sliced, or that
   significantly changing peaks tend to be regions broader than
   the slice width.


```{r, broad_hit_counts}
if (length(AtacStats) > 0) {
   
   broad_hit_counts <- lapply(nameVector(dimnames(AtacStats$hit_array)[[1]]), function(p){
      sapply(nameVector(dimnames(AtacStats$hit_array)[[2]]), function(i){
         k <- AtacStats$hit_array[p,i,1][[1]];
         k_bp <- values(fc_gr[names(k)])$broad_peak;
         c(peak_hits=length(k_bp),
            broad_peak_hits=length(table(k_bp)));
      })
   });
   jamba::printDebug("broad_hit_counts:");
   broad_hit_counts;
}
   
if (length(AtacStatsMgm) > 0) {
   broad_hit_counts_mgm <- lapply(nameVector(dimnames(AtacStatsMgm$hit_array)[[1]]), function(p){
      sapply(nameVector(dimnames(AtacStatsMgm$hit_array)[[2]]), function(i){
         k <- AtacStatsMgm$hit_array[p,i,1][[1]];
         k_bp <- values(fc_gr[names(k)])$broad_peak;
         c(peak_hits=length(k_bp),
            broad_peak_hits=length(table(k_bp)));
      })
   });
   jamba::printDebug("broad_hit_counts ",
      c("(", "MGM", ">=", MGM, ")"),
      ":", sep="");
   broad_hit_counts_mgm;

}
## Briefly compare All peak hits to Mgm peak hits
## signedVennPlotSets(list(mgm4=AtacStatsMgm$hit_array[1,2,1][[1]]
## all=AtacStats$hit_array[1,2,1][[1]]))

```

### Heatmap of Hits {.tabset}

#### Replicates {.tabset}

A series of heatmaps is shown below, one heatmap for each of the
different statistical comparisons and hit cutoff thresholds used.

##### All Peaks

Base filename: `heatmap_hits_all`

```{r, heatmap_hits_all, fig.height=14, fig.width=4 + sqrt(length(iSamplesUse)) * 1.5}
#
peak_hits_q <- NULL;
if (length(AtacStats) > 0) {
   hm_cutoffs <- dimnames(AtacStats$hit_array)$Cutoffs;
   
   for (hm_cutoff in hm_cutoffs) {
      hm_hitlist <- hit_array_to_list(AtacStats,
         assay_names=assay_names,
         cutoff_names=hm_cutoff)
      hm_hits_use <- setdiff(unique(unlist(lapply(hm_hitlist, names))), NA)
      if (all(length(hm_hits_use) == 0) ||
            length(hm_hits_use) >= 20000) {
         # no heatmap
      } else {
         hm_all <- tryCatch({
            heatmap_se(fc_se_b[, iSamplesUse],
               assay_name=assay_names,
               sestats=AtacStats,
               row_type=paste0("hits at: ",
                  gsub("^hit ", "", hm_cutoff)),
               data_type="signal",
               color_max=3,
               controlSamples=iControlSamples,
               control_label="versus control samples",
               rowData_colnames=c("feature_type_winner"),
               row_split="feature_type_winner",
               show_row_names=FALSE,
               top_colnames=c("Run", "Group"),
               sample_color_list=sample_color_list,
               column_cex=0.4,
               column_split=c("Group"),
               column_title_rot=90)
         }, error=function(e){
            NULL
         });
         if (length(hm_all) > 0) {
            hm_all_d <- ComplexHeatmap::draw(hm_all,
               column_title=attr(hm_all, "hm_title"),
               merge_legends=TRUE)
            peak_hits_q <- unlist(unname(heatmap_row_order(hm_all_d)));
         }
      }
   }
}

```

##### MGM Peaks

Base filename: `heatmap_hits_mgm`

```{r, heatmap_hits_mgm, fig.height=14, fig.width=4 + sqrt(length(iSamplesUse)) * 1.5}
#
peak_hits_mgm_q <- NULL;
if (length(AtacStatsMgm) > 0) {
   hm_cutoffs <- dimnames(AtacStatsMgm$hit_array)$Cutoffs;

   for (hm_cutoff in hm_cutoffs) {
      hm_hitlist <- hit_array_to_list(AtacStatsMgm,
         assay_names=assay_names,
         cutoff_names=hm_cutoff)
      hm_hits_use <- setdiff(unique(unlist(lapply(hm_hitlist, names))), NA)
      if (all(length(hm_hits_use) == 0) ||
            length(hm_hits_use) >= 20000) {
         # no heatmap
      } else {
         hm_mgm <- tryCatch({
            heatmap_se(fc_se_b[, iSamplesUse],
               assay_name=assay_names,
               sestats=AtacStatsMgm,
               row_type=paste0AtacStatsMgm,
               data_type="signal",
               color_max=3,
               controlSamples=iControlSamples,
               control_label="versus control samples",
               rowData_colnames=c("feature_type_winner"),
               row_split="feature_type_winner",
               show_row_names=FALSE,
               top_colnames=c("Run", "Group"),
               sample_color_list=sample_color_list,
               column_cex=0.6,
               column_split=c("Group"),
               column_title_rot=90)
         }, error=function(e){
            NULL
         });
         if (length(hm_mgm) > 0) {
            hm_mgm_d <- ComplexHeatmap::draw(hm_mgm,
               column_title=attr(hm_mgm, "hm_title"),
               merge_legends=TRUE)
            peak_hits_mgm_q <- unlist(unname(heatmap_row_order(hm_mgm_d)));
         }
      }
   }
}

if (length(AtacStatsMgm) > 0) {
   use_sestats <- AtacStatsMgm;
}

```


#### Group Means {.tabset}

##### All Peaks

Base filename: `heatmap_groups_all`

```{r, heatmap_groups_all, fig.height=14, fig.width=4 + sqrt(length(unique(iGroup[iSamplesUse]))) * 1.5}
# prepare group mean SE
fc_se_b_grp <- se_collapse_by_column(fc_se_b,
   columns=iSamplesUse,
   column_groups="Group",
   assay_names=assay_names)

# now repeat steps above
if (length(AtacStats) > 0) {
   hm_cutoffs <- dimnames(AtacStats$hit_array)$Cutoffs;
   
   for (hm_cutoff in hm_cutoffs) {
      hm_hitlist <- hit_array_to_list(AtacStats,
         assay_names=assay_names,
         cutoff_names=hm_cutoff)
      hm_hits_use <- setdiff(unique(unlist(lapply(hm_hitlist, names))), NA)
      if (all(length(hm_hits_use) == 0) ||
            length(hm_hits_use) >= 20000) {
         # no heatmap
      } else {
         hm_all_grp <- tryCatch({
            heatmap_se(fc_se_b_grp[, iSamplesUse],
               assay_name=assay_names,
               sestats=AtacStats,
               row_type=paste0("hits at: ",
                  gsub("^hit ", "", hm_cutoff)),
               data_type="signal",
               color_max=3,
               controlSamples=unique(colData(
                  fc_se[,iControlSamples,drop=FALSE])[[group_colname]]),
               control_label="versus control group",
               rowData_colnames=c("feature_type_winner"),
               row_split="feature_type_winner",
               show_row_names=FALSE,
               top_colnames=c("Group"),
               sample_color_list=sample_color_list,
               column_cex=0.4,
               # column_split=c("Group"),
               column_title_rot=90)
         }, error=function(e){
            NULL
         });
         if (length(hm_all_grp) > 0) {
            hm_all_grp_d <- ComplexHeatmap::draw(hm_all_grp,
               column_title=attr(hm_all_grp, "hm_title"),
               merge_legends=TRUE)
         }
      }
   }
}
```

##### MGM Peaks

Base filename: `heatmap_groups_mgm`

```{r, heatmap_groups_mgm, fig.height=14, fig.width=4 + sqrt(length(unique(iGroup[iSamplesUse]))) * 1.5}

# now repeat steps above
if (length(AtacStatsMgm) > 0) {
   hm_cutoffs <- dimnames(AtacStatsMgm$hit_array)$Cutoffs;
   
   for (hm_cutoff in hm_cutoffs) {
      hm_hitlist <- hit_array_to_list(AtacStatsMgm,
         assay_names=assay_names,
         cutoff_names=hm_cutoff)
      hm_hits_use <- setdiff(unique(unlist(lapply(hm_hitlist, names))), NA)
      if (all(length(hm_hits_use) == 0) ||
            length(hm_hits_use) >= 20000) {
         # no heatmap
      } else {
         hm_all_grp <- tryCatch({
            heatmap_se(fc_se_b_grp[, iSamplesUse],
               assay_name=assay_names,
               sestats=AtacStatsMgm,
               row_type=paste0("hits from MGM sub-analysis at: ",
                  gsub("^hit ", "", hm_cutoff)),
               data_type="signal",
               color_max=3,
               controlSamples=unique(colData(
                  fc_se[,iControlSamples,drop=FALSE])[[group_colname]]),
               control_label="versus control group",
               rowData_colnames=c("feature_type_winner"),
               row_split="feature_type_winner",
               show_row_names=FALSE,
               top_colnames=c("Group"),
               sample_color_list=sample_color_list,
               column_cex=0.4,
               # column_split=c("Group"),
               column_title_rot=90)
         }, error=function(e){
            NULL
         });
         if (length(hm_all_grp) > 0) {
            hm_all_grp_d <- ComplexHeatmap::draw(hm_all_grp,
               column_title=attr(hm_all_grp, "hm_title"),
               merge_legends=TRUE)
         }
      }
   }
}
```



### ATAC Mode Annotation (Optional)

**`r if (ATAC && length(AtacStats) > 0 && exists("iPeaksMgm") && length(iPeaksMgm) > 0) {("ATAC-mode is enabled.")} else {("ATAC-mode is disabled.")}`**

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseAtac1" role="button" aria-expanded="false" aria-controls="collapseAtac1">(Expand...)</a></p>
<div class="collapse" id="collapseAtac1"><div class="card card-body">

This workflow is enabled only when environment variable `ATAC=1`,
and when stats have been performed, and when the max group mean
filter has been applied.

This workflow defines a subset of genes with "active promoters" if
any ATAC-seq peaks are found in the `"Promoter"` region of the gene.
For this purpose, we require the peak to meet the MGM threshold
for detection above background.

The purpose of this step is to reduce the total genes defined in the GTF
to the subset of genes that were observed to have "active promoter",
so the subset of GTF genes can be used to annotate peaks.
The result is improved accuracy in assigning genes to peaks,
by removing GTF entries that are not associated with active
transcription in the given experiment.

In practice, most actively transcribed genes, or genes that are
poised to become actively transcribed, have open chromatin
seen by ATAC-seq across most or all conditions of an experiment.

Peaks are annotated to the nearese MGM-filtered gene TSS, as
defined by the promoter regions, along with the distance to that region.

</div></div>

```{r, peak_promoters}
mgmpg <- NULL;
mgmpd <- NULL;
if (ATAC && length(AtacStats) > 0 && exists("iPeaksMgm") && length(iPeaksMgm) > 0) {
   
   ## iPeaksMgm that are promoters
   ## - call these active promoters
   ##
   ## annotate all peaks by nearest active promoter
   ## - promoter,tts,intron,exon -- gene_name intersected
   ## - all cases, nearest mgm4 promoter, and its gene_name
   ## - distance to nearest mgm4 promoter
   
   promoters_gr <- subset(genome_regions,
      grepl("promoter", 
         ignore.case=TRUE, 
         feature_type));
   #values(promoters_gr)[,"gene_strand"] <- as.character(strand(promoters_gr));
   
   upstream <- upstream_promoter;
   downstream <- downstream_promoter;
   
   tryCatch({
      mgm_promoter_gr <- IRanges::subsetByOverlaps(promoters_gr,
         fc_gr[iPeaksMgm],
         ignore.strand=TRUE);
      mgm_tss_gr <- promoter_to_tss(mgm_promoter_gr[,c("gene_name", "gene_id")],
         upstream=upstream,
         downstream=downstream);
      
      mgmpg <- paste0("mgm", MGM, "_annoTSS_gene_name");
      mgmpgid <- paste0("mgm", MGM, "_annoTSS_gene_id");
      mgmpd <- paste0("mgm", MGM, "_annoTSS_distance");
      
      ## nearest mgm_annoTSS distance
      fc_gr_dist <- GenomicRanges::distanceToNearest(fc_gr,
         ignore.strand=TRUE,
         select="all",
         mgm_tss_gr);
      #220020
      ## Add comma-delimited values per column as needed
      genecol1 <- head(provigrep(c("^Geneid", "^Gene"), colnames(values(fc_gr))), 1);
      fc_gr_dist_df <- renameColumn(
         data.frame(
            as.data.frame(values(fc_gr[queryHits(fc_gr_dist)])[,genecol1,drop=FALSE]),
            gene_distance=values(fc_gr_dist)$distance,
            as.data.frame(values(mgm_tss_gr[subjectHits(fc_gr_dist)])[,c("gene_name", "gene_id"),drop=FALSE])
         ),
         from=c("gene_name", "gene_id", "gene_distance"),
         to=c(mgmpg, mgmpgid, mgmpd)
      )
      for (i in c(mgmpg, mgmpgid, mgmpd)) {
         printDebug(i);
         fc_gr_dist_df1 <- unique(fc_gr_dist_df[,c(genecol1, i),drop=FALSE]);
         if (igrepHas("_distance", i)) {
            fc_gr_dist_df1vals <- nameVector(fc_gr_dist_df1[,c(i, genecol1),drop=FALSE]);
            values(fc_gr)[,i] <- c(NA, 0)[1];
         } else {
            fc_gr_dist_df1vals <- cPaste(split(fc_gr_dist_df1[[i]], fc_gr_dist_df1[[genecol1]]))
            values(fc_gr)[,i] <- head(c(NA, ""), 1);
         }
         imatch <- match(names(fc_gr_dist_df1vals),
            values(fc_gr)[[genecol1]]);
         if (length(imatch) > 0) {
            if (igrepHas("_distance", i)) {
               values(fc_gr)[imatch,i] <- fc_gr_dist_df1vals;
            } else {
               values(fc_gr)[imatch,i] <- as.character(fc_gr_dist_df1vals);
            }
         }
      }
   }, error=function(e){
      printDebug("Error during ATAC peak TSS annotation:",
         fgText="red");
      print(e);
   });
} else {
   ATAC <- FALSE;
}
```

### Annotate by Full Gene Name

This section uses each `_gene_id` column to
retrieve the Entrez gene description, using the R package
`"genejam"` if available. The resulting column has suffix
`_gene_description`.

```{r, add_gene_names}
if (suppressPackageStartupMessages(require(genejam))) {
   if (igrepHas("gene_id", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_id", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genenamecol <- gsub("gene_id", "gene_name", genecol);
         genedescol <- gsub("gene_id", "gene_description", genecol);
         #jamba::printDebug(c("Converting from '", genecol,
         #   "' to '", genedescol, "'"), sep="");
         genedf0 <- genejam::freshenGenes(
            data.frame(intermediate=values(fc_gr)[[genecol]],
               stringsAsFactors=FALSE),
            try_list=NULL,
            sep=";",
            intermediate="intermediate",
            final=c("SYMBOL", "GENENAME"));
         genedf3 <- trim_multigenedesc(genedf0, 4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            jamba::printDebug("   No gene descriptions were found for the input data.", fgText="red2");
         }
      }
   } else if (igrepHas("gene_name", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_name", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genedescol <- gsub("gene_name", "gene_description", genecol);
         jamba::printDebug(c("Converting from '", genecol,
            "' to '", genedescol, "'"), sep="");
         genedf2 <- genejam::freshenGenes(
            data.frame(gene_id=values(fc_gr)[[genecol]],
               stringsAsFactors=FALSE),
            try_list=c("SYMBOL2EG", "ALIAS2EG"),
            sep=";",
            intermediate="intermediate",
            final=c("SYMBOL", "GENENAME"))
         genedf3 <- trim_multigenedesc(genedf2[,c("intermediate", "SYMBOL", "GENENAME"),drop=FALSE], 4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            jamba::printDebug("   No gene descriptions were found for the input data.", fgText="red2");
         }
      }
   }
}
```


### Export stats results for hits (all peaks)

Statistical results are saved in text file for the full set
of peaks.
Look in the folder defined by OUTDIR:

> **`r OUTDIR`**


```{r, stats_df, dependson=c("run_limma")}
if (length(AtacStats) > 0) {
   ## Subset statsDF for rows with at least one hit
   for (iSig in dimnames(AtacStats$hit_array)[[3]]) {
      iDF_all <- AtacStats$stats_df[[iSig]];
      if (nrow(iDF_all) == 0) {
         next;
      }

      ## Add nearest promoter df
      pro_cols1 <- provigrep(c("^gene_id",
         "^gene_name",
         "feature_type($|_winner)",
         "_gene",
         "_distance",
         "broad_peak|num_slices"),
         colnames(values(fc_gr)));
      pro_cols <- intersect(
         colnames(values(fc_gr)),
         pro_cols1
      );
      if (length(pro_cols) > 0) {
         idf_match <- match(iDF_all[,1], values(fc_gr)$Gene);
         promoter_df <- data.frame(check.names=FALSE,
            stringsAsFactors=FALSE,
            values(fc_gr[idf_match,pro_cols]));
         iDF_all[,pro_cols] <- promoter_df;
      }
      
      ## Add genome coordinate annotations
      coord_cols <- c("Chr", "Start", "End", "Length");
      coord_df <- data.frame(check.names=FALSE,
         stringsAsFactors=FALSE,
         rowData(
         fc_se[match(iDF_all[,1], rowData(fc_se)$Geneid),]
         )[,coord_cols]);
      iDF_all2 <- data.frame(check.names=FALSE,
         stringsAsFactors=FALSE,
         iDF_all[,1,drop=FALSE],
         coord_df,
         iDF_all[,-1,drop=FALSE]);

      txt_ext <- paste0(".stats_nomgm",
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".txt");
      txt_file <- file.path(OUTDIR,
         paste0("slicejam_stats",
            txt_ext))
      # txt_file <- file.path(OUTDIR,
      #    gsub("[.]fc$|[.]fc[.]txt$",
      #       ignore.case=TRUE,
      #       txt_ext,
      #       basename(fc_file)));

      data.table::fwrite(
         x=iDF_all2,
         file=txt_file,
         sep="\t",
         col.names=TRUE);
      kable_coloring(
         data.frame(check.names=FALSE,
            `Rows saved:`=formatInt(nrow(iDF_all2)),
            `Output file:`=txt_file),
         row.names=FALSE,
         caption="File saved:") %>%
      kableExtra::column_spec(column=2,
         border_left="1px solid #DDDDDD")

      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all2,
         MGM=0,
         type="all");
      bed_ext <- paste0(".stats_nomgm",
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".bed");
      bed_file <- file.path(OUTDIR,
         paste0("slicejam_stats",
            bed_ext))
      # bed_file <- file.path(OUTDIR,
      #    gsub("[.]fc$|[.]fc[.]txt$",
      #       ignore.case=TRUE,
      #       bed_ext,
      #       basename(fc_file)));
      ## Add track header line
      trackname <- paste0(
         format(Sys.time(), "%Y%m%d%H%M"),
         "all");
      trackline <- paste0('track name=',
         trackname,
         ' description="',
         format(Sys.time(), "%Y%m%d%H%M"),
         ' all peaks"');
      writeLines(text=trackline,
         con=bed_file);
      data.table::fwrite(
         x=statsbed,
         file=bed_file,
         append=TRUE,
         sep="\t",
         col.names=FALSE);
      kable_coloring(
         data.frame(check.names=FALSE,
            `Rows saved:`=formatInt(nrow(statsbed)),
            `BED file:`=bed_file),
         row.names=FALSE,
         caption="File saved:") %>%
      kableExtra::column_spec(column=2,
         border_left="1px solid #DDDDDD")
   }
}
```

### Export stats results for hits (subset peaks)

Statistical results are saved in text file for the subset
of peaks with max group mean >= **`r MGM`**.
Look in the folder defined by OUTDIR: **`r OUTDIR`**.

A BED file is also saved during this step.

```{r, stats_df_mgm, dependson=c("run_limma_mgm")}
if (length(AtacStatsMgm) > 0) {
   ## Subset statsDF for rows with at least one hit
   for (iSig in dimnames(AtacStatsMgm$hit_array)[[3]]) {
      iDF_all <- AtacStatsMgm$stats_df[[iSig]];
      if (nrow(iDF_all) == 0) {
         next;
      }

      ## Add nearest promoter df
      pro_cols1 <- provigrep(c("^gene_id",
         "^gene_name",
         "feature_type($|_winner)",
         "_gene",
         "_distance",
         "broad_peak|num_slices"),
         colnames(values(fc_gr)));
      pro_cols <- intersect(
         colnames(values(fc_gr)),
         pro_cols1
      );
      if (length(pro_cols) > 0) {
         idf_match <- match(iDF_all[,1], values(fc_gr)$Gene);
         promoter_df <- data.frame(check.names=FALSE,
            stringsAsFactors=FALSE,
            values(fc_gr[idf_match,pro_cols]));
         iDF_all[,pro_cols] <- promoter_df;
      }
      
      ## Add genome coordinate annotations
      coord_cols <- c("Chr", "Start", "End", "Length");
      coord_df <- data.frame(check.names=FALSE,
         stringsAsFactors=FALSE,
         rowData(
            fc_se[match(iDF_all[,1], rowData(fc_se)$Geneid),]
         )[,coord_cols]);
      iDF_all2 <- data.frame(check.names=FALSE,
         stringsAsFactors=FALSE,
         iDF_all[,1,drop=FALSE],
         coord_df,
         iDF_all[,-1,drop=FALSE]);

      txt_ext <- paste0(".stats_mgm",
         MGM,
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".txt");
      txt_file <- file.path(OUTDIR,
         paste0("slicejam_stats",
            txt_ext))
      # txt_file <- file.path(OUTDIR,
      #    gsub("[.]fc$|[.]fc[.]txt$",
      #       ignore.case=TRUE,
      #       txt_ext,
      #       basename(fc_file)));
      
      cat(file="debug_log.txt", txt_file, "\n");
      data.table::fwrite(
         x=iDF_all2,
         file=txt_file,
         sep="\t",
         col.names=TRUE);
      kable_coloring(
         data.frame(check.names=FALSE,
            `Rows saved:`=formatInt(nrow(iDF_all2)),
            `Output file:`=txt_file),
         row.names=FALSE,
         caption="File saved:") %>%
      kableExtra::column_spec(column=2,
         border_left="1px solid #DDDDDD")

      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all2,
         MGM=MGM,
         type="all");
      if (length(statsbed) > 0) {
         bed_ext <- paste0(".stats_mgm",
            MGM,
            ifelse(ATAC, "_atac1", ""),
            ".", iSig, ".bed");
         bed_file <- file.path(OUTDIR,
            paste0("slicejam_stats",
               bed_ext))
         # bed_file <- file.path(OUTDIR,
         #    gsub("[.]fc$|[.]fc[.]txt$",
         #       ignore.case=TRUE,
         #       bed_ext,
         #       basename(fc_file)));
         ## Add track header line
         trackname <- paste0(
            format(Sys.time(), "%Y%m%d%H%M"),
            "mgm");
         trackline <- paste0('track name=',
            trackname,
            ' description="',
            format(Sys.time(), "%Y%m%d%H%M"),
            ' mgm', MGM, ' peaks"');
         writeLines(text=trackline,
            con=bed_file);
         data.table::fwrite(
            x=statsbed,
            file=bed_file,
            append=TRUE,
            sep="\t",
            col.names=FALSE);
         kable_coloring(
            data.frame(check.names=FALSE,
               `Rows saved:`=formatInt(nrow(statsbed)),
               `BED file:`=bed_file),
            row.names=FALSE,
            caption="File saved:") %>%
         kableExtra::column_spec(column=2,
            border_left="1px solid #DDDDDD")
      }
      
      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all2,
         MGM=MGM,
         type="hits");
      if (length(statsbed) > 0) {
         bed_ext <- paste0(".stathits_mgm",
            MGM,
            ifelse(ATAC, "_atac1", ""),
            ".", iSig, ".bed");
         bed_file <- file.path(OUTDIR,
            paste0("slicejam_stats",
               bed_ext))
         # bed_file <- file.path(OUTDIR,
         #    gsub("[.]fc$|[.]fc[.]txt$",
         #       ignore.case=TRUE,
         #       bed_ext,
         #       basename(fc_file)));
         ## Add track header line
         trackname <- paste0(
            format(Sys.time(), "%Y%m%d%H%M"),
            "hit");
         trackline <- paste0('track name=',
            trackname,
            ' description="',
            format(Sys.time(), "%Y%m%d%H%M"),
            ' mgm', MGM, ' hits"');
         writeLines(text=trackline,
            con=bed_file);
         data.table::fwrite(
            x=statsbed,
            file=bed_file,
            append=TRUE,
            sep="\t",
            col.names=FALSE);
         kable_coloring(
            data.frame(check.names=FALSE,
               `Rows saved:`=formatInt(nrow(statsbed)),
               `BED file:`=bed_file),
            row.names=FALSE,
            caption="File saved:") %>%
         kableExtra::column_spec(column=2,
            border_left="1px solid #DDDDDD")
      }
   }
}
```




## Export RData for follow-up analyses

**`r if (SAVE_RDATA) {("RData was saved.")} else {("RData was not saved.")}`**


```{r, export_rdata, dependson=c("check_reps", "run_limma", "run_limma_mgm")}
if (SAVE_RDATA) {
   rdata_ext <- ".RData";
   fc_dat_file <- file.path(OUTDIR,
      "slicejam_analysis.RData")

   if (!igrepHas("RData", fc_dat_file)) {
      fc_dat_file <- paste0(fc_dat_file, ".RData");
   }
   save.image(fc_dat_file);
   (kable_coloring(
      data.frame(check.names=FALSE,
         `RData file saved:`=fc_dat_file),
      row.names=FALSE));
} else {
   (kable_coloring(
      data.frame(check.names=FALSE,
         `RData session not saved:`=paste0("SAVE_RDATA=", SAVE_RDATA)),
      row.names=FALSE));
}
```

## R Session Info

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseSessionInfo1" role="button" aria-expanded="false" aria-controls="collapseSessionInfo1">(R Session Information)</a></p>
<div class="collapse" id="collapseSessionInfo1"><div class="card card-body">

```{r r_session_info, cache=FALSE}
sessioninfo::session_info()
```

</div></div>
