---
title: "Slicejam Analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
    df_print: kable
    self_contained: false
---
   
About this document:

* prepared on **`r format(Sys.time(), "%a %b %d %X %Y")`**
* using **`r R.version.string`**
* R slicejam package version **`r packageVersion("slicejam")`**.

### R environment setup

This document can be reproduced using the R code displayed in
this document. The initial steps involve loading R packages
into memory.

```{r, outdir}
OUTDIR <- Sys.getenv("OUTDIR");
if (length(OUTDIR) == 0 || nchar(OUTDIR) == 0) {
   OUTDIR <- ".";
}
CACHEDIR <- Sys.getenv("CACHEDIR");
if (length(CACHEDIR) == 0 || nchar(CACHEDIR) == 0) {
   CACHEDIR <- file.path(OUTDIR, "cache", "");
}
if (length(grep("/$", CACHEDIR)) == 0) {
   CACHEDIR <- file.path(CACHEDIR, "");
}
SAVE_RDATA <- Sys.getenv("SAVE_RDATA");
if (length(SAVE_RDATA) == 0 || nchar(SAVE_RDATA) == 0 || SAVE_RDATA %in% c("0", "FALSE", "F")) {
   SAVE_RDATA <- FALSE;
} else {
   SAVE_RDATA <- TRUE;
}
ATAC <- Sys.getenv("ATAC");
if (length(ATAC) == 0 || nchar(ATAC) == 0 || is.na(ATAC) || ATAC %in% c(0, "0", "F", "NA", "N", "n")) {
   ATAC <- FALSE;
} else {
   ATAC <- TRUE;
}
GROUPCHECK <- Sys.getenv("GROUPCHECK");
if (length(GROUPCHECK) == 0 ||
      nchar(GROUPCHECK) == 0 || 
      is.na(GROUPCHECK) || 
      GROUPCHECK %in% c(0, "0", "F", "NA", "N", "n")) {
   GROUPCHECK <- FALSE;
} else {
   GROUPCHECK <- TRUE;
}
GTF <- Sys.getenv("GTF");
if (length(GTF) == 0 || nchar(GTF) == 0) {
   #GTF <- "/ddn/gs1/project/nextgen/post/archer1/gokey_working/EA_Jul052017_RNAseq/hg19_refgene_041619_minimus_exons.gtf";
   #GTF <- "/ddn/gs1/project/nextgen/post/archer1/gokey_working/Copied_Intersections/110519_ATAC_broadpeak/To_James/hg19_refgene_041619_minimus_exons_v2.gtf";
   #GTF <- "/ddn/gs1/project/nextgen/post/archer1/gokey_working/Adelman/Scripts/fetch_refGene_gtf/test/hg19_121619_genes.gtf"; # 20dec2019
   stop("The GTF environment variable is required.");
}
if (!file.exists(GTF)) {
   stop(paste0("The GTF was not found:", GTF));
}
```

```{r setup, include=TRUE}
#    css: liao_style.css
## To render this page:
## In the farrisSeq folder run:
## rmarkdown::render("Adam19_Rnaseq_Analysis_vM21.Rmd", "html_document", knit_root_dir=getwd(), output_dir=getwd());
##    css: /Users/wardjm/Projects/R-scripts/liao_style.css

## Define the cache so it is dependent upon:
## - the versions of a few R packages,
## - the environmental variable FC_FILE
knitr::opts_chunk$set(cache=TRUE,
   autodep=TRUE,
   dev=c("png", "cairo_pdf"),
   dev.args=list(
      pointsize=14
   ),
   fig.align="center",
   pdf.options(encoding="ISOLatin9.enc"),
   cache.path=CACHEDIR,
   cache.extra=list(
      packageVersion("jamba"),
      packageVersion("jamma"),
      packageVersion("splicejam"),
      paste0("FC_FILE:", Sys.getenv("FC_FILE")),
      paste0("GTF:", Sys.getenv("GTF"))
   )
);

options("kable_styling_position"="left",
   "kable_styling_full_width"=FALSE,
   "kable_styling_bootstrap_options"=c("striped",
      "hover",
      "condensed",
      "responsive")
);
## Set font to Arial which helps compatibility in Adobe Illustrator,
## and helps display of up and down arrow characters in Venn diagrams
par("family"="Arial",
   "lend"="square",
   "ljoin"="mitre");
```


```{r, r_initialization, cache=FALSE}
## Prepend a common R package install location to .libPaths()
#.libPaths(
#   file.path(dirname(Sys.getenv("HOME")),
#      "wardjm/R/x86_64-pc-linux-gnu-library/3.5")
#);
print(".libPaths():");
print(.libPaths());

## Load R packages
suppressPackageStartupMessages(library(jamba));
if (!"CBioRUtils" %in% search()) {
   cbioPath <- file.path(
      dirname(Sys.getenv("HOME")),
      "wardjm/Builds/NIEHS_IntegrativeBIx-svn/Bpipe/bpipeScripts/CBioRUtils.R"
   );
   CBioRUtils <- new.env(parent=.GlobalEnv);
   suppressMessages(
      sys.source(cbioPath,
         envir=CBioRUtils,
         keep.source=TRUE));
   attach(CBioRUtils);
} else {
   ## Refresh CBioRUtils.R
   refreshCBioRUtils();
}
suppressPackageStartupMessages(library(SummarizedExperiment));
suppressPackageStartupMessages(library(colorjam));
suppressPackageStartupMessages(library(jamma));
suppressPackageStartupMessages(library(splicejam));
suppressPackageStartupMessages(library(data.table));
suppressPackageStartupMessages(library(matrixStats));
suppressPackageStartupMessages(library(extrafont));
suppressPackageStartupMessages(library(GenomicFeatures));
suppressPackageStartupMessages(library(rtracklayer));
suppressPackageStartupMessages(library(methods));
suppressPackageStartupMessages(library(multienrichjam));

```

## Description of workflow

* Import the first `.fc` file found in the current working directory.

    * Optionally use "FC_FILE" environmental variable.

* Curate filename columns by one of two processes:

    * If environment variable `"CURATE_YAML"` is defined, and points to
    a file that exists, the yaml curation workflow is used.
    * Otherwise the following rules are used to derive Sample, Group, Run,
    assuming the colnames are in the rough format **RUN**_*GROUP*_**REP**:
    
        * Samples are defined by calling `basename()` to remove the
         directory of each file (if it exists), then removing
         the file stem, and any recognized keywords such as `"dedup"`,
         `"cutadapt"`, `"fastq"`, etc.
        * Sample is split using the delimiter `"[-_.]"` and the
         first word is used for `"Run"`, the second word is used
         for `"Group"`.
        * The delimiter therefore can be any of: hyphen `"-"`,
         underscore `"_"`, or decimal `"."`.
        * Group is converted to factor, whose levels are ordered
         so that several recognized control labels are first:
         `"Control","ctrl","ctl","veh","blank","ntc","Luc"`,
         followed by the groups in the order they appear as column
         names in the featureCounts file.

* An R object `SummarizedExperiment` is created, using the columns
from featureCounts to populate rows, such as
`"Geneid", "Chr", "Start", "End", "Strand", "Length"`.
* Assign categorical colors to the unique values in the `"Group"` column.

* Normalization and batch adjustment

    * Quantile-normalize peak counts across all samples.
    * When the `"Run"` column contains more than one unique value,
    apply batch adjustment using the `"Run"` column to define batch,
    and `"Group"` to define groups, using the quantile-normalized data.

* Create MA-plots:

    * un-normalized data, global-centered
    * un-normalized data, centered by "Run"
    * quantile-normalized data, global-centered
    * quantile-normalized data, centered by "Run"
    * quantile-normalized data, centered by "Group","Run"
    * quantile-normalizeddata, global-centered, batch-adjusted by "Run"

* Create sample correlation heatmaps:

   * quantile-normalized data, centered by the first factor level
   in "Group", intended to act as a control group.
   * quantile-normalized data, batch-adjusted by "Run", centered by the
   first factor level in "Group".

* Statistical comparisons

   * Use the `"Group"` column to define groups, design, and contrasts.
   
       * Note that contrasts are limited to comparisons to the first
       `"Group"` level.
   
   * Use quantile-normalized, batch-adjusted counts.
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05
   * Apply secondary max group mean filtering with MGM=`{r cat(MGM)}`, to
   reduce the number of peaks used for statistical testing
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05

* Create volcano plots

    * Plot each contrast using the P-value, then using adjusted P-value.

* Create Venn diagrams:

    * Include each set of two contrasts with "all peak" hits.
    * Include each set of two contrasts with "MGM-filtered peak" hits.

* Annotate peaks

    * Associate each peak with one or more overlapping feature_types
    * Call one feature_type_winner per peak, based upon order: promoter, TTS, exon, intron, intergenic
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "all peaks" comparison
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "MGM-filtered peaks" comparison

* Re-annotate peaks into broad_peaks by merging adjacent slices

    * Count broad_peak hits from the "all peaks" comparison.
    * Count broad_peak hits from the "MGM-filtered peaks" comparison.

* Heatmap of peak hits across samples

    * Heatmap of peak hits from the "all peaks" comparison.
    * Heatmap of peak hits from the "MGM-filtered peaks" comparison.

* Optional ATAC-mode annotation to "active promoter genes"

    * Define active promoter genes by presence of ATAC-seq peak in promoter region.
    * Annotate peaks by nearest active promoter peak, along with distance.

* Save stats results to text files

    * Save "all peak" stats to tab-delimited text file.
    * Save "MGM-filtered peak" stats to tab-delimited text file.
    * This file is also used to create the "table of Venn intersections."

* Optionally save RData session

## Analysis workflow


### Import .fc file

Currently this workflow will use a `".fc"` file defined
via the environmental variable `"FC_FILE"`, or if that
variable is not defined, it searches for the first `".fc"`
file in the current working directory.

```{r, import_fc}
## Try to use FC_FILE
fc_env <- Sys.getenv("FC_FILE");
if (nchar(fc_env) > 0) {
   if (!file.exists(fc_env)) {
      stop(paste0("FC_FILE does not exists as a file:", fc_env,
         " in working directory:", getwd()));
   }
   fc_file <- fc_env;
} else {
   fc_file <- head(list.files(pattern="[.]fc$"), 1);
}
printDebug("Using fc_file:", fc_file);

## Import
fc_data <- data.frame(check.names=FALSE,
   stringsAsFactors=FALSE,
   data.table::fread(fc_file, skip="Geneid"));
## Add zero-padding to peak numbers
fc_data[,"Geneid"] <- paste0("peak_",
   jamba::padInteger(seq_len((nrow(fc_data)))));
rownames(fc_data) <- fc_data[,"Geneid"];

# featureCounts colnames
fc_colnames <- intersect(c("Geneid","Chr","Start","End","Strand","Length"),
   colnames(fc_data));
if (length(fc_colnames) == 0) {
   stop(paste0("None of the featureCounts colnames were detected: ", cPaste(fc_colnames)));
}
filenames <- setdiff(colnames(fc_data), fc_colnames);

curation_yaml <- Sys.getenv("CURATION_YAML",
   "curation.yaml");
if (length(curation_yaml) > 0 && nchar(curation_yaml) > 0 && file.exists(curation_yaml)) {
   curationL <- tryCatch({
      if (length(curation_yaml) == 1 && file.exists(curation_yaml)) {
         yaml::yaml.load(readLines(curation_yaml));
      } else {
         yaml::yaml.load(curation_yaml);
      }
   }, error=function(e){
      jamba::printDebug("Error: ",
         c("curation_yaml file '",
         curation_yaml,
         "' coult not be parsed by yaml::yaml.load()"),
         sep="",
         fgText=c("red", "dodgerblue"));
      list();
   });
   if (length(curationL) > 0) {
      sample_df <- splicejam::curateVtoDF(filenames, curationL);
      sample_colname <- head(vigrep("^sample[s]*$", colnames(sample_df)), 1)
      if (length(sample_colname) > 0) {
         iSamples <- jamba::makeNames(sample_df[[sample_colname]]);
      } else {
         jamba::printDebug("Error:",
            c("No '","Sample","' column was recognized."),
            sep="",
            fgText=c("red", "dodgerblue"));
         stop("No Sample column was recognized.");
      }
      run_colname <- head(vigrep("^run$", colnames(sample_df)), 1)
      if (length(run_colname) > 0) {
         Run <- sample_df[[run_colname]];
      } else {
         jamba::printDebug("Error:",
            c("No '","Run","' column was recognized."),
            sep="",
            fgText=c("red", "dodgerblue"));
      }
      group_colname <- head(vigrep("^Group$", colnames(sample_df)), 1)
      if (length(group_colname) > 0) {
         ## Ensure group names have no invalid characters like "-"
         sample_df[[group_colname]] <- gsub("[- ]+",
            ".",
            sample_df[[group_colname]]);
         iGroup <- sample_df[[group_colname]];
         iGroup <- factor(iGroup,
            levels=provigrep(c("UL3", "p2w5", "p4w4",
               "Control|ctrl", "Veh", "NTC", "blank",
               "^Luc",
               "."),
               unique(iGroup)));
         sample_df[[group_colname]] <- iGroup;
      } else {
         jamba::printDebug("Error:",
            c("No '","Group","' column was recognized."),
            sep="",
            fgText=c("red", "dodgerblue"));
         stop("No Group column was recognized.");
      }
      filename_colname <- head(vigrep("^filename$", colnames(sample_df)), 1);
      if (length(filename_colname) == 0) {
         filename_colname <- "Filename";
         sample_df[[filename_colname]] <- filenames;
      }
      rownames(sample_df) <- sample_df[[sample_colname]];
      knitr::kable(sample_df,
         caption="Sample Groups defined by curation.yaml");
      if (any(!sapply(c("iSamples", "Run", "iGroup"), exists))) {
         stop("One or more of iSamples,Run,Group could not be defined.");
      }
   } else {
      curation_yaml <- NULL;
   }
}
if (length(curation_yaml) == 0) {
   iSamples <- gsub("[-_.](v[0-9]|Coord|DeDup|dudup|cutadapt|sickle|sanger|trim|bam|fastq|fq).*$",
      ignore.case=TRUE,
      "",
      basename(filenames));
   Run <- gsub("[-_.].+$", "", iSamples);
   iGroup <- gsub("^([^-_.]+)[-_.]+([^-_.]+)[-_.].*$",
      "\\2",
      iSamples);
   iGroup <- gsub("[- ]+",
      ".",
      iGroup);
   iGroup <- factor(iGroup,
      levels=provigrep(c("UL3", "p2w5", "p4w4",
         "Control|ctrl|ctl", "Veh", "NTC", "blank",
         "."),
         unique(iGroup)));
   group_colname <- "Group";
   sample_colname <- "Sample";
   filename_colname <- "Filename";
   run_colname <- "Run";
   sample_df <- data.frame(
      check.names=FALSE,
      stringsAsFactors=FALSE,
      Filename=filenames,
      Sample=iSamples,
      Group=Group,
      Run=Run,
      row.names=iSamples
   );
   knitr::kable(sample_df,
      caption="Sample Groups defined by basic rules (not curation.yaml)");
}

## Optional GROUPCHECK which exits here to allow reviewing the sample grouping
if (GROUPCHECK) {
   txt_ext <- paste0(".curated_samples.txt");
   sample_txt_file <- file.path(OUTDIR,
      gsub("[.]fc$|[.]fc[.]txt$",
         txt_ext,
         basename(fc_file)));
   data.table::fwrite(
      x=sample_df,
      file=sample_txt_file,
      sep="\t",
      col.names=TRUE);
   stop(paste0("Curated sample data for review: '",
      sample_txt_file, "'"));
}

## Rename featureCounts columns to match iSamples
fc_data <- jamba::renameColumn(fc_data,
   from=filenames,
   to=iSamples);

## Make SummarizedExperiment
fc_se <- SummarizedExperiment(
   assays=list(counts=log2(1+as.matrix(fc_data[,iSamples,drop=FALSE]))),
   rowData=fc_data[,fc_colnames,drop=FALSE],
   colData=sample_df);

## Re-order samples by Group
iSamples <- mixedSortDF(colData(fc_se),
   byCols=group_colname)[[sample_colname]];
fc_se <- fc_se[,iSamples];
Run <- nameVector(colData(fc_se)[[run_colname]], colnames(fc_se));
iGroup <- nameVector(colData(fc_se)[[group_colname]], colnames(fc_se));

## Define sample colors
g_levels <- levels(colData(fc_se)[[group_colname]]);
n_levels <- length(g_levels);
sample_colors <- nameVector(
   head(rainbowJam(n_levels+3)[c(-3,-4,-6)], n_levels),
   g_levels);
color_sub1 <- nameVector(sample_colors[colData(fc_se)[[group_colname]]],
   colData(fc_se)[[sample_colname]]);
run_colors <- group2colors(unique(Run));

run_colors1 <- nameVector(run_colors[colData(fc_se[,iSamples])[[run_colname]]],
   iSamples);
```

### Plot distribution of peak sizes

If the featureCounts output included column name `"Length"`,
the plots below will show the distribution of lengths.

```{r, peak_sizes, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes",
      xlab="Peak width",
      breaks=100);
   
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes (log-scales x-axis",
      log="x",
      xlab="Peak width",
      breaks=100);
}
```


### Apply quantile normalization


```{r, norm_quantile, dependson="import_fc"}
## Raw data
iMatrix <- assays(fc_se[,iSamples])[["counts"]];

## Evaluate quantile normalization
fc_se <- doNorm(fc_se,
   signal="counts",
   methodList=c("quantile"));

## Quantile-normalized data
iMatrixQ <- assays(fc_se[,iSamples])[["quantile_counts"]];

```

### Apply batch-adjustment by "Run", grouped by "Group"

```{r, batch_adjust_run, dependson="norm_quantile"}
## Batch adjustment
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   fc_se_b <- tryCatch({
      removeAllNormBatchEffect(
         fc_se[,iSamples],
         groupColumn=group_colname,
         batchColumn=run_colname,
         signal=c("counts","quantile_counts"),
         adjustMethod="limma");
   }, error=function(e){
      printDebug("Error during removeAllNormBatchEffect():");
      print(e);
      NULL;
   });
   if (length(fc_se_b) > 0) {
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se_b[,iSamples])[["countsbatch"]];
      ## quantile-normalized data, batch-adjusted
      iMatrixQB <- assays(fc_se_b[,iSamples])[["quantile_countsbatch"]];
   } else {
      ## Do not perform batch adjustment
      fc_se_b <- fc_se;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
      ## quantile-normalized data, batch-adjusted
      iMatrixQB <- assays(fc_se[,iSamples])[["quantile_counts"]];
   }
} else {
   ## Do not perform batch adjustment
   fc_se_b <- fc_se;
   ## batch-adjusted raw data
   iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
   ## quantile-normalized data, batch-adjusted
   iMatrixQB <- assays(fc_se[,iSamples])[["quantile_counts"]];
}


```

### Create MA-plots

#### Un-normalized data, global-centered

```{r, ma_layout, dependson="import_fc"}
## Determine blank plot positions to organize Group and Run
Group_ctm <- as.matrix(table(colData(fc_se)[,c(group_colname, run_colname)]));
sub_n <- colMaxs(Group_ctm);
ctm_n <- matrix(cumsum(rep(sub_n, nrow(Group_ctm))),
   ncol=ncol(Group_ctm),
   byrow=TRUE);
#ctm_n;
ma_ncol <- sum(sub_n);
blank_pos <- unlist(lapply(1:nrow(Group_ctm), function(i){
   rmNULL(lapply(1:ncol(Group_ctm), function(j){
      iseq <- seq(from=ctm_n[i,j], by=-1, length.out=sub_n[j]);
      head(iseq, sub_n[j] - Group_ctm[i,j])
   }))
}));

```

```{r, ma_raw_global, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
jp_raw <- jammaplot(
   noiseFloor(iMatrix[,iSamples,drop=FALSE],
      minimum=4,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("un-normalized data", "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Un-normalized data, centered by "Run"

```{r, ma_raw_run, fig.height=10, fig.width=14, dependson=c("import_fc", "ma_layout")}
jp_raw_run <- jammaplot(
   noiseFloor(iMatrix[,iSamples,drop=FALSE],
      minimum=4,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("un-normalized data", "centered by 'Run'"),
   centerGroups=colData(fc_se[,iSamples])[[run_colname]],
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Quantile-normalized data, global-centered

```{r, ma_q_global, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
jp_q <- jammaplot(
   noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
      minimum=4,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("quantile-normalized data", "global-centered"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=Run[iSamples],
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Quantile-normalized data, centered by "Run"

```{r, ma_q_run, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
jp_q_run <- jammaplot(
   noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
      minimum=4,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("quantile-normalized data", "centered by 'Run'"),
   centerGroups=colData(fc_se[,iSamples])[[run_colname]],
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=2,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Quantile-normalized data, centered by "Group", Run"

If there are replicates per Group_Run, then MA-plots will
be displayed below, centered by Group and Run.

```{r, ma_q_group_run, fig.height=10, fig.width=14, dependson=c("norm_quantile", "ma_layout")}
grouprun_ct <- tcount(pasteByRow(colData(fc_se[,iSamples])[,c(group_colname,run_colname)]));
if (any(grouprun_ct > 1)) {
   jp_q_group_run <- jammaplot(
      noiseFloor(iMatrixQ[,iSamples,drop=FALSE],
         minimum=4,
         newValue=NA),
      ncol=ma_ncol,
      blankPlotPos=blank_pos,
      maintitle=c("quantile-normalized data", "centered by 'Group','Run'"),
      centerGroups=pasteByRow(colData(fc_se[,iSamples])[,c("Group","Run")]),
      groupSuffix="",
      subtitlePreset="bottomleft",
      subtitle=paste0("by ", iGroup[iSamples], ",", Run[iSamples]),
      subtitleBoxColor=get("blendColors", envir=CBioRUtils)(
         color_sub1[iSamples],
         run_colors1[iSamples]),
      useMean=TRUE,
      displayMAD=TRUE,
      outlierMAD=2,
      titleBoxColor=color_sub1[iSamples]
   )
}
```

#### Quantile-normalized data, global-centered, batch-adjusted by "Run"

This set of MA-plots represents the data used for statistical comparisons.
Note that panels display the MAD factor, which is a multiple of
variability of each sample to the median variability across all samples.
The threshold "2x MAD" is defined as one replicate having at least
twice the median variability across all samples.

> **At this point, outlier samples are not removed from
downstream statistical comparisons.**

```{r, ma_q_b_run, fig.height=10, fig.width=14, dependson=c("batch_adjust_run", "ma_layout")}
jp_q_b_run <- jammaplot(
   noiseFloor(iMatrixQB[,iSamples,drop=FALSE],
      minimum=4,
      newValue=NA),
   ncol=ma_ncol,
   blankPlotPos=blank_pos,
   maintitle=c("quantile-normalized data", "global-centered", "batch-adjusted by 'Run'"),
   groupSuffix="",
   subtitlePreset="bottomleft",
   subtitle=paste0("adj. by ", Run[iSamples]),
   subtitleBoxColor=run_colors1[iSamples],
   useMean=TRUE,
   displayMAD=TRUE,
   outlierMAD=1.951,
   titleBoxColor=color_sub1[iSamples]
)
```

#### Optional subset of sample outliers

Initially, all samples are used, but this section is a placeholder
for outlier detection and removal.

```{r, sample_outliers}
## Optional subset of samples (e.g. MA-plot outliers)
iSamplesUse <- iSamples;
printDebug("Using samples:", iSamplesUse);
```

### Sample Correlation Heatmaps


#### Quantile-normalized data, centered by Group control

```{r, cor_hm, fig.height=12, fig.width=12, dependson=c("norm_quantile", "sample_outliers")}
## Define centering control samples
iControl <- head(levels(iGroup), 1);
iControlSamples <- intersect(
   subset(colData(fc_se), colData(fc_se)[[group_colname]] %in% iControl)$Sample,
   iSamplesUse);
printDebug("Using control: ", iControl);
printDebug("Control Samples: ", iControlSamples);

## Center iMatrixQ
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQctr <- centerGeneData(iMatrixQ[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorM <- cor(use="pairwise.complete.obs",
   iMatrixQctr[,iSamplesUse,drop=FALSE]);

## hcluster
iCorM_hc <- tryCatch({
   amap::hcluster(iCorM);
}, error=function(e){
   printDebug("Error during hcluster(iCorM):");
   print(e);
   FALSE;
});

## Heatmap
heatmap.3(
   iCorM,
   hclCutoff=60,
   ColSideColors=data.frame(
      Group=color_sub1[iSamplesUse],
      Run=run_colors1[iSamplesUse]),
   ColSideColorsNote=data.frame(
      Group=iGroup[iSamplesUse],
      Run=Run[iSamplesUse]),
   cexColSideColorsNote=c(0.5, 1.5),
   RowSideColors=color_sub1[iSamplesUse],
   Rowv=iCorM_hc,
   Colv=iCorM_hc,
   margins=c(20,20),
   colLensFactor=5,
   add.expr=box(),
   plotAltRowDendro=FALSE,
   main=c("quantile-normalized data"),
   subMain=paste0("centered by ", iControl)
)
```


#### Quantile-normalized data, batch-adjusted, centered by Group control

```{r, cor_hm_b, fig.height=12, fig.width=12, dependson=c("batch_adjust_run", "sample_outliers")}
## Center iMatrixQB
## center all samples, but only relative to control samples in iSamplesUse
iMatrixQBctr <- centerGeneData(iMatrixQB[,iSamples,drop=FALSE],
   mean=TRUE,
   controlSamples=iControlSamples);

## Correlation matrix
iCorMB <- cor(use="pairwise.complete.obs",
   iMatrixQBctr[,iSamplesUse,drop=FALSE])

## hcluster
iCorMB_hc <- tryCatch({
   amap::hcluster(iCorMB);
}, error=function(e){
   printDebug("Error during hcluster(iCorM):");
   print(e);
   FALSE;
});

## Heatmap
heatmap.3(
   iCorMB,
   hclCutoff=60,
   ColSideColors=data.frame(
      Group=color_sub1[iSamplesUse],
      Run=run_colors1[iSamples]),
   ColSideColorsNote=data.frame(
      Group=iGroup[iSamplesUse],
      Run=Run[iSamplesUse]),
   cexColSideColorsNote=c(0.5, 1.5),
   RowSideColors=color_sub1[iSamplesUse],
   Rowv=iCorMB_hc,
   Colv=iCorMB_hc,
   margins=c(20,20),
   colLensFactor=5,
   add.expr=box(),
   plotAltRowDendro=FALSE,
   main=c("quantile-normalized data", "batch-adjusted by 'Run'"),
   subMain=paste0("centered by ", iControl)
)
```


### Statistical Comparisons

#### Confirm biological replicates

If there is only one replicate per sample group, then
the collection of samples are used to estimate
variability during statistical testing. This procedure
uses the moderated t-test from the `limma` R package,
which uses an improved estimation of variability
by including pooled variability from neighboring
measurements.


```{r, check_reps}
ctGroup <- table(iGroup[iSamplesUse]);
if (any(ctGroup > 1)) {
   printDebug("At least one sample group contains replicates.");
   denovo_controls <<- FALSE;
} else {
   printDebug("No sample group contains replicates. Pooled replicates will be created.");
   denovo_controls <<- TRUE;
   fc_se_ctrl <- fc_se_b[,iSamplesUse];
   colData(fc_se_ctrl)[,group_colname] <- "Control";
   colnames(fc_se_ctrl) <- jamba::makeNames(rep("Control", ncol(fc_se_ctrl)));
   colData(fc_se_ctrl)[,"Sample"] <- colnames(fc_se_ctrl);
   fc_se_b <- cbind(fc_se_ctrl, fc_se_b);
   iSamplesUse <- colnames(fc_se_b);

}
```

#### Define Design and Contrast Matrices

```{r, stats_design, dependson=c("batch_adjust_run", "sample_outliers", "check_reps")}
###############################################
## Statistical comparisons
conDesB <- splicejam::groups2contrasts(
   as.data.frame(colData(fc_se_b[,iSamplesUse])[,group_colname,drop=FALSE]),
   returnDesign=TRUE);
iContrasts <- conDesB$iContrasts;
iDesign <- conDesB$iDesign;

## Optionally remove denovo controls from contrasts if needed
if (denovo_controls) {
   iRemove <- igrep("Control-|-Control", colnames(iContrasts));
   if (length(iRemove) > 0) {
      iContrasts <- iContrasts[, -iRemove, drop=FALSE];
   }
}

color_sub_im <- c(`1`="#2222AA", `0`="white", `-1`="#BB5555");

kable_coloring(
   colorSub=color_sub_im,
   as.data.frame(iDesign),
   row.names=TRUE,
   caption="The sample design matrix: 'iDesign'"
   ) %>%
   kable_styling(bootstrap_options=c("striped","bordered")) %>%
   row_spec(0, background="#DDDDDD");

kable_coloring(
   colorSub=color_sub_im,
   as.data.frame(iContrasts),
   row.names=TRUE,
   caption="The group contrast matrix: 'iContrasts'"
   ) %>%
   kable_styling(bootstrap_options=c("striped","bordered")) %>%
   row_spec(0, background="#DDDDDD");
```

#### Limma stats

Limma uses a "moderated t-test" for statistics, and the method
below enables "limma-voom" which is intended for
use with count-based data.

```{r, run_limma, dependson=c("stats_design")}
## Find best available signal
signalList <- head(
   provigrep(c("quantile_countsbatch", "quantile_counts"),
      names(assays(fc_se_b))),
   1);
## Use all peaks
iPeaks <- rownames(fc_se);

## Determine whether we can use Voom
useVoom <- TRUE;
if (all(colSums(iDesign) == 1)) {
   useVoom <- FALSE;
}

## Stats using all batches
cutoffMaxGroupMean <- 4;
AtacStats <- tryCatch({
   allNormStatsTests(allNorm=fc_se_b[,iSamplesUse],
      signalList=signalList,
      useVoom=useVoom,
      iContrasts=iContrasts,
      iDesign=iDesign[iSamplesUse,,drop=FALSE],
      cutoffMaxGroupMean=cutoffMaxGroupMean,
      returnAllDFs=TRUE,
      returnLmFit=TRUE,
      iGenes=iPeaks,
      cutoffFold=c(1.5),
      cutoffP=NULL,
      transformAveExpr="none",
      cutoffAdjP=c(0.05),
      iSamples=iSamplesUse,
      collapseByGene=FALSE,
      verbose=FALSE);
}, error=function(e){
   printDebug("Error during allNormStatsTests():");
   print(e);
   NULL;
});
if (length(AtacStats) > 0) {
   printDebug("Stats hits by contrast and hit criteria:");
   AtacStats$statsHitsA;
}
```

#### Limma stats with subset peaks by signal threshold

The method below is identical to that used above, except that
peaks are filtered by minimum signal, since we apply a minimum
signal threshold for candidate peak hits. Peak mean counts
(normalized, batch-adjusted) are required to be at least
this same threshold in at least one sample group to be
included in the moderated t-test.

The default threshold is **4** after `log2(1+x)` transformation,
which corresponds to **15 counts** `(2^4 - 1)` per peak, averaged
across replicates in each sample group. This same threshold
is reflected in the MA-plots above, since the x-axis minimum
value is also **4**.

The maxGroupMean threshold is defined by the environmental
variable `MGM`, used to customize the threshold during the stats
tests:

```{r, mgm_arg}
## Get maxGroupMean threshold argument
MGM <- Sys.getenv("MGM", 4);
MGM <- jamba::rmNA(as.numeric(unlist(strsplit(MGM, "[ ,]+"))));
if (length(MGM) == 0) {
   MGM <- c(4);
}
if (length(MGM) > 1) {
   jamba::printDebug("maxGroupMean thresholds: ", MGM);
} else {
   jamba::printDebug("maxGroupMean threshold: ", MGM);
}
```

The max group mean threshold is: **`{r cat(MGM)}`**

```{r, run_limma_mgm4, dependson=c("stats_design","run_limma","mgm_arg")}
## Use the lmFit1 coefficients directly, which ensures
## we apply the threshold to exactly the same data 
## used in the limma test
if (length(AtacStats) > 0) {
   
   if ("quantile_countsbatch" %in% names(AtacStats$statsLmFits)) {
      coef_1 <- coefficients(AtacStats$statsLmFits$quantile_countsbatch$lmFit1);
   } else {
      coef_1 <- coefficients(AtacStats$statsLmFits$quantile_counts$lmFit1);
   }
   if (length(coef_1) == 0 || !is.matrix(coef_1)) {
      AtacStatsMgm4 <- NULL;
   } else {
      iPeaksMgm4 <- rownames(coef_1)[rowMaxs(coef_1) >= head(MGM, 1)];
      printDebug("Filtered from: ",
         format(nrow(coef_1), big.mark=","),
         " total peaks to ",
         format(length(iPeaksMgm4), big.mark=","),
         " (", format(length(iPeaksMgm4)*100/nrow(coef_1), digits=1), " %)",
         " with log2(1+x) group mean >= ",
         MGM,
         " (normalized, batch-adjusted) in at least one sample group.");
      
      ## Stats using all batches
      AtacStatsMgm4 <- tryCatch({
         allNormStatsTests(allNorm=fc_se_b[iPeaksMgm4,iSamplesUse],
            signalList=signalList,
            useVoom=TRUE,
            iContrasts=iContrasts,
            iDesign=iDesign[iSamplesUse,,drop=FALSE],
            cutoffMaxGroupMean=c(MGM),
            returnAllDFs=TRUE,
            returnLmFit=TRUE,
            iGenes=iPeaksMgm4,
            cutoffFold=c(1.5, 1.5),
            cutoffP=c(0.05,0.05),
            transformAveExpr="none",
            cutoffAdjP=c(1,0.05),
            iSamples=iSamplesUse,
            collapseByGene=FALSE,
            verbose=FALSE);
      }, error=function(e){
         printDebug("Error during allNormStatsTests():");
         print(e);
         NULL;
      });
   }
   if (length(AtacStatsMgm4) > 0) {
      printDebug("Stats hits by contrast and hit criteria ",
         c("(", "MGM ", ">= ", MGM, ")"), ":",
         sep="");
      AtacStatsMgm4$statsHitsA;
   }
}
```

#### Volcano plots (all peaks)

```{r, volcano_plots, fig.height=12, fig.width=12, dependson=c("run_limma")}
if (length(AtacStats) > 0) {

   for (iSig in dimnames(AtacStats$statsHitsA)[[3]]) {
      for (iCon in dimnames(AtacStats$statsHitsA)[[2]]) {
         iDF <- AtacStats$statsDFs[[iSig]][[iCon]];
         par("mar"=c(7,7,4,3));
         vp1 <- volcanoPlot(iDF,
            pvalueCutoff=0.05,
            fcCutoff=log2(1.5),
            intensityCutoff=0,
            main=iCon,
            subMain="All peaks",
            pvCol="P.Value");
         volcanoPlot(iDF,
            pvalueCutoff=0.05,
            fcCutoff=log2(1.5),
            intensityCutoff=0,
            main=iCon,
            subMain="All peaks",
            pvCol="adj.P.Val");
      }
   }
}
```

#### Volcano plots (subset peaks)

Volcano plots below were created using the limma results
after subsetting peaks for max group mean >= 
**`{r cat(MGM)}`**.

```{r, volcano_plots_mgm4, fig.height=12, fig.width=12, dependson=c("run_limma_mgm4")}
if (length(AtacStatsMgm4) > 0) {
   for (iSig in dimnames(AtacStatsMgm4$statsHitsA)[[3]]) {
      for (iCon in dimnames(AtacStatsMgm4$statsHitsA)[[2]]) {
         iDF <- AtacStatsMgm4$statsDFs[[iSig]][[iCon]];
         par("mar"=c(7,7,4,3))
         tryCatch({
            volcanoPlot(iDF,
               pvalueCutoff=0.05,
               fcCutoff=log2(1.5),
               intensityCutoff=MGM,
               main=iCon,
               subMain=paste0("Subset peaks mgm >= ", MGM),
               pvCol="P.Value");
            volcanoPlot(iDF,
               pvalueCutoff=0.05,
               fcCutoff=log2(1.5),
               intensityCutoff=MGM,
               main=iCon,
               subMain=paste0("Subset peaks mgm >= ", MGM),
               pvCol="adj.P.Val");
         }, error=function(e){
            jamba::printDebug("Error during volcanoPlot(), skipping.");
            jamba::printDebug("iSig:", iSig);
            jamba::printDebug("iCon:", iCon);
            print(e);
         });
      }
   }
}
```


#### Venn diagrams (all peaks)

```{r, venn_pairs, fig.height=10, fig.width=10, dependson=c("run_limma")}
if (length(AtacStats) > 0) {
   pair_l <- unlist(recursive=FALSE,
      apply(combn(dimnames(AtacStats$statsHitsA)[[2]], 2), 2, list));
   
   par("mar"=c(5,4,4,2));
   for (iSig in dimnames(AtacStats$statsHitsA)[[3]]) {
      for (iHit in dimnames(AtacStats$statsHitsA)[[1]]) {
         for (iCon in pair_l) {
            statsHitsAvenn(AtacStats$statsHitsA,
               iCon=iCon,
               iHit=iHit,
               iSig=iSig);
         }
      }
   }
}
```

#### Venn diagrams (subset peaks)

```{r, venn_pairs_mgm4, fig.height=10, fig.width=10, dependson=c("run_limma_mgm4")}
if (length(AtacStatsMgm4) > 0) {
   pair_l_mgm4 <- unlist(recursive=FALSE,
      apply(combn(dimnames(AtacStatsMgm4$statsHitsA)[[2]], 2), 2, list));
   
   par("mar"=c(5,4,4,2));
   for (iSig in dimnames(AtacStatsMgm4$statsHitsA)[[3]]) {
      for (iHit in dimnames(AtacStatsMgm4$statsHitsA)[[1]]) {
         for (iCon in pair_l_mgm4) {
            statsHitsAvenn(AtacStatsMgm4$statsHitsA,
               iCon=iCon,
               iHit=iHit,
               iSig=iSig);
         }
      }
   }
}
```


### Annotate peaks

The GTF file was used to define genomic regions. Each peak is annotated
with only one of the following categories, assigned in order.
Therefore a peak that overlaps a promoter and an exon is only
annotated "promoter".

* **promoters**: 2,000 upstream and 200 downstream the TSS for each transcript
* **TTS**: the last stranded nucleotide position in each transcript
* **exon**: each stranded exon per transcript
* **intron**: non-exonic region within a stranded transcript region
* **intergenic**: all other regions.

#### Define genome regions (or load previously defined genome regions)

The GTF file is used to derive a filename to store genome regions
derived from the gene-transcript-exon features.

The GTF file:

`r jamba::printDebug(unlist(strsplit(GTF, "/")), collapse="/", sep="")`

is renamed to have file extension:

`.genome_regions.RData`

If the RData file does not exist, genome regions are defined, then
saved into that RData file.

```{r, refgene_txdb}

#refgene_gtf <- "/ddn/gs1/project/nextgen/post/archer1/gokey_working/EA_Jul052017_RNAseq/hg19_refgene_041619_minimus_exons.gtf";
refgene_gtf <- GTF;

## genome_regions may be saved to a file already
gtf_gr_file <- gsub("[.](gff|gtf|gff3)$",
   ".genome_regions.RData",
   ignore.case=TRUE,
   refgene_gtf);
if (!file.exists(gtf_gr_file) && file.exists(basename(gtf_gr_file))) {
   gtf_gr_file <- basename(gtf_gr_file);
}
if (file.exists(gtf_gr_file)) {
   genome_regions_o <- load(gtf_gr_file);
   if (!exists("genome_regions")) {
      printDebug("The RData file '",
         gtf_gr_file,
         "' did not contain R object '",
         "genome_regions", "'.");
      printDebug("Instead, it contained these R objects:",
         genome_regions_o);
   }
}

```

If genome regions must be defined, they will be defined below:

```{r, define_genome_regions_txdb}
## If genome_regions is not define, create it
if (!exists("genome_regions")) {
   printDebug("Defining genome regions from GTF features.");
   txdb_file <- gsub("[.]gtf", ".txdb",
      ignore.case=TRUE,
      refgene_gtf);
   if (file.exists(txdb_file)) {
      refgene_txdb <- AnnotationDbi::loadDb(txdb_file);
   } else {
      refgene_txdb <- GenomicFeatures::makeTxDbFromGFF(refgene_gtf);
   }
}
```

```{r, define_genome_regions_tx2geneDF}
if (!exists("genome_regions")) {
   ## tx2geneDF
   system.time(tx2geneDF <- unique(makeTx2geneFromGtf(refgene_gtf,
      geneAttrNames=c("gene_id", "gene_name"),
      txAttrNames=c("transcript_id"),
      geneFeatureType="exon",
      txFeatureType="exon")));
   rownames(tx2geneDF) <- tx2geneDF$transcript_id;
}
```

```{r, define_genome_regions_featuresBy}
if (!exists("genome_regions")) {

   ## Exons
   exonsByGene <- exonsBy(refgene_txdb, by="gene");
   values(exonsByGene@unlistData)$gene_name <- rep(
      tx2geneDF[match(names(exonsByGene),
         tx2geneDF$gene_id), "gene_name"],
      elementNROWS(exonsByGene));
   values(exonsByGene@unlistData)$gene_id <- rep(
      names(exonsByGene),
      elementNROWS(exonsByGene));
   
   ## transcript ranges
   txByGene <- transcriptsBy(refgene_txdb);
   values(txByGene@unlistData)$gene_name <- tx2geneDF[match(values(txByGene@unlistData)$tx_name,
      tx2geneDF$transcript_id), "gene_name"];
   values(txByGene@unlistData)$gene_id <- tx2geneDF[match(values(txByGene@unlistData)$tx_name,
      tx2geneDF$transcript_id), "gene_id"];
   
   ## TTS per transcript range, extend -1000,+1000 around TTS
   ttsByTx <- flank(txByGene@unlistData, start=FALSE, width=-1);
   ttsByTx <- flank(ttsByTx, width=1000, both=TRUE);
   
   ## promoters
   ## Use default values for GenomicFeatures::promoters()
   ## but store them here to make sure they do not change
   upstream <- 2000;
   downstream <- 200;
   promoters_gr <- promoters(refgene_txdb,
      upstream=upstream,
      downstream=downstream,
      use.names=TRUE);
   values(promoters_gr)$gene_name <- tx2geneDF[match(values(promoters_gr)$tx_name,
      tx2geneDF$transcript_id), "gene_name"];
   values(promoters_gr)$gene_id <- tx2geneDF[match(values(promoters_gr)$tx_name,
      tx2geneDF$transcript_id), "gene_id"];
   
   ## Introns
   intronsByGene <- GenomicRanges::setdiff(txByGene@unlistData,
      exonsByGene@unlistData);
   intronsByGene <- annotateGRfromGR(intronsByGene,
      txByGene@unlistData[,c("gene_name", "gene_id")]);

   ## Assemble each layer of genomic region
   promoter_name <- paste0("Promoters (-", upstream, ",+", downstream, ")");
   genome_regions_l <- list(
      promoters=promoters_gr[,c("gene_name", "gene_id")],
      exons=exonsByGene@unlistData[,c("gene_name", "gene_id")],
      introns=intronsByGene[,c("gene_name", "gene_id")],
      `TTS(-1000,+1000)`=ttsByTx[,c("gene_name", "gene_id")]);
   names(genome_regions_l)[1] <- promoter_name;
   genome_regions <- GRangesList(genome_regions_l)@unlistData;
   values(genome_regions)$feature_type <- rep(names(genome_regions_l),
      lengths(genome_regions_l));
   
   ## Save to the RData file
   tryCatch({
      save(list=c("genome_regions", "tx2geneDF"),
         file=gtf_gr_file);
      printDebug("Genome regions were saved for later re-use:",
         gtf_gr_file);
   }, error=function(e){
      save(list=c("genome_regions", "tx2geneDF"),
         file=basename(gtf_gr_file));
      gtf_gr_file <- basename(gtf_gr_file);
      printDebug("Genome regions were saved for later re-use:",
         gtf_gr_file);
   });
} else {
   printDebug("Genome regions were already defined.");
}
```

#### Annotate peaks by genome region


```{r, peak_genes}
## Rules:
## - first pass: annotate anything that overlaps the peak
##    - gene_name, feature_type
## - second pass: choose feature_type_winner, in order:
##    - promoter
##    - TTS
##    - exon
##    - intron
##    - anything else is "intergenic"
## - third pass: assign gene of "winner"
## - fourth pass: distance to nearest gene body
##    - gene_name of nearest gene body
##

## Annotate ATAC peaks
fc_gr <- as(rowData(fc_se), "GRanges");
values(fc_gr) <- renameColumn(values(fc_gr),
   from="Geneid",
   to="Gene");
#write.table(file="fc_gr.txt", as.data.frame(head(fc_gr, 10)), sep="\t");

## Expand comma-delimited genes from genome_regions
gr_expl <- strsplit(values(genome_regions)$gene_name, ",");
gr_expi <- rep(seq_along(genome_regions), lengths(gr_expl));
genome_regions_exp <- unname(genome_regions)[gr_expi];
values(genome_regions_exp)$gene_name <- unname(unlist(gr_expl));

## Remove spaces and commas from feature_type column
values(genome_regions_exp)$feature_type <- gsub("[, ]+",
   "",
   values(genome_regions_exp)$feature_type);

##
values(genome_regions_exp)[,"gene_feature_type"] <- pasteByRow(
   values(genome_regions_exp)[,c("gene_name", "feature_type")],
   sep=" ");

# Annotate each peak by each genome_region overlapped
## ~20 seconds, substantially faster than annotateGRfromGR()
printDebug("Annotating peaks by genome_regions");
system.time({
   fco <- GenomicRanges::findOverlaps(fc_gr,
      ignore.strand=TRUE,
      genome_regions_exp);
   grdt <- data.table::data.table(
      fc=values(fc_gr[queryHits(fco)])$Gene,
      gene_name=values(genome_regions_exp[subjectHits(fco)])$gene_name,
      gene_id=values(genome_regions_exp[subjectHits(fco)])$gene_id,
      feature_type=factor(
         values(genome_regions_exp[subjectHits(fco)])$feature_type,
         levels=provigrep(c("promoter", "tts", "exon", "intron", "intergenic", "."),
            unique(values(genome_regions_exp[subjectHits(fco)])$feature_type))),
      gene_feature_type=values(genome_regions_exp[subjectHits(fco)])$gene_feature_type);
   ## Iterate each column, combine multi-row features into one row
   ## then annotate using unique, sorted, comma-delimited values
   grd_colnames <- c("gene_name", 
      "gene_id", 
      "feature_type", 
      "gene_feature_type");
   grd_vals <- lapply(nameVector(grd_colnames), function(i){
      #printDebug(i);
      icols <- c("fc", i);
      idt <- grdt[, ..icols];
      ## expand comma-delimited entries
      grdt0ft <- unique(mixedSortDF(idt, byCols=1:2)[,c(1,2)]);
      grv <- unstrsplit(CharacterList(split(as.character(grdt0ft[[2]]), grdt0ft[[1]])), sep=",");
      #grv <- cPaste(CharacterList(split(as.character(grdt0ft[[2]]), grdt0ft[[1]])), verbose=TRUE);
      grv;
   });
   ## Add each column to fc_gr
   for (i in names(grd_vals)) {
      printDebug(i);
      imatch <- match(names(grd_vals[[i]]), values(fc_gr)$Gene);
      if ("feature_type" == i) {
         values(fc_gr)[,i] <- unname(c("intergenic", grd_vals[[i]][1])[1]);
      } else {
         values(fc_gr)[,i] <- unname(c(NA, grd_vals[[i]][1])[1]);
      }
      values(fc_gr)[imatch,i] <- grd_vals[[i]];
   }
})

## is_duplicate()
is_duplicate <- function(x) {
   duplicated(x) | duplicated(x,
      fromLast=TRUE);
}

## sort for best feature_type
printDebug("Annotating peaks by best feature_type");
system.time({
   grdt0 <- mixedSortDF(grdt,
      byCols=c("fc","feature_type","gene_name","gene_id"));
   ## find first row per peak
   grdt0_peak <- grdt0[["fc"]];
   grdt0_peaku <- unique(grdt0_peak);
   whichu <- match(grdt0_peaku, grdt0_peak);
   ## First row per peak
   grdt0_pk <- grdt0[whichu,][["fc"]];
   grdt0_bestft <- grdt0[whichu,][["feature_type"]];
   names(grdt0_bestft) <- grdt0_pk;
   
   ## subset for peaks having the best feature_type
   bestftcolnames <- c("fc", "feature_type", "gene_name", "gene_id");
   grdt0_hasbestft <- unique(
      subset(grdt0, 
         feature_type == grdt0_bestft[grdt0_peak])[,..bestftcolnames]);

   ## Comma-delimit values for each column
   ## peak_089268
   for (i in c("feature_type", "gene_name", "gene_id")) {
      inewcolname <- paste0(i, "_winner");
      #printDebug(i, ", ", inewcolname);
      ikeepcolnames <- c("fc", i);
      grdt0_hasbestft_sub <- unique(grdt0_hasbestft[, ..ikeepcolnames]);
      if (anyDuplicated(grdt0_hasbestft_sub[["fc"]])) {
         ## split only duplicate entries
         fcdupes <- is_duplicate(grdt0_hasbestft_sub[["fc"]]);
         #grdt0_bestvalue_dupe <- cPaste(split(grdt0_hasbestft_sub[[i]][fcdupes],
         #   grdt0_hasbestft_sub[["fc"]][fcdupes]));
         grdt0_bestvalue_dupe <- unstrsplit(
            sep=",",
            CharacterList(
               split(grdt0_hasbestft_sub[[i]][fcdupes],
                  grdt0_hasbestft_sub[["fc"]][fcdupes])));
         grdt0_bestvalue_nondupe <- nameVector(as.character(grdt0_hasbestft_sub[[i]][!fcdupes]),
            grdt0_hasbestft_sub[["fc"]][!fcdupes]);
         grdt0_bestvalue <- c(grdt0_bestvalue_dupe,
            grdt0_bestvalue_nondupe);
      } else {
         grdt0_bestvalue <- nameVector(as.character(grdt0_hasbestft_sub[[i]]),
            grdt0_hasbestft_sub[["fc"]]);
      }
      imatch <- match(names(grdt0_bestvalue),
         values(fc_gr)$Gene);
      if ("feature_type" %in% i) {
         values(fc_gr)[,inewcolname] <- "intergenic";
      } else {
         values(fc_gr)[,inewcolname] <- c(NA, "")[1];
      }
      if (length(imatch) > 0) {
         values(fc_gr)[imatch,inewcolname] <- as.character(grdt0_bestvalue);
      }
   }
});

## gene distance analysis
fc_gr_genedist <- GenomicRanges::distanceToNearest(fc_gr,
   ignore.strand=TRUE,
   genome_regions,
   select="all");
fc_gr_genedist_df <- data.frame(
   Gene=values(fc_gr[queryHits(fc_gr_genedist)])[["Gene"]],
   nearest_gene_distance=values(fc_gr_genedist)$distance,
   renameColumn(
      as.data.frame(values(genome_regions[subjectHits(fc_gr_genedist)])[,c("gene_name", "gene_id")]),
      from=c("gene_name", "gene_id"),
      to=c("nearest_gene_name", "nearest_gene_id"))
)
for (i in c("nearest_gene_name", "nearest_gene_id", "nearest_gene_distance")) {
   #printDebug(i);
   fc_gr_genedist_df1 <- unique(fc_gr_genedist_df[,c("Gene", i)]);
   if (igrepHas("_distance", i)) {
      fc_gr_genedist_df1vals <- nameVector(fc_gr_genedist_df1[,c(i, "Gene")]);
      values(fc_gr)[,i] <- c(NA, 0)[1];
   } else {
      #fc_gr_genedist_df1vals <- cPaste(split(fc_gr_genedist_df1[[i]], fc_gr_genedist_df1[["Gene"]]))
      fc_gr_genedist_df1vals <- unstrsplit(sep=",",
         CharacterList(
            split(fc_gr_genedist_df1[[i]], fc_gr_genedist_df1[["Gene"]])
         )
      )
      values(fc_gr)[,i] <- c(NA, "")[1];
   }
   imatch <- match(names(fc_gr_genedist_df1vals), values(fc_gr)$Gene);
   if (length(imatch) > 0) {
      if (igrepHas("_distance", i)) {
         values(fc_gr)[imatch,i] <- fc_gr_genedist_df1vals;
      } else {
         values(fc_gr)[imatch,i] <- as.character(fc_gr_genedist_df1vals);
      }
   }
}

feature_type_winner <- values(fc_gr)$feature_type_winner;
names(feature_type_winner) <- values(fc_gr)$Gene;

```

#### Peak feature_type incidence matrix

```{r, peak_featuretype_im}
## For fun make an logical matrix of feature_type
ftcols <- c("Promoter", "TTS", "exon", "intron", "intergenic")
fc_typem <- as.data.frame(lapply(nameVector(ftcols), function(i){
   j <- grepl(i, values(fc_gr)$feature_type);
   names(j) <- values(fc_gr)$Gene;
   j;
}));
m1 <- as.matrix(fc_typem);
m1_counts <- (t(m1) %*% (m1));
color_sub_imct <- colorjam::vals2colorLevels(
   nameVector(sort(unique(as.vector(m1_counts)))),
   lens=10);
m1k <- kable_coloring(as.data.frame(m1_counts),
   colorSub=color_sub_imct,
   color_cells=TRUE,
   row.names=TRUE) %>%
   kable_styling(bootstrap_options=c("striped","bordered")) %>%
   row_spec(0, background="#DDDDDD", color="#333333", bold=TRUE) %>%
   column_spec(1, background="#DDDDDD", color="#333333", bold=TRUE)
print(m1k)

#           Promoter  TTS  exon intron intergenic
#Promoter      10951 1155  2754   3263          0
#TTS            1155 5972  3811   2398          0
#exon           2754 3811 16191  12770          0
#intron         3263 2398 12770  72976          0
#intergenic        0    0     0      0     208245

```

#### Pie chart of statistical hits by genome region

A pie chart is created for each statistical comparison, showing
the number of peaks overlapping each genome region. A peak
is only assigned one genome region, with preference in order:
promoter, TTS, exon, intron, intergenic.

For comparison, the pie chart using all peaks is also shown.


```{r, pie_functions}
feature_to_pie <- function
(x,
 do_pie=TRUE,
 main=NULL,
 colorSub=NULL,
 ...)
{
   j <- table(rmNA(x));
   if (length(j) == 0) {
      return(NULL);
   }
   j_pct <- format(j / sum(j) * 100, digits=1, trim=TRUE);
   j_df <- data.frame(
      j,
      pct=paste0(j_pct, "%")
   );
   main_v <- strsplit(main, "\n")[[1]];
   if (length(main_v) == 4) {
      names(main_v) <- c("Contrast", "Hit", "Signal", "PeakSet");
   } else {
      main_v <- c(Contrast=main, Hit="", Signal="", PeakSet="");
   }
   j_df[,names(main_v)] <- data.frame(as.list(main_v));
   j_names <- names(j);
   colorSub <- rmNA(colorSub[j_names]);
   if (length(colorSub) == length(j_names)) {
      colorSub <- nameVector(rainbowJam(length(j_names)),
         j_names);
   }
   names(j) <- paste0(j_names,
      " (",
      format(trim=TRUE, big.mark=",", j),
      "; ",
      j_pct,
      "%)");
   main <- paste0(main,
      "\n(",
      sum(j),
      " total)");
   if (do_pie) {
      pie(j,
         col=rainbowJam(length(j)));
      title(main=main,
         ...);
   }
   return(j_df);
}

peak_pie_by_region <- function
(statsHitsA,
 feature_type_winner=NULL,
 PeakSet="",
 iHit=NULL,
 iSig=NULL,
 iCon=NULL,
 iPeaks=NULL,
 do_pie=TRUE,
 main=NULL,
 colorSub=NULL,
 ...)
{
   ## Wrapper function for peak-to-region
   if (length(iPeaks) > 0) {
      k <- iPeaks;
      if (length(main) == 0) {
         main <- "All Peaks";
      }
      printDebug("peak_pie_by_region(): ",
         "main:", main);
      j_df <- feature_to_pie(feature_type_winner[k],
         main=main,
         colorSub=colorSub,
         do_pie=do_pie);
      return(j_df);
   } else {
      if (length(iHit) == 0) {
         iHit <- dimnames(statsHitsA)[[1]];
      } else if (is.numeric(iHit)) {
         iHit <- dimnames(statsHitsA)[[1]][iHit];
      }
      if (length(iSig) == 0) {
         iSig <- dimnames(statsHitsA)[[3]];
      } else if (is.numeric(iSig)) {
         iSig <- dimnames(statsHitsA)[[3]][iSig];
      }
      if (length(iCon) == 0) {
         iCon <- dimnames(statsHitsA)[[2]];
      } else if (is.numeric(iSig)) {
         iCon <- dimnames(statsHitsA)[[2]][iSig];
      }
      printDebug("peak_pie_by_region(): ",
         "\niCon:", iCon,
         "\niHit:", iHit,
         "\niSig:", iSig);
      j_df_l <- rbindList(lapply(iHit, function(iHit1) {
         rbindList(lapply(iSig, function(iSig1) {
            rbindList(lapply(iCon, function(iCon1) {
               k <- names(statsHitsA[iHit1, iCon1, iSig1][[1]]);
               main <- paste(c(iCon1, iHit1, iSig1, PeakSet), collapse="\n");
               j_df <- feature_to_pie(feature_type_winner[k],
                  main=main,
                  colorSub=colorSub,
                  do_pie=do_pie);
               j_df;
            }))
         }))
      }));
      return(j_df_l);
   }
}
```


```{r, peak_region_pies, fig.height=12, fig.width=12}
if (length(AtacStats) > 0) {
   par("mfrow"=c(3,3), "mar"=c(5,5,6,8));
   colorSub_pie <- nameVector(
      rainbowJam(length(unique(feature_type_winner))),
         unique(feature_type_winner));
   j_df_s <- peak_pie_by_region(AtacStats$statsHitsA,
      PeakSet="All Peaks Tested",
      do_pie=TRUE,
      colorSub=colorSub_pie,
      feature_type_winner=feature_type_winner);
   if (length(j_df_s) > 0) {
      j_df_all <- peak_pie_by_region(iPeaks=rownames(fc_se),
         do_pie=TRUE,
         colorSub=colorSub_pie,
         feature_type_winner=feature_type_winner);
      j_df <- rbind(j_df_s, j_df_all);
      
      ## Print the text results
      kable_coloring(j_df,
         row.names=TRUE,
         caption="Data used for the pie charts:"
         ) %>%
         kable_styling(bootstrap_options=c("striped","bordered")) %>%
         row_spec(0, background="#DDDDDD");
   }
}
```

#### Pie chart (subset peaks)

The pie charts below reflect the peaks after filtering
for the max group mean threshold: **`{r cat(MGM)}`**


```{r, peak_region_pies_mgm4, fig.height=12, fig.width=12}
if (length(AtacStatsMgm4) > 0) {
   par("mfrow"=c(3,3), "mar"=c(5,5,6,8));
   j_df_mgm4_s <- peak_pie_by_region(AtacStatsMgm4$statsHitsA,
      PeakSet=paste0("Subset Peaks Mgm", MGM),
      colorSub=colorSub_pie,
      do_pie=TRUE,
      feature_type_winner=feature_type_winner);
   if (length(j_df_mgm4_s) > 0) {
      j_df_mgm4 <- peak_pie_by_region(iPeaks=iPeaksMgm4,
         do_pie=TRUE,
         colorSub=colorSub_pie,
         main=paste0("Subset Peaks Mgm", MGM),
         feature_type_winner=feature_type_winner)
      j_df_all <- peak_pie_by_region(iPeaks=iPeaks,
         do_pie=TRUE,
         colorSub=colorSub_pie,
         main="All Peaks",
         feature_type_winner=feature_type_winner)
      
      ## Print the text results
      j_df_mgm4_tall <- rbind(j_df_mgm4_s, j_df_mgm4);
      ## Print the text results
      kable_coloring(j_df_mgm4_tall,
         row.names=TRUE,
         caption="Data used for the pie charts:"
         ) %>%
         kable_styling(bootstrap_options=c("striped","bordered")) %>%
         row_spec(0, background="#DDDDDD");
   }
}
```


### Annotate peaks using continuous numbering

Peaks are re-annotated so that contiguous peaks are numbered
in column "broak_peak", with slices of each peak given suffix
numbers in column "broak_peak_number". This process makes it
feasible to count the number of contiguous peaks, that is
before peaks were sliced to a max width threshold.

We refer to "broad peaks" as the contiguous peaks.

```{r, peak_renumbering}
fc_gr_red <- reduce(fc_gr);
values(fc_gr_red)$broad_peak <- makeNames(
   rep("broad", length(fc_gr_red)),
   doPadInteger=TRUE,
   renameOnes=TRUE,
   suffix="_");
fc_gr <- annotateGRfromGR(fc_gr,
   fc_gr_red);
values(fc_gr)$broad_peak_number <- makeNames(
   values(fc_gr)$broad_peak,
   suffix="_",
   doPadInteger=TRUE,
   renameOnes=TRUE);
fc_gr_bp_ct <- table(values(fc_gr)$broad_peak);
values(fc_gr)$num_slices <- as.vector(fc_gr_bp_ct[values(fc_gr)$broad_peak]);
rowData(fc_se)$broad_peak <- values(fc_gr)$broad_peak;
rowData(fc_se)$broad_peak_number <- values(fc_gr)$broad_peak_number;
rowData(fc_se)$broad_peak_slices <- fc_gr_bp_ct[values(fc_gr)$broad_peak];
```


#### Broad peak hit counts

The peak statistical hits are re-counted by unique "broad_peak" values,
to indicate how many unique broad peaks are found to be statistically
changing.

If most peaks are not sliced, this number should be almost identical
to the number of sliced peak hits above.

Similarly, if most hits involve only one slice from broad_peaks that
have multiple slices, the number of broad_peak hits should be nearly
identical to the number of sliced peak hits above.

Lastly, if the number of broad_peak hits is substantially lower than
the number of sliced peak hits above, it suggests that broad_peaks
are statistically changing across multiple slices per peak.


```{r, broad_hit_counts}
if (length(AtacStats) > 0) {
   
   broad_hit_counts <- lapply(nameVector(dimnames(AtacStats$statsHitsA)[[1]]), function(p){
      sapply(nameVectorN(AtacStats$statsHitsA[p,,]), function(i){
         k <- AtacStats$statsHitsA[p,i,1][[1]];
         k_bp <- values(fc_gr[names(k)])$broad_peak;
         c(peak_hits=length(k_bp),
            broad_peak_hits=length(table(k_bp)));
      })
   });
   printDebug("broad_hit_counts:");
   broad_hit_counts;
}
   
if (length(AtacStatsMgm4) > 0) {
   broad_hit_counts_mgm4 <- lapply(nameVector(dimnames(AtacStatsMgm4$statsHitsA)[[1]]), function(p){
      sapply(nameVectorN(AtacStatsMgm4$statsHitsA[p,,]), function(i){
         k <- AtacStatsMgm4$statsHitsA[p,i,1][[1]];
         k_bp <- values(fc_gr[names(k)])$broad_peak;
         c(peak_hits=length(k_bp),
            broad_peak_hits=length(table(k_bp)));
      })
   });
   printDebug("broad_hit_counts ",
      c("(", "MGM", ">=", MGM, ")"),
      ":", sep="");
   broad_hit_counts_mgm4;

}
## Briefly compare All peak hits to Mgm4 peak hits
## signedVennPlotSets(list(mgm4=AtacStatsMgm4$statsHitsA[1,2,1][[1]], all=AtacStats$statsHitsA[1,2,1][[1]]))

```

### Heatmap of peak hits

```{r, heatmap_hits, fig.height=14, fig.width=10}
if (length(AtacStats) > 0) {
   stats_dim <- dim(AtacStats$statsHitsA);
   peak_hits <- unique(unlist(lapply(AtacStats$statsHitsA[,1:2,], names)));
   peak_hits_q <- unique(unlist(lapply(AtacStats$statsHitsA[stats_dim[1],1:2,], names)));
}
if (length(AtacStatsMgm4) > 0) {
   mgm4_stats_dim <- dim(AtacStatsMgm4$statsHitsA);
   peak_hits_mgm4 <- unique(unlist(lapply(AtacStatsMgm4$statsHitsA[,1:2,], names)));
   peak_hits_mgm4_q <- unique(unlist(lapply(AtacStatsMgm4$statsHitsA[mgm4_stats_dim[1],1:2,], names)));
   #lengths(list(peak_hits=peak_hits, peak_hits_q=peak_hits_q,
   #   peak_hits_mgm4=peak_hits_mgm4, peak_hits_mgm4_q=peak_hits_mgm4_q))
   
   if (denovo_controls) {
      iSamplesUse <- unvigrep("^Control_v", iSamplesUse);
   }
   if (length(peak_hits_mgm4_q) > 1) {
      iM <- iMatrixQBctr[peak_hits_mgm4_q,iSamplesUse,drop=FALSE];
      hc <- hcluster(iM, link="ward", method="euclidean");
      
      
      ## Subset only promoter peak hits
      #peak_hits_mgm4_q_p <- subset(rowData(fc_se[peak_hits_mgm4_q,]), feature_type_winner %in% "promoter")$Geneid;
      #iM <- iMatrixQBctr[peak_hits_mgm4_q_p,];
      #hc_p <- hcluster(iM, link="ward", method="euclidean");
      
      hm_mgm4 <- heatmap.3(iM,
      #   Colv=FALSE,
         main=paste0("Hits (MGM >= ", MGM, ")"),
         farbeLim=4,
         Rowv=hc,
         hclCutoff=60,
         labRow=NA,
         keysize=c(0.7,1),
         ColSideColors=data.frame(
            Group=color_sub1[iSamplesUse],
            Run=run_colors1[iSamplesUse]),
         ColSideColorsNote=data.frame(
            Group=iGroup[iSamplesUse],
            Run=Run[iSamplesUse]),
         cexColSideColorsNote=c(1.5, 1.5),
         cexColSideColors=1.5,
         colLensFactor=5,
         add.expr=box(),
         margins=c(17,7)
      )
   }
}
```

### Heatmap of peak hits using group means


```{r, heatmap_groupmeans, fig.height=14, fig.width=10}
if (length(AtacStatsMgm4) > 0) {
   if (denovo_controls) {
      iSamplesUse <- unvigrep("^Control_v", iSamplesUse);
   }
   if (length(peak_hits_mgm4_q) > 1) {
      iM_grp <- rowGroupMeans(iM[,iSamplesUse],
         useMedian=FALSE,
         groups=iGroup[iSamplesUse]);
      
      hm_mgm4_grp <- heatmap.3(iM_grp,
      #   Colv=FALSE,
         main=paste0("Hits (MGM >= ", MGM, ")"),
         Rowv=hc,
         hclCutoff=60,
         labRow=NA,
         keysize=c(0.7,1),
         ColSideColors=data.frame(
            Group=sample_colors[colnames(iM_grp)]),
         ColSideColorsNote=data.frame(
            Group=nameVector(colnames(iM_grp))),
         cexColSideColorsNote=c(1.5),
         cexColSideColors=1.5,
         colLensFactor=5,
         add.expr=box(),
         margins=c(7,7)
      )
   }
}
```




### Peak to ATAC-promoter analysis (optional)

**`r jamba::printDebug("ATAC:", ATAC)`**

**`{r if (ATAC && length(AtacStats) > 0 && exists("iPeaksMgm4") && length(iPeaksMgm4) > 0) {cat("ATAC-mode is enabled.")} else {cat("ATAC-mode is disabled."}}`**

This workflow is enabled only when environment variable `ATAC=1`,
and when stats have been performed, and when the max group mean
filter has been applied.

When enabled, this workflow defines a subset of genes as having
"active promoters" if there are any ATAC-seq peaks found in the
upstream promoter/TSS region of the gene. In practice, most genes
that are active, or poised to be active, have an ATAC-seq peak
across all conditions of an experiment. That said, this process
greatly reduced the total set of GTF-defined genes, which can make
the process of annotating peaks by genes much more accurate
and informative.

Peaks are annotated by the neared Mgm-filtered promoter peak,
along with the distance to that peak.


```{r, peak_promoters}
mgmpg <- NULL;
mgmpd <- NULL;
if (ATAC && length(AtacStats) > 0 && exists("iPeaksMgm4") && length(iPeaksMgm4) > 0) {
   
   ## iPeaksMgm4 that are promoters
   ## - call these active promoters
   ##
   ## annotate all peaks by nearest active promoter
   ## - promoter,tts,intron,exon -- gene_name intersected
   ## - all cases, nearest mgm4 promoter, and its gene_name
   ## - distance to nearest mgm4 promoter
   
   promoters_gr <- subset(genome_regions,
      grepl("promoter", 
         ignore.case=TRUE, 
         feature_type));
   #values(promoters_gr)[,"gene_strand"] <- as.character(strand(promoters_gr));
   
   promoter2tss <- function
   (gr,
    upstream=2000,
    downstream=200)
   {
      ## Purpose is to convert coordinates of promoters back to original TSS
      gr2pos <- strand(gr) %in% "+";
      newstart <- ifelse(gr2pos,
         start(gr) + upstream,
         start(gr) + (downstream-1));
      newend <- ifelse(gr2pos,
         end(gr) - (downstream-1),
         end(gr) - upstream);
      ranges(gr) <- IRanges(start=newstart,
         end=newend);
      return(gr);
   }
   upstream <- 2000;
   downstream <- 200;
   
   tryCatch({
      mgm_promoter_gr <- IRanges::subsetByOverlaps(promoters_gr,
         fc_gr[iPeaksMgm4],
         ignore.strand=TRUE);
      mgm_tss_gr <- promoter2tss(mgm_promoter_gr[,c("gene_name", "gene_id")],
         upstream=upstream,
         downstream=downstream);
      
      mgmpg <- paste0("mgm", MGM, "_annoTSS_gene_name");
      mgmpgid <- paste0("mgm", MGM, "_annoTSS_gene_id");
      mgmpd <- paste0("mgm", MGM, "_annoTSS_distance");
      
      ## nearest mgm_annoTSS distance
      fc_gr_dist <- GenomicRanges::distanceToNearest(fc_gr,
         ignore.strand=TRUE,
         select="all",
         mgm_tss_gr);
      #220020
      ## Add comma-delimited values per column as needed
      genecol1 <- head(provigrep(c("^Geneid", "^Gene"), colnames(values(fc_gr))), 1);
      fc_gr_dist_df <- renameColumn(
         data.frame(
            as.data.frame(values(fc_gr[queryHits(fc_gr_dist)])[,genecol1,drop=FALSE]),
            gene_distance=values(fc_gr_dist)$distance,
            as.data.frame(values(mgm_tss_gr[subjectHits(fc_gr_dist)])[,c("gene_name", "gene_id"),drop=FALSE])
         ),
         from=c("gene_name", "gene_id", "gene_distance"),
         to=c(mgmpg, mgmpgid, mgmpd)
      )
      for (i in c(mgmpg, mgmpgid, mgmpd)) {
         printDebug(i);
         fc_gr_dist_df1 <- unique(fc_gr_dist_df[,c(genecol1, i),drop=FALSE]);
         if (igrepHas("_distance", i)) {
            fc_gr_dist_df1vals <- nameVector(fc_gr_dist_df1[,c(i, genecol1),drop=FALSE]);
            values(fc_gr)[,i] <- c(NA, 0)[1];
         } else {
            fc_gr_dist_df1vals <- cPaste(split(fc_gr_dist_df1[[i]], fc_gr_dist_df1[[genecol1]]))
            values(fc_gr)[,i] <- head(c(NA, ""), 1);
         }
         imatch <- match(names(fc_gr_dist_df1vals),
            values(fc_gr)[[genecol1]]);
         if (length(imatch) > 0) {
            if (igrepHas("_distance", i)) {
               values(fc_gr)[imatch,i] <- fc_gr_dist_df1vals;
            } else {
               values(fc_gr)[imatch,i] <- as.character(fc_gr_dist_df1vals);
            }
         }
      }
   }, error=function(e){
      printDebug("Error during ATAC peak TSS annotation:",
         fgText="red");
      print(e);
   });
} else {
   ATAC <- FALSE;
}
```

### Annotate peaks with full gene name

This section uses the gene symbol or Entrez Gene ID to
retrieve the Entrez gene description, using the R package
`"genejam"` (if available).

```{r, add_gene_names}
if (suppressPackageStartupMessages(require(genejam))) {
   trim_multigenedesc <- function(genedf3, max_genes=4){
      multigrep <- paste0("^(",
         paste(rep("[^;]+", max_genes), collapse=";"),
         ");.+$");
      multirows <- grepl(multigrep, genedf3$SYMBOL);
      if (any(multirows)) {
         jamba::printDebug(c("   Trimming ", formatInt(sum(multirows)), " multi-gene rows."), sep="");
         genedf3$GENENAME <- gsub(multigrep, "\\1", genedf3$GENENAME);
         genedf3$SYMBOL <- gsub(multigrep, "\\1", genedf3$SYMBOL);
      }
      return(genedf3);
   }
   if (igrepHas("gene_id", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_id", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genenamecol <- gsub("gene_id", "gene_name", genecol);
         genedescol <- gsub("gene_id", "gene_description", genecol);
         jamba::printDebug(c("Converting from '", genecol, "' to '", genedescol, "'"), sep="");
         genedf1 <- genejam::freshenGenes(data.frame(gene_id=values(fc_gr)[[genecol]]),
            try_list=c("SYMBOL"),
            final=NULL);
         if (genenamecol %in% colnames(values(fc_gr))) {
            jamba::printDebug(c("   Including values from '", genenamecol, "'"), sep="");
            genedf1[,genenamecol] <- values(fc_gr)[[genenamecol]];
            genedf1 <- renameColumn(genedf1[,c("found",genenamecol),drop=FALSE],
               from="found", to="gene_id_found");
         } else {
            genedf1 <- renameColumn(genedf1[,"found",drop=FALSE],
               from="found", to="gene_id_found");
         }
         genedf2 <- genejam::freshenGenes(genedf1,
            try_list=c("SYMBOL2EG", "ALIAS2EG"),
            sep=";",
            final=c("SYMBOL", "GENENAME"))
         genedf3 <- data.frame(check.names=FALSE,
            genedf1,
            genedf2[,c("intermediate", "SYMBOL", "GENENAME"),drop=FALSE]);
         genedf3 <- trim_multigenedesc(genedf3, 4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            jamba::printDebug("   No gene descriptions were found for the input data.", fgText="red2");
         }
      }
   } else if (igrepHas("gene_name", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_name", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genedescol <- gsub("gene_name", "gene_description", genecol);
         jamba::printDebug(c("Converting from '", genecol, "' to '", genedescol, "'"), sep="");
         genedf2 <- genejam::freshenGenes(data.frame(gene_id=values(fc_gr)[[genecol]]),
            try_list=c("SYMBOL2EG", "ALIAS2EG"),
            sep=";",
            final=c("SYMBOL", "GENENAME"))
         genedf3 <- data.frame(check.names=FALSE,
            genedf1,
            genedf2[,c("intermediate", "SYMBOL", "GENENAME"),drop=FALSE]);
         genedf3 <- trim_multigenedesc(genedf3, 4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            jamba::printDebug("   No gene descriptions were found for the input data.", fgText="red2");
         }
      }
   }
}
```


### Export stats results for hits (all peaks)

Statistical results are saved in text file for the full set
of peaks.
Look in the folder defined by OUTDIR: **`{r cat(OUTDIR)}`**.

```{r, export_beds}
statsdf2bed <- function
(statsdf,
 type=c("all", "hits"),
 ...)
{
   ## This function converts Stats data.frame to BED format
   ## using the BED name field to encode additional info
   type <- match.arg(type);
   
   hit_cols <- vigrep("^hit ", colnames(statsdf));
   hit_cols_names <- gsub("^.+ ", "", hit_cols);
   hit_cols_names2 <- gsub("_v2", "", makeNames(hit_cols_names));
   names(hit_cols) <- hit_cols_names2;
   hit_cols_use <- hit_cols[unvigrep("_v[0-9]+$", hit_cols_names2)];
   hits_l <- lapply(hit_cols_use, function(i){statsdf[[i]] != 0})
   hits_v <- Reduce("|", hits_l);
   
   ## Optionally subset for stats hits
   if ("hits" %in% type) {
      statsdf <- statsdf[hits_v,,drop=FALSE];
   }
   
   hits_im <- as.matrix(statsdf[,hit_cols_use,drop=FALSE]);
   hits_score <- rowSums(abs(hits_im));
   ## Make BED name
   hits_vl <- lapply(nameVectorN(hit_cols_use), function(i){
      j <- statsdf[[hit_cols[i]]];
      ifelse(j == 0,
         "",
         paste0(j, "_", i))
   });
   ## _annoTSS_gene_name
   mgmpg <- paste0("mgm", MGM, "_annoTSS_gene_name");
   mgmpgid <- paste0("mgm", MGM, "_annoTSS_gene_id");
   mgmpd <- paste0("mgm", MGM, "_annoTSS_distance");

   statsdf$ft_winner <- gsub("[ ()].*$",
      "", 
      statsdf$feature_type_winner);

   hits_vm <- do.call(cbind, c(
      lapply(nameVector(c("Gene", "ft_winner", mgmpg, mgmpd)), function(i){
         rmNA(naValue="", statsdf[[i]])
      }),
      hits_vl
   ));
   hits_label <- pasteByRow(hits_vm, sep="|");
   statsdf$label <- hits_label;
   statsdf$score <- hits_score;
   statsdf$strand <- "+";
   statsbed <- statsdf[,c("Chr", "Start", "End", "label", "strand", "score"),drop=FALSE];
   return(statsbed);
}
```

```{r, stats_df, dependson=c("run_limma")}
if (length(AtacStats) > 0) {
   ## Subset statsDF for rows with at least one hit
   for (iSig in dimnames(AtacStats$statsHitsA)[[3]]) {
      iDF_all <- AtacStats$statsDF[[iSig]];
      hit_cols <- vigrep("^hit ", colnames(iDF_all));
      hits_l <- lapply(hit_cols, function(i){iDF_all[,i] != 0})
      hits_v <- Reduce("|", hits_l);
      #iDF_all_hits <- iDF_all[hits_v,,drop=FALSE];
      iDF_all_hits <- iDF_all;
      
      ## Add nearest promoter df
      pro_cols1 <- provigrep(c("^gene_id",
         "^gene_name",
         "feature_type($|_winner)",
         "_gene",
         "_distance",
         "broad_peak|num_slices"),
         colnames(values(fc_gr)));
      pro_cols <- intersect(
         colnames(values(fc_gr)),
         pro_cols1
      );
      if (length(pro_cols) > 0) {
         idf_match <- match(iDF_all_hits$Gene, values(fc_gr)$Gene);
         promoter_df <- as.data.frame(values(fc_gr[idf_match,pro_cols]));
         iDF_all_hits[,pro_cols] <- promoter_df;
      }
      
      ## Add genome coordinate annotations
      coord_cols <- c("Chr", "Start", "End", "Length");
      coord_df <- data.frame(rowData(
         fc_se[match(iDF_all_hits$Gene, rowData(fc_se)$Geneid),]
         )[,coord_cols]);
      iDF_all_hits2 <- insertDFcols(iDF_all_hits,
         coord_df,
         colnum=2);
      
      txt_ext <- paste0(".stats_nomgm",
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".txt");
      txt_file <- file.path(OUTDIR,
         gsub("[.]fc$|[.]fc[.]txt$",
            txt_ext,
            basename(fc_file)));

      data.table::fwrite(
         x=iDF_all_hits2,
         file=txt_file,
         sep="\t",
         col.names=TRUE);
      printDebug("Wrote ",
         formatInt(nrow(iDF_all_hits2)),
         " rows to: ",
         txt_file);

      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all_hits2);
      bed_ext <- paste0(".stats_nomgm",
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".bed");
      bed_file <- file.path(OUTDIR,
         gsub("[.]fc$|[.]fc[.]txt$",
            bed_ext,
            basename(fc_file)));
      data.table::fwrite(
         x=statsbed,
         file=bed_file,
         sep="\t",
         col.names=FALSE);
   }
}
```

### Export stats results for hits (subset)

Statistical results are saved in text file for the subset
of peaks with max group mean >= **`{r cat(MGM)}`**.
Look in the folder defined by OUTDIR: **`{r cat(OUTDIR)}`**.

A BED file is also saved during this step.

```{r, stats_df_mgm4, dependson=c("run_limma_mgm4")}
if (length(AtacStatsMgm4) > 0) {
   ## Subset statsDF for rows with at least one hit
   for (iSig in dimnames(AtacStatsMgm4$statsHitsA)[[3]]) {
      iDF_all <- AtacStatsMgm4$statsDF[[iSig]];
      hit_cols <- vigrep("^hit ", colnames(iDF_all));
      hits_l <- lapply(hit_cols, function(i){iDF_all[,i] != 0})
      hits_v <- Reduce("|", hits_l);
      #iDF_all_hits <- iDF_all[hits_v,,drop=FALSE];
      iDF_all_hits <- iDF_all;
   
      ## Add nearest promoter df
      pro_cols1 <- provigrep(c("gene_id",
         "gene_name",
         "feature_type",
         "_distance",
         "broad_peak"),
         colnames(values(fc_gr)));
      pro_cols <- intersect(
         colnames(values(fc_gr)),
         pro_cols1
      );
      promoter_df <- as.data.frame(values(fc_gr[iDF_all_hits$Gene])[,pro_cols]);
      iDF_all_hits[,pro_cols] <- promoter_df;
      
      ## Add genome coordinate annotations
      coord_cols <- c("Chr", "Start", "End", "Length");
      coord_df <- data.frame(rowData(
         fc_se[match(iDF_all_hits$Gene, rowData(fc_se)$Geneid),]
         )[,coord_cols]);
      iDF_all_hits2 <- insertDFcols(iDF_all_hits,
         coord_df,
         colnum=2);
      
      txt_ext <- paste0(".stats_mgm",
         MGM,
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".txt");
      txt_file <- file.path(OUTDIR,
         gsub("[.]fc$|[.]fc[.]txt$",
            txt_ext,
            basename(fc_file)));
      
      cat(file="debug_log.txt", txt_file, "\n");
      writeTable(outfile=txt_file,
         outdata=iDF_all_hits2);
      printDebug("Wrote ",
         formatInt(nrow(iDF_all_hits2)),
         " rows to: ",
         txt_file);
      
      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all_hits2);
      bed_ext <- paste0(".stats_mgm",
         MGM,
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".bed");
      bed_file <- file.path(OUTDIR,
         gsub("[.]fc$|[.]fc[.]txt$",
            bed_ext,
            basename(fc_file)));
      data.table::fwrite(
         x=statsbed,
         file=bed_file,
         sep="\t",
         col.names=FALSE);
      
      ## Save all peaks in BED format
      statsbed <- statsdf2bed(iDF_all_hits2, type="hits");
      bed_ext <- paste0(".statshits_mgm",
         MGM,
         ifelse(ATAC, "_atac1", ""),
         ".", iSig, ".bed");
      bed_file <- file.path(OUTDIR,
         gsub("[.]fc$|[.]fc[.]txt$",
            bed_ext,
            basename(fc_file)));
      data.table::fwrite(
         x=statsbed,
         file=bed_file,
         sep="\t",
         col.names=FALSE);

   }
}
```




## Export RData for follow-up analyses


```{r, export_rdata, dependson=c("check_reps", "run_limma", "run_limma_mgm4")}
if (SAVE_RDATA) {
   fc_dat_file <- file.path(OUTDIR,
      gsub("[.]fc$|[.]fc[.]txt$",
         ".RData",
         fc_file));
   if (!igrepHas("RData", fc_dat_file)) {
      fc_dat_file <- paste0(fc_dat_file, ".RData");
   }
   save.image(fc_dat_file);
   printDebug("Saved RData to:",
      fc_dat_file);
} else {
   jamba::printDebug("Skipped saving the RData session, because ",
      "SAVE_RDATA=",
      SAVE_RDATA);
}
```


## Change log

#### 18jul2019

* Added comments to MA-plot of quantile-normalized, batch-adjusted data.
* Added display of MAD values to the above MA-plot, with note saying
outliers are not removed.
* Fixed issue with `genomic_regions` where names like
`"Promoter(-2000,+200)"` were replaced with `"promoter"`.
Now it uses the proper name.
* Note this file saves genome regions to an RData file
`"hg19_refgene_041619_minimus_exons.genome_regions.RData"`.
It does not name the file with promoter and TTS ranges, which
means the ranges are fixed relative to the RData file. In
future if we want to change the ranges around promoters and TTS,
we need to add range numbers to the RData filename.


#### 27aug2019

* Added step to check for replicates per sample group. When there
are no replicates, the whole set of singlets are copied into
a new matrix annotated "Control" to be used to estimate
variability in limma.
* Subsequent steps test for success of each limma contrast
before running each subsequent section.

#### 14nov2019

* Changed maxGroupMean "Mgm4" threshold to be user-configurable
with environmental variable `MGM`, which defaults to `MGM=4`.
* The script `run_gokeyATAC.R` was updated to pass the `MGM`
environmental variable, and includes `_mgm#` in the output file
and directory only if `MGM` is not 4, preserving previous results.

#### 15nov2019

* Changed output to use environmental variable `OUTDIR`, default `"."`
which is the current working directory. The script
`run_gokeyATAC.R` will define `OUTDIR` to use the same subfolder
where the Rmarkdown html file is created.
* Added `CACHEDIR` as optional method of defining a specific cache directory,
the default is the `OUTDIR` with `"cache/"` appended to the end.
* Fixed several issues where plots were labeled "Mgm4" even when `MGM`
was not 4.
* Changed stats output files to check for existing file before writing
a new file.

#### 20nov2019

* Added new section for peak annotation by genome region. This process
annotates all peaks using all genes defined in the GTF file. Genome
regions are defined as promoter, TTS, exon, intron, or intergenic.
First, all peaks are annotated with any intersecting genome region,
along with the corresponding gene(s) where applicable. Next, a
"winner" is chosen among feature_types, in order: promoter, TTS,
exon, intron. Then, the corresponding gene_name(s) for the
feature_type_winner are annotated for each peak. Finally, the distance
to nearest gene, along with the nearest_gene_name are annotated,
which is mostly interesting for peaks in "intergenic" regions.
* Added environment variable `ATAC` which should be 0 or 1 (0 by default).
When ATAC=0 then peaks are not annotated to the nearest ATAC-positive
promoter region. When ATAC=1, then only gene promoters with an ATAC peak
overlapping the promoter are used for annotation -- intended to subset
the total set of all possible genes, to the subset of genes for which
data support their active expression.

#### 21nov2019

* New optional environmental variable `GTF` with file path to
the GTF file that should be used when annotating peaks to genes
and nearest genes. Note that changing this file invalidates the cache,
causing the whole pipeline to be repeated.
* The script will now save the stats files even if they previously exist,
in order to update the peak annotations.

#### 13dec2019

* Add "gene_id" to set of columns retained by genomic_regions,
and include "gene_id" in subsequent output columns.
* Change in nearest Mgm4 promoter logic to use TSS of promoters overlapping
Mgm peaks.

   * take promoters_gr that overlap Mgm4 peaks, call them mgm_promoters_gr
   * convert mgm_promoters_gr to mgm_tss_gr (adjust back to original TSS position)
   * determine distance to nearest mgm_tss_gr per fc_gr peak
   * use `distanceToNearest(..., 'select="all"')`
   to return all hits with same distance
   * use column headers "mgm4_annoTSS_gene_name", "mgm4_annoTSS_gene_id", "mgm4_annoTSS_distance"

#### 02jan2020

* Ideas for naming this tool: The tool stacks layers of peaks, merges
overlapping peaks, slices the merged peaks by max width, then compares
peak depth across layers using differential analysis.
* peak, merge, slice, diff, comp
* geologic terms: stratum/strata; mesa; slate; mesa; plateau
* divergent

#### 06jan2020

* Fixed bug in `peak_genes` caused by syntax differences that work with
`data.frame` and `matrix`, but not `data.table` objects. Also made the
peak annotation steps markedly faster by only running the step to
comma-delimit multiple values for the subset of rows that require that
logic.

#### 03feb2020

* Changed the volcano plots with MGM threshold to use the actual
`MGM` threshold, previously they used MGM=4 which caused minor
discrepancies between the reported hits from limma, and the volcano
plot numbers. This change removes that discrepancy.

#### 17apr2020

* Initial migration to Github.
* Changed instances of `CellLine` to `Group`.

#### 29apr2020

* use `distanceToNearest(..., 'select="all"', ignore.strand=TRUE)`
to return all hits with same distance, and to return hits from either strand
* use `subsetByOverlaps(..., ignore.strand=TRUE)` to return overlaps
from either strand.
* use `findOverlaps(..., ignore.strand=TRUE)` to return overlaps
from either strand.
* Export BED files for all peaks, MGM-filtered peaks, and MGM peak hits.
* Enable use of `"curation.yaml"` file to convert featureCounts column
names to Sample, Group, Run columns used for sample grouping
and batch adjustment.
* Added optional GROUPCHECK which allows reviewing the sample grouping
without running the rest of the pipeline.
