---
title: "Slicejam Analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
    df_print: kable
    self_contained: true
    keep_md: true
---
   
About this document:

* prepared on **`r format(Sys.time(), "%a %b %d %X %Y")`**
* using **`r R.version.string`**
* R slicejam package version **`r packageVersion("slicejam")`**.

### R environment setup

This document can be reproduced using the R code displayed in
this document. The initial steps involve loading R packages
into memory.

#### Rmarkdown options

```{r, setup, include=TRUE}
#    css: liao_style.css
## To render this page:
## In the farrisSeq folder run:
## rmarkdown::render("Adam19_Rnaseq_Analysis_vM21.Rmd", "html_document", knit_root_dir=getwd(), output_dir=getwd());
##    css: /Users/wardjm/Projects/R-scripts/liao_style.css

## Define the cache so it is dependent upon:
## - the versions of a few R packages,
## - the environmental variable FC_FILE
knitr::opts_chunk$set(cache=TRUE,
   autodep=TRUE,
   dev=c("png", "cairo_pdf"),
   dev.args=list(
      pointsize=14
   ),
   fig.align="center",
   fig.height=9,
   fig.width=8,
   pdf.options(encoding="ISOLatin9.enc"),
   cache.path=Sys.getenv("CACHEDIR", "cache"),
   cache.lazy=FALSE,
   cache.extra=list(
      packageVersion("jamba"),
      packageVersion("jamma"),
      packageVersion("platjam"),
      packageVersion("slicejam"),
      packageVersion("jamses"),
      paste0("CURATION_TXT:", Sys.getenv("CURATION_TXT"))
   )
);

# add to prevent transparent background for PNG output
par("bg"="white");

options("kable_styling_position"="left",
   "kable_styling_full_width"=FALSE,
   "kable_styling_bootstrap_options"=c("striped",
      "hover",
      "condensed",
      "responsive")
);
## Set font to Arial which helps compatibility in Adobe Illustrator,
## and helps display of up and down arrow characters in Venn diagrams
par("family"="Arial",
   "lend"="square",
   "ljoin"="mitre");
```

#### SliceJam Parameters

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseSlicejam1" role="button" aria-expanded="false" aria-controls="collapseSlicejam1">(Expand...)</a></p>
<div class="collapse" id="collapseSlicejam1"><div class="card card-body">

The parameters used for this analysis are defined below.

```{r, slicejam_minimum_version}
# check version of slicejam
min_slicejam_version <- "0.0.37.900";
if (compareVersion(as.character(packageVersion("slicejam")), "0.0.37.900") < 0) {
   jamba::kable_coloring(
   data.frame(check.names=FALSE,
      stringsAsFactors=FALSE,
      `.libPaths()`=.libPaths()),
   caption=paste0(
      "R library paths in use."));

   stop(paste0("Slicejam must be at least slicejam-", min_slicejam_version,
      ", please verify .libPaths():\n",
      jamba::cPaste(paste0("'", .libPaths(), "'"), sep="\n")));
}
```


```{r, slicejam_args, include=TRUE, cache=FALSE}
sliceargs <- slicejam::get_slicejam_args(verbose=TRUE,
   name="analysis",
   type="data.frame");
if ("sliceargs_df" %in% names(attributes(sliceargs))) {
   sliceargs_df <- attr(sliceargs, "sliceargs_df");
   suppressPackageStartupMessages(library(kableExtra));
   jamba::kable_coloring(
      sliceargs_df,
      caption=paste0(
         "slicejam arguments:"));
}

DRYRUN <- sliceargs$DRYRUN;
ATAC <- sliceargs$ATAC;
fc_file <- sliceargs$fc_file;
fc_filepath <- sliceargs$fc_filepath;
CURATION_TXT <- sliceargs$CURATION_TXT;
GTF <- sliceargs$GTF;
GTFNAME <- sliceargs$GTFNAME;
MASK <- sliceargs$MASK;
DRYRUN <- sliceargs$DRYRUN;
OUTDIR <- sliceargs$OUTDIR;
CACHEDIR <- sliceargs$CACHEDIR;
ATAC <- sliceargs$ATAC;
MGM <- sliceargs$MGM;
NORM <- sliceargs$NORM;
NORMMIN <- sliceargs$NORMMIN;
NORMSHORT <- sliceargs$NORMSHORT;
GROUPCHECK <- sliceargs$GROUPCHECK;
FACTOR_ORDER <- sliceargs$FACTOR_ORDER;
SAVE_RDATA <- sliceargs$SAVE_RDATA;

detectedTx <- sliceargs$DETECTED_TX;
detectedGenes <- sliceargs$DETECTED_GENES;

upstream_promoter <- sliceargs$UPSTREAM_PROMOTER;
downstream_promoter <- sliceargs$DOWNSTREAM_PROMOTER;
upstream_tts <- sliceargs$UPSTREAM_TTS;
downstream_tts <- sliceargs$DOWNSTREAM_TTS;

fc_base <- sliceargs$fc_base;
fc_basedir <- sliceargs$fc_basedir;
fc_html <- sliceargs$fc_html;
knit_root_dir <- sliceargs$knit_root_dir;

```


#### R package initialization


```{r, r_initialization, cache=FALSE}
library(pacman)
suppressPackageStartupMessages(p_load_gh("jmw86069/jamba"));

jamba::kable_coloring(
   data.frame(check.names=FALSE,
      stringsAsFactors=FALSE,
      `.libPaths()`=.libPaths()),
   caption=paste0(
      "R library paths in use."));

options("warn"=-1);
suppressPackageStartupMessages(p_load_gh("jmw86069/colorjam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/slicejam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/jamses"));
suppressPackageStartupMessages(p_load_gh("jmw86069/jamma"));
suppressPackageStartupMessages(p_load_gh("jmw86069/splicejam"));
suppressPackageStartupMessages(p_load_gh("jmw86069/multienrichjam"));

suppressPackageStartupMessages(p_load("SummarizedExperiment"));
suppressPackageStartupMessages(p_load("data.table"));
suppressPackageStartupMessages(p_load("matrixStats"));
suppressPackageStartupMessages(p_load("extrafont"));
#suppressPackageStartupMessages(library(GenomicFeatures));
suppressPackageStartupMessages(p_load("rtracklayer"));
suppressPackageStartupMessages(p_load(methods));
suppressPackageStartupMessages(p_load("kableExtra"));

```

</div></div>

## Description of workflow

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseDescription1" role="button" aria-expanded="false" aria-controls="collapseDescription1">(Expand...)</a></p>
<div class="collapse" id="collapseDescription1"><div class="card card-body">

* Import the first `.fc` file found in the current working directory.

    * Optionally use "FC_FILE" environmental variable.

* Curate filename columns by one of two processes:

    * A file defined by `"CURATION_TXT"` is used to parse the colnames
    in the featureCounts file into three columns: `"Sample"`, `"Group"`, `"Run"`

* An R object `SummarizedExperiment` is created, using the columns
from featureCounts to populate rows, such as
`"Geneid", "Chr", "Start", "End", "Strand", "Length"`.
* Assign categorical colors to the unique values in the `"Group"` column.

* Normalization and batch adjustment

    * Normalize peak counts across all samples using the method defined by `NORM`.
    * When `"Run"` contains more than one unique value,
    apply batch adjustment using the `"Run"` column to define batch,
    and `"Group"` to define groups, using normalized data.

* Create MA-plots:

    * un-normalized data, global-centered
    * un-normalized data, centered by "Run"
    * normalized data, global-centered
    * normalized data, centered by "Run"
    * normalized data, centered by "Group","Run"
    * normalizeddata, global-centered, batch-adjusted by "Run"

* Create sample correlation heatmaps:

   * normalized data, centered by the first factor level
   in "Group", intended to act as a control group.
   * normalized data, batch-adjusted by "Run", centered by the
   first factor level in "Group".

* Statistical comparisons

   * Use the `"Group"` column to define groups, design, and contrasts.
   
       * Note that contrasts are limited to comparisons to the first
       `"Group"` level.
   
   * Use normalized, batch-adjusted counts.
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05
   * Apply secondary max group mean filtering with MGM=**`r MGM`**, to
   reduce the number of peaks used for statistical testing
   * Apply cutoffs: maxGroupMean >= 4; fold change >= 1.5; adjusted P-value <= 0.05

* Create volcano plots

    * Plot each contrast using the P-value, then using adjusted P-value.

* Create Venn diagrams:

    * Include each set of two contrasts with "all peak" hits.
    * Include each set of two contrasts with "MGM-filtered peak" hits.

* Annotate peaks

    * Associate each peak with one or more overlapping feature_types
    * Call one feature_type_winner per peak, based upon order: promoter, TTS, exon, intron, extragenic
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "all peaks" comparison
    * Create pie charts showing distribution of feature_type for differential
    peaks from the "MGM-filtered peaks" comparison

* Re-annotate peaks into broad_peaks by merging adjacent slices

    * Count broad_peak hits from the "all peaks" comparison.
    * Count broad_peak hits from the "MGM-filtered peaks" comparison.

* Heatmap of peak hits across samples

    * Heatmap of peak hits from the "all peaks" comparison.
    * Heatmap of peak hits from the "MGM-filtered peaks" comparison.

* Optional ATAC-mode annotation to "active promoter genes"

    * Define active promoter genes by presence of ATAC-seq peak in promoter region.
    * Annotate peaks by nearest active promoter peak, along with distance.

* Save stats results to text files

    * Save "all peak" stats to tab-delimited text file.
    * Save "MGM-filtered peak" stats to tab-delimited text file.
    * This file is also used to create the "table of Venn intersections."

* Optionally save RData session

</div></div>


## Analysis workflow


### Import featureCounts data

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseImport1" role="button" aria-expanded="false" aria-controls="collapseImport1">(Expand...)</a></p>
<div class="collapse" id="collapseImport1"><div class="card card-body">

```{r, import_fc}
## Import
fc_data <- import_featurecounts(fc_file);

## Add zero-padding to peak labels
fc_data[,1] <- paste0("peak_",
   jamba::padInteger(seq_len(nrow(fc_data))));
rownames(fc_data) <- fc_data[,1];

# featureCounts colnames
fc_colnames <- attr(fc_data, "fc_colnames");
filenames <- attr(fc_data, "filenames");
if (length(fc_colnames) == 0) {
   stop(paste0("None of the featureCounts colnames were detected: ",
      jamba::cPaste(fc_colnames)));
}


## Use curation_txt if that file exists
curation_txt <- CURATION_TXT;
curation_df <- data.table::fread(curation_txt,
   data.table=FALSE,
   sep="\t");
filename_colname <- "Filename";
sample_df <- slicejam::curate_to_df_by_pattern(
   x=filenames,
   input_colname=filename_colname,
   df=curation_df);
if (nrow(sample_df) == 0) {
   jamba::printDebug("Error: ",
      c("The curation_txt file '",
         curation_txt,
         "' did not match filenames."),
      sep="",
      fgText=c("red", "dodgerblue"));
   jamba::printDebug("filenames: ",
      paste0("\n'", filenames, "'"));
   jamba::printDebug("curation_txt contents:");
   print(curation_df);
   stop("curation_txt file did not match filenames.");
}

## Check for samples not curated, and therefore dropped from the analysis
if (length(sample_df[[filename_colname]]) < length(filenames)) {
   samples_dropped <- setdiff(filenames,
      sample_df[[filename_colname]]);
   samples_retained <- sample_df[[filename_colname]];
   samples_retmax <- max(c(
      length(samples_dropped),
      length(samples_retained)));
   samples_dropped <- unlist(rep(list(samples_dropped, ""),
      c(1, samples_retmax - length(samples_dropped))));
   samples_retained <- unlist(rep(list(samples_retained, ""),
      c(1, samples_retmax - length(samples_retained))));
   
   kable_coloring(
      data.frame(
         samples_retained=samples_retained,
         samples_dropped=samples_dropped),
      row.names=FALSE,
      caption=paste0(
         "Samples retained and dropped."));
}

sample_colname <- head(jamba::provigrep(
   c("^label$",
      "^sample$",
      "^sample.*(name|id)$"),
   colnames(sample_df)), 1);
if (length(sample_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Label","' nor '", "Sample", "' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Sample column was recognized.");
}
sample_df[[sample_colname]] <- jamba::makeNames(sample_df[[sample_colname]]);
rownames(sample_df) <- sample_df[[sample_colname]];

run_colname <- head(jamba::provigrep(
   c("^batch$",
      "^run$"),
   colnames(sample_df)), 1);
if (length(run_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Run","' or '","Batch","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
}

group_colname <- head(jamba::provigrep(
   c("^Group$",
      "^group.*name$",
      "group$"),
   colnames(sample_df)), 1);
if (length(group_colname) == 0) {
   jamba::printDebug("Error: ",
      c("No '","Group","' column was recognized."),
      sep="",
      fgText=c("red", "dodgerblue"));
   stop("No Group column was recognized.");
}
## Ensure group names have no invalid characters like "-"
sample_df[[group_colname]] <- gsub("[%]", "pct",
   gsub("[- ]+",
      ".",
      sample_df[[group_colname]]));
sample_df[[group_colname]] <- factor(sample_df[[group_colname]],
   levels=unique(sample_df[[group_colname]]));

normgroup_colname <- head(jamba::provigrep(
   c("^normgroup$",
      "norm.*group"),
   colnames(sample_df)), 1);
if (length(normgroup_colname) == 0) {
   normgroup_colname <- "normgroup";
   sample_df[[normgroup_colname]] <- sample_df[[group_colname]];
}

## Optional color assignment
color_colname <- head(jamba::provigrep(
   c("^Color$", "Color$"),
   colnames(sample_df)), 1);

## Sort sample_df by group
sample_df <- jamba::mixedSortDF(sample_df,
   byCol=c(
      normgroup_colname,
      group_colname,
      run_colname,
      sample_colname));

## Define useful sample annotation vectors
iSamples <- sample_df[[sample_colname]];
Run <- jamba::nameVector(sample_df[[run_colname]],
   iSamples);
iGroup <- jamba::nameVector(sample_df[[group_colname]],
   iSamples);
iNormGroup <- jamba::nameVector(sample_df[[normgroup_colname]],
   iSamples);

## Make sure the order in filenames is propagated
filenames <- jamba::nameVector(sample_df[[filename_colname]],
   iSamples);
fc_data <- fc_data[,c(fc_colnames, filenames),drop=FALSE];

## Print sample annotation table
curation_color_list <- platjam::design2colors(
   plot_type="none",
   sample_df)
kable_coloring(sample_df,
   colorSub=curation_color_list,
   caption=paste0(
      "Sample Groups defined by curation_txt: '",
      basename(curation_txt),
      "'"));
if (any(!sapply(c("iSamples", "Run", "iGroup", "iNormGroup"), exists))) {
   stop("One or more of iSamples,Run,Group could not be defined.");
}

## Rename featureCounts columns to match iSamples
kable_coloring(
   data.frame(
      Rename_from=filenames,
      Rename_to=iSamples),
   row.names=FALSE,
   caption=paste0(
      "Renamed featureCounts colnames."));
fc_data <- jamba::renameColumn(fc_data,
   from=filenames,
   to=iSamples);

## Make SummarizedExperiment
fc_se <- SummarizedExperiment(
   assays=list(counts=log2(1+as.matrix(fc_data[,iSamples,drop=FALSE]))),
   rowData=fc_data[,fc_colnames,drop=FALSE],
   colData=sample_df);

## Re-order samples by Group
iSamples <- mixedSortDF(colData(fc_se),
   byCols=group_colname)[[sample_colname]];
fc_se <- fc_se[,iSamples];
Run <- nameVector(colData(fc_se)[[run_colname]],
   colnames(fc_se));
iGroup <- nameVector(colData(fc_se)[[group_colname]],
   colnames(fc_se));
```

</div></div>

#### Assign colors

Base filename: `define_colors`

```{r define_colors, fig.width=16, fig.height=9}
## Define sample colors
color_sub <- NULL;
if (length(color_colname) == 1) {
   # assume group color assignment
   colordf <- unique(data.frame(check.names=FALSE,
      colData(fc_se)[,c(group_colname, color_colname), drop=FALSE]));
   if (length(jamba::tcount(colordf[[group_colname]], 2)) == 0) {
      # no duplicate groups, therefore these are group colors
      color_sub <- jamba::nameVector(
         colordf[,c(color_colname, group_colname)]);
   } else {
      # these must be sample colors
      colordf <- data.frame(check.names=FALSE,
         colData(fc_se)[,c(sample_colname, color_colname), drop=FALSE]);
      color_sub <- jamba::nameVector(
         colordf[,c(color_colname, sample_colname)]);
   }
}
sample_color_list <- platjam::design2colors(
   data.frame(check.names=FALSE,
      colData(fc_se)),
   group_colnames=group_colname,
   # lightness_colnames=run_colname,
   end_hue_pad=0,
   color_sub=color_sub)

group_colors <- sample_color_list[[group_colname]];
run_colors <- sample_color_list[[run_colname]];

color_sub1 <- nameVector(group_colors[colData(fc_se)[[group_colname]]],
   colData(fc_se)[[sample_colname]]);

run_colors1 <- nameVector(run_colors[colData(fc_se[,iSamples])[[run_colname]]],
   iSamples);
```

#### Statistical Contrasts

Base filename: `define_contrasts`

```{r, define_contrasts, fig.height=12, fig.width=12}
sample_txt_file <- file.path(OUTDIR,
   "curated_samples.txt");
data.table::fwrite(
   x=sample_df,
   file=sample_txt_file,
   sep="\t",
   col.names=TRUE);

## define experimental design groups and contrasts
sedesign <- jamses::groups_to_sedesign(
   data.frame(check.names=FALSE,
      colData(fc_se[, iSamples])[, group_colname, drop=FALSE]),
   factor_order=FACTOR_ORDER,
   max_depth=2);
iDesign <- sedesign@design;
iContrasts <- sedesign@contrasts;

contrastNames <- NULL;
if (length(iContrasts) > 0) {
   contrastNames <- colnames(iContrasts);
}
contrast_txt_file <- file.path(OUTDIR,
   "contrasts.txt");
kc <- NULL;
if (file.exists(contrast_txt_file)) {
   contrastNames <- gsub("^[ \t\r\n]+|[ \t\r\n]+$",
      "",
      readLines(contrast_txt_file));
   iContrasts <- tryCatch({limma::makeContrasts(
         contrasts=contrastNames,
         levels=iDesign);
   }, error=function(e){
      jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
         contrast_txt_file);
      print(e);
      NULL;
   })
   if (length(iContrasts) > 0) {
      rownames(iContrasts) <- colnames(iDesign);
      sedesign@contrasts <- iContrasts;
      kc <- kable_coloring(
         data.frame(check.names=FALSE,
            Note="contrastNames were loaded from file: ",
            File=paste0("'", contrast_txt_file, "'")),
         row.names=FALSE);
   } else {
      contrastNames <- NULL;
      iContrasts <- NULL;
   }
} else {
   if (length(contrastNames) > 0) {
      writeLines(text=contrastNames,
         con=contrast_txt_file);
      kc <- kable_coloring(
         data.frame(check.names=FALSE,
            Note="contrastNames were stored in file: ",
            File=paste0("'", contrast_txt_file, "'")),
         row.names=FALSE);
   } else {
      kc <- kable_coloring(
         data.frame(Note=paste0("There are no contrasts defined."),
         row.names=FALSE));
   }
}
kc;

kable_coloring(
   data.frame(
      `Contrast Names`=contrastNames,
      `Abbreviated Comp`=jamses::contrast2comp(contrastNames)),
   row.names=FALSE,
   caption=paste0(
      "Contrasts defined:"));

# plot sedesign
plot_sedesign(sedesign,
   arrow_ex=0.7,
   axis1=2,
   axis2=1,
   axis3=3,
   axis4=4)

## Optional GROUPCHECK which exits here to allow reviewing the sample grouping
if (GROUPCHECK) {
   #txt_ext <- paste0(".curated_samples.txt");
   #sample_txt_file <- file.path(OUTDIR,
   #   gsub("[.]fc$|[.]fc[.]txt$",
   #      txt_ext,
   #      basename(fc_file)));
   printDebug("Curated sample data: '",
      sample_txt_file, "'");
   printDebug("Contrast names: '",
      contrast_txt_file, "'");
   knitr::knit_exit(fully=TRUE);
}

```

### Annotate Peaks 

#### Peak Sizes {.tabset}

If the featureCounts output included column name `"Length"`,
the plots below will show the distribution of lengths.

##### Distribution

Base filename: `peak_sizes`

```{r, peak_sizes, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes",
      xaxt="n",
      yaxt="n",
      breaks=100);
   title(xlab="Peak width", line=4)
   axis1at <- pretty(par("usr")[1:2])
   axis(1,
      las=2,
      cex.axis=1.3,
      at=axis1at,
      labels=formatInt(axis1at))
   axis2at <- pretty(par("usr")[3:4])
   axis(2,
      las=2,
      cex.axis=1.3,
      at=axis2at,
      labels=formatInt(axis2at))
}
```

##### Distribution (log)

Base filename: `peak_sizes_log`

```{r, peak_sizes_log, fig.height=6, fig.width=6}
if ("Length" %in% colnames(rowData(fc_se))) {
   plotPolygonDensity(rowData(fc_se)$Length,
      main="Peak sizes (log-scaled x-axis)",
      log="x",
      xlab="Peak width",
      breaks=100);
}
```

#### MGM Threshold

The max group mean (MGM) threshold is defined by the environmental
variable `MGM`, used to customize the threshold during the stats
tests, applied to signal log2-transformed with `log2(1 + x)`,
and therefore the default `MGM=4` defines a threshold of `15 counts`
in normal space, using `( 2^MGM ) - 1` to reverse the
transform: `log2(1 + x)`


```{r, mgm_arg}
## Get maxGroupMean threshold argument
if (length(MGM) == 0) {
   MGM <- c(`4`=4);
}
if (!is.numeric(MGM)) {
   MGM <- jamba::rmNA(
      as.numeric(unlist(strsplit(MGM, "[ ,]+")))
   );
}
names(MGM) <- as.character(MGM);
kable_coloring(
   data.frame(check.names=FALSE,
      `max group mean threshold (MGM)`=MGM),
   row.names=FALSE);
```


### Data Normalization

#### Apply Normalization

```{r, apply_norm, dependson="import_fc"}
kable_coloring(
   data.frame(check.names=FALSE,
      `normalization method (NORM)`=NORM),
   row.names=FALSE);

## Raw data
iMatrix <- assays(fc_se[,iSamples])[["counts"]];

## Perform normalization
if ("median" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("mediangroup" %in% NORM) {
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("jammanorm"),
      param=list(
         jammanorm=list(
            minimum_mean=NORMMIN,
            centerGroups=iNormGroup,
            controlSamples=iSamples,
            useMean=FALSE)
      )
   );
   normname <- "jammanorm_counts";
} else if ("quantile" %in% NORM) {
   ## Evaluate quantile normalization
   fc_se <- jamses::se_normalize(fc_se,
      assay_names="counts",
      method=c("quantile"));
   normname <- "quantile_counts";
} else if ("none" %in% NORM) {
   normname <- "counts";
}

## Normalized data
iMatrixQ <- assays(fc_se[,iSamples])[[normname]];

```

#### Batch Adjustment

When there are multiple values in the `"Run"` column,
batch adjustment is performed using `limma::removeBatchEffect()`,
solely for data visualization.
Batch is defined by `"Run"`, with group defined by `"Group"`.

```{r, batch_adjust_run, dependson="apply_norm"}
## Batch adjustment
batch_adjusted <- FALSE;
if (length(unique(colData(fc_se[,iSamples])[[run_colname]])) > 1) {
   fc_batch <- nameVector(colData(fc_se[,iSamples])[[run_colname]],
      iSamples);
   fc_group <- nameVector(colData(fc_se[,iSamples])[[group_colname]],
      iSamples);
   fc_se_b <- tryCatch({
      jamses::se_normalize(fc_se[,iSamples],
         assay_names=unique(c("counts", normname)),
         method=c("limma_batch_adjust"),
         params=list(
            limma_batch_adjust=list(
               batch=fc_batch[iSamples],
               group=fc_group[iSamples]))
      );
   }, error=function(e){
      printDebug("Error during batch adjustment by se_normalize():");
      print(e);
      NULL;
   });
   bnormname <- paste0("limma_batch_adjust_", normname);
   if (length(fc_se_b) > 0) {
      batch_adjusted <- TRUE;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se_b[,iSamples])[["limma_batch_adjust_counts"]];
      ## quantile-normalized data, batch-adjusted
      iMatrixQB <- assays(fc_se_b[,iSamples])[[bnormname]];
   } else {
      ## Do not perform batch adjustment
      fc_se_b <- fc_se;
      ## batch-adjusted raw data
      iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
      ## quantile-normalized data, batch-adjusted
      bnormname <- normname;
      iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
   }
} else {
   ## Do not perform batch adjustment
   fc_se_b <- fc_se;
   ## batch-adjusted raw data
   iMatrixB <- assays(fc_se[,iSamples])[["counts"]];
   ## quantile-normalized data, batch-adjusted
   bnormname <- normname;
   iMatrixQB <- assays(fc_se[,iSamples])[[bnormname]];
}


```


#### Detected Peaks

This section defines "detected peaks" for downstream analysis,
which requires the max group mean in any one group should be
at least `MGM` at each threshold defined.
This step should be performed using normalized data if possible.

```{r detected_rows}
# calculate row group means
iMatrixQBgrp <- rowGroupMeans(
   assays(fc_se_b[,iSamples])[[bnormname]],
   groups=colData(fc_se[,iSamples])$Group,
   useMedian=FALSE)
mgm_values <- apply(iMatrixQBgrp, 1, max, na.rm=TRUE)
names(mgm_values) <- rownames(iMatrixQBgrp);

# generate a list so we can handle multiple thresholds
iPeaks_list <- lapply(MGM, function(imgm) {
   iPeaksMgm <- names(mgm_values)[(!is.na(mgm_values) &
         mgm_values >= imgm)];
})
names(iPeaks_list) <- paste0("MGM >= ", MGM);
mgm_list <- nameVector(MGM, names(iPeaks_list))

# print table summary
filtered_to <- lengths(iPeaks_list);
mgm_n <- length(mgm_values);
filtered_df <- data.frame(check.names=FALSE,
   `Peak Set`=c("All peaks",
      names(iPeaks_list)),
   `Size`=c(mgm_n,
      lengths(iPeaks_list)),
   Percentage=paste0(
      round(100 * c(mgm_n,
         lengths(iPeaks_list)) / mgm_n), "%"))
kable_coloring(
   filtered_df,
   row.names=FALSE,
   caption=paste0(
      "Result of MGM filtering."))

# default uses the first (non-empty) vector in the list
iPeaks_list <- rmNULL(iPeaks_list)
if (length(iPeaks_list) == 0) {
   iPeaksMgm <- rownames(fc_se);
} else {
   iPeaksMgm <- iPeaks_list[[1]];
}

```

### MA-Plots {.tabset}


```{r, ma_layout, dependson="apply_norm"}
## Determine blank plot positions to organize Group and Run
blank_pos <- NULL;
ma_ncol <- ceiling(sqrt(ncol(fc_se)));
ma_nrow <- ceiling(sqrt(ncol(fc_se)));
tryCatch({
   Group_ctm <- as.matrix(table(
      data.frame(check.names=FALSE,
         colData(fc_se)[,c(group_colname, run_colname)])));
   sub_n <- apply(Group_ctm, 2, function(i){
      max(i, na.rm=TRUE)});
   ctm_n <- matrix(cumsum(rep(sub_n, nrow(Group_ctm))),
      ncol=ncol(Group_ctm),
      byrow=TRUE);
   #ctm_n;
   ma_ncol <- sum(sub_n);
   ma_nrow <- nrow(Group_ctm);
   blank_pos <- unlist(lapply(seq_len(nrow(Group_ctm)), function(i){
      jamba::rmNULL(lapply(seq_len(ncol(Group_ctm)), function(j){
         iseq <- seq(from=ctm_n[i,j], by=-1, length.out=sub_n[j]);
         head(iseq, sub_n[j] - Group_ctm[i,j])
      }))
   }));
}, error=function(e){
   print(e);
})
ma_height <- 5 + sqrt(ma_nrow) * 4;
ma_width <- 5 + sqrt(ma_ncol) * 4.5;
```

Base filename: `ma_plots`

```{r ma_plots, fig.height=ma_height, fig.width=ma_width, results='asis'}

# signals to include
signal_names <- unique(c("counts", normname, bnormname));


# peaks to use
ipeaks_names <- c(list(
   `All Peaks`=rownames(fc_se)),
   iPeaks_list)

# centering and controls
centerby_list <- list(
   # `centered by first group`=NULL,
   `global-centered`=NULL,
   `centered by Run`="Run")
# control samples in the first group
iControl <- head(levels(iGroup), 1);
iControlSamples <- iSamples[
   colData(fc_se[, iSamples])[[group_colname]] %in% iControl]
centering_controls <- list(
   # `centered by first group`=iControlSamples,
   `global-centered`=iSamples,
   `centered by Run`=iSamples)


for (ipeaks_name in names(ipeaks_names)) {
   ipeaks_use <- ipeaks_names[[ipeaks_name]];
   cat(paste0("\n\n#### ", ipeaks_name, " {.tabset}\n\n"));
   for (signal_name in signal_names) {
      cat(paste0("\n\n##### Signal: ", signal_name, " {.tabset}\n\n"));
      for (centering_name in names(centerby_list)) {
         cat(paste0("\n\n###### ", centering_name, "\n\n"));
         # determine centering
         centerGroups <- NULL;
         centerby_colnames <- centerby_list[[centering_name]];
         if (length(centerby_colnames) > 0) {
            centerGroups <- pasteByRow(data.frame(check.names=FALSE,
               colData(fc_se_b[, iSamples])[, centerby_colnames, drop=FALSE]))
            names(centerGroups) <- iSamples;
         }
         
         iControlSamples <- centering_controls[[centering_name]];
         # create MA-plots
         jp_attr <- attributes(jammaplot(
            fc_se_b[ipeaks_use, iSamples],
            assay_name=signal_name,
            controlSamples=iControlSamples,
            centerGroups=centerGroups,
            noise_floor=1,
            noise_floor_value=NA,
            ncol=ma_ncol,
            nrow=ma_nrow,
            blankPlotPos=blank_pos,
            maintitle=paste0("Signal: ", signal_name,
               "Peaks: ", ipeaks_name,
               "\n", centering_name),
            groupSuffix="",
            subtitlePreset="bottomleft",
            subtitle=Run[iSamples],
            subtitleBoxColor=run_colors1[iSamples],
            useMean=TRUE,
            displayMAD=TRUE,
            outlierMAD=2,
            ablineV=MGM,
            titleBoxColor=color_sub1[iSamples]
         ))
      }
   }
}

```


### Sample outliers

Initially, all samples are used, but this section is a placeholder
for outlier detection and removal.

```{r, sample_outliers}
## Optional subset of samples (e.g. MA-plot outliers)
iSamplesUse <- iSamples;
iSamplesUseList <- split(iSamplesUse, SummarizedExperiment::colData(fc_se[, iSamplesUse])$Group)
maxlen <- max(lengths(iSamplesUseList), na.rm=TRUE);
iSamplesUseList <- lapply(iSamplesUseList, function(i){
   c(i, rep("", length.out=maxlen - length(i)))
})
iSamplesUse_df <- as.data.frame(iSamplesUseList)

kable_coloring(
   caption="Samples used for downstream analysis, organized by group.",
   iSamplesUse_df,
   row.names=FALSE,
   colorSub=sample_color_list)
```

### Sample Correlation {.tabset}

Note that only detected rows with the first MGM threshold were used
to calculate sample correlation.

Base filename: `correlation_heatmaps`

```{r correlation_heatmaps, fig.height=12, fig.width=12, results='asis'}
# Goals:
# - iterate rows: iPeaks_list
# - iterate signal: raw, normalized, batch-adjusted
# - iterate centering: global, group-centered

# peaks to use
ipeaks_names <- c(
   # list(`All Peaks`=rownames(fc_se)),
   iPeaks_list)
# only one signal is used when batch-adjustment was not performed
signal_names <- unique(c(normname, bnormname))
# centering and controls
centerby_list <- list(
   `centered by first group`=NULL,
   `global-centered`=NULL,
   `centered by Run`="Run")
# control samples in the first group
iControl <- head(levels(iGroup), 1);
iControlSamples <- iSamplesUse[
   colData(fc_se[, iSamplesUse])[[group_colname]] %in% iControl]
centering_controls <- list(
   `centered by first group`=iControlSamples,
   `global-centered`=iSamplesUse,
   `centered by Run`=iSamplesUse)

for (ipeaks_name in names(ipeaks_names)) {
   ipeaks_use <- ipeaks_names[[ipeaks_name]];
   cat(paste0("\n\n#### ", ipeaks_name, " {.tabset}\n\n"));
   for (signal_name in signal_names) {
      cat(paste0("\n\n##### Signal: ", signal_name, "  {.tabset}\n\n"));
      for (centering_name in names(centerby_list)) {
         cat(paste0("\n\n###### ", centering_name, "\n\n"));
         # determine centering
         centerby_colnames <- centerby_list[[centering_name]];
         iControlSamples <- centering_controls[[centering_name]];
         # generate heatmap
         hm_cor <- heatmap_se(
            fc_se_b[,iSamples],
            rows=ipeaks_use,
            assay_name=signal_name,
            correlation=TRUE,
            top_colnames=c("Run", "Group"),
            centerby_colnames=centerby_colnames,
            controlSamples=iControlSamples,
            # row_type=paste0("hits at: ",
            #    gsub("^hit ", "", hm_cutoff)),
            data_type="signal",
            color_max=1,
            # control_label="versus control samples",
            show_row_names=FALSE,
            sample_color_list=sample_color_list,
            column_cex=0.4,
            column_split=c("Group"),
            row_split=c("Group"),
            column_title_rot=90)
         ComplexHeatmap::draw(hm_cor,
            column_title=attr(hm_cor, "hm_title"),
            merge_legends=TRUE)
      }
   }
}

```


### Peak Width by Signal {.tabset}

This section is intended to help evaluate peak width, and peak slice width
when using `mergeSplitCountBed`.

A vertical dotted line is drawn to indicate the MGM threshold.

Base filename: `peak_width_plots`

```{r peak_width_plots, fig.height=ma_height*1.5, fig.width=ma_width*1.5, results='asis'}
# only one signal is used when batch-adjustment was not performed
ipeaks_names <- c(
   list(`All Peaks`=rownames(fc_se)),
   iPeaks_list)
signal_names <- unique(c("counts", normname, bnormname));

mgm_list[["All Peaks"]] <- 0;

for (ipeaks_name in names(ipeaks_names)) {
   ipeaks_use <- ipeaks_names[[ipeaks_name]];
   mgm_cutoff <- mgm_list[[ipeaks_name]];
   cat(paste0("\n\n#### ", ipeaks_name, " {.tabset}\n\n"));
   for (signal_name in signal_names) {
      cat(paste0("\n\n##### Signal: ", signal_name, "\n\n"));
      # do plot
      total_n <- length(iSamples) + length(blank_pos);
      ma_nrow <- ceiling(total_n / ma_ncol);
      par(mfrow=c(ma_nrow, ma_ncol));
      par(mar=c(5, 4, 2, 2) + 0.1);
      panel_num <- 0;
      for (iCol in iSamples) {
         panel_num <- panel_num + 1;
         if (panel_num %in% blank_pos) {
            next;
         }
         iRows <- rownames(iMatrix);
         ix <- noiseFloor(
            as.vector(assays(fc_se_b[ipeaks_use, iCol])[[signal_name]]),
            minimum=1,
            newValue=NA);
         iy <- log2(1 + rowData(fc_se[ipeaks_use, ])$Length);
         plotSmoothScatter(
            x=ix,
            y=iy,
            xlab="log2(1 + x) raw counts",
            ylab="Peak width",
            main="",
            yaxt="n");
         # place title atop the plot, using plot panel width
         jamba::drawLabels(preset="top",
            adjPreset="top",
            txt=iCol,
            panelWidth="minimum",
            boxColor=color_sub1[iCol],
            boxBorderColor=jamba::makeColorDarker(color_sub1[iCol]),
            labelCol=jamba::setTextContrastColor(color_sub1[iCol]),
            labelCex=1.3,
            drawBox=TRUE);
         minorLogTicksAxis(2,
            logBase=2,
            offset=1,
            displayBase=10);
         abline(v=mgm_cutoff,
            lty="dotted",
            col="#333333");
         jamba::drawLabels(preset="topleft",
            adjPreset="bottomright",
            txt=Run[iCol],
            boxColor=run_colors1[iCol],
            boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
            labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
            labelCex=1.0,
            drawBox=TRUE);
         total_points <- sum(!is.na(ix) & !is.na(iy));
         jamba::drawLabels(preset="bottomright",
            txt=paste0(
               formatInt(total_points),
               " pts"),
            boxColor="#FFFFFF77",
            boxBorderColor="#99999977",
            labelCol="black",
            labelCex=1,
            drawBox=TRUE);
      }
      par(mfrow=c(1, 1));
      # end plot
   }
}
```


```{r, peak_width_signal_norm2, fig.height=ma_height*1.5, fig.width=ma_width, dependson=c("import_fc", "ma_layout")}
if (FALSE) {
total_n <- length(iSamples) + length(blank_pos);
ma_nrow <- ceiling(total_n / ma_ncol);
par(mfrow=c(ma_nrow, ma_ncol));
par(mar=c(5, 4, 3, 1) + 0.1);
panel_num <- 0;
for (iCol in iSamples) {
   panel_num <- panel_num + 1;
   if (panel_num %in% blank_pos) {
      next;
   }
   iRows <- rownames(iMatrix);
   ix <- noiseFloor(iMatrixQB[, iCol],
         minimum=1,
         newValue=NA);
   iy <- rowData(fc_se[iRows, ])$Length;
   plotSmoothScatter(
      x=ix,
      y=iy,
      xlab="log2(1 + x) raw counts",
      ylab="Peak width",
      main="");
   # place title atop the plot, using plot panel width
   jamba::drawLabels(preset="top",
      adjPreset="top",
      txt=iCol,
      panelWidth="minimum",
      boxColor=color_sub1[iCol],
      boxBorderColor=jamba::makeColorDarker(color_sub1[iCol]),
      labelCol=jamba::setTextContrastColor(color_sub1[iCol]),
      labelCex=1.3,
      drawBox=TRUE);
   abline(v=MGM,
      lty="dotted",
      col="#333333");
   jamba::drawLabels(preset="topleft",
      adjPreset="bottomright",
      txt=Run[iCol],
      boxColor=run_colors1[iCol],
      boxBorderColor=jamba::makeColorDarker(run_colors1[iCol]),
      labelCol=jamba::setTextContrastColor(run_colors1[iCol]),
      labelCex=1.0,
      drawBox=TRUE);
   total_points <- sum(!is.na(ix) & !is.na(iy));
   total_points_mgm <- sum(!is.na(ix) & !is.na(iy) & ix >= MGM);
   jamba::drawLabels(preset="bottomright",
      txt=paste0(
         formatInt(total_points),
         " pts\n",
         formatInt(total_points_mgm),
         " pts MGM", MGM),
      boxColor="#FFFFFF77",
      boxBorderColor="#99999977",
      labelCol="black",
      labelCex=1,
      drawBox=TRUE);
}
par(mfrow=c(1, 1));
}
```

### Statistical Comparisons

#### Confirm Replicates

If there is only one replicate per sample group, then
the collection of samples are used to estimate
variability during statistical testing. This procedure
uses the moderated t-test from the `limma` R package,
which uses an improved estimation of variability
by including pooled variability from neighboring
measurements.


```{r, check_reps}
## check for any available contrasts
if (length(contrastNames) == 0) {
   printDebug("There were no statistical contrasts defined, exiting now.");
   knitr::knit_exit(fully=TRUE);
}

##
ctGroup <- table(iGroup[iSamplesUse]);
denovo_controls <<- FALSE;
if (any(ctGroup > 1)) {
   Outcome <- "At least one sample group contains replicates.";
} else {
   Outcome <- "No sample groups contain replicates. Pooled replicates will be created.";
   denovo_controls <<- TRUE;
   fc_se_ctrl <- fc_se_b[,iSamplesUse];
   colData(fc_se_ctrl)[,group_colname] <- "DenovoControl";
   colnames(fc_se_ctrl) <- jamba::makeNames(rep("DenovoControl", ncol(fc_se_ctrl)));
   colData(fc_se_ctrl)[,"Sample"] <- colnames(fc_se_ctrl);
   fc_se_b <- cbind(fc_se_ctrl, fc_se_b);
   iSamplesUse <- colnames(fc_se_b);
   
   ## update design and contrasts
   sedesign <- jamses::groups_to_sedesign(
      data.frame(check.names=FALSE,
         colData(fc_se[, iSamplesUse])[, group_colname, drop=FALSE]),
      factor_order=FACTOR_ORDER,
      max_depth=1);
   iDesign <- sedesign@design;
   if (length(contrastNames) > 0) {
      iContrasts <- tryCatch({limma::makeContrasts(
            contrasts=contrastNames,
            levels=iDesign);
      }, error=function(e){
         jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
            contrast_txt_file);
         print(e);
         sedesign@contrasts;
      })
   } else {
      iContrasts <- sedesign@contrasts;
   }
}
kable_coloring(
   data.frame(Outcome=Outcome),
   row.names=FALSE,
   caption="Samples were evaluated for presence of biological replicates.")
```


#### Design Matrices

```{r, stats_design, dependson=c("batch_adjust_run", "sample_outliers", "check_reps")}
###############################################
# Statistical comparisons

## subset by iSamplesUse
sedesign <- jamses::validate_sedesign(sedesign,
   min_reps=1,
   samples=iSamplesUse,
   verbose=FALSE);
iDesign <- sedesign@design;
if (length(contrastNames) > 0) {
   iContrasts <- tryCatch({limma::makeContrasts(
         contrasts=contrastNames,
         levels=iDesign);
   }, error=function(e){
      jamba::printDebug("Error during limma::makeContrasts() from contrasts file: ",
         contrast_txt_file);
      print(e);
      sedesign@contrasts;
   });
   sedesign@contrasts <- iContrasts;
} else {
   iContrasts <- sedesign@contrasts;
}

## Optionally remove denovo controls from contrasts if needed
if (denovo_controls) {
   iRemove <- igrep("DenovoControl-|-DenovoControl", colnames(iContrasts));
   if (length(iRemove) > 0) {
      iContrasts <- iContrasts[, -iRemove, drop=FALSE];
      sedesign@contrasts <- iContrasts;
   }
}

## Print design matrix
color_sub_im <- c(`-1`="#2656CC",
   `0`="#DDDDDD",
   `1`="#CC2626");
kable_coloring(
   data.frame(check.names=FALSE,
      iDesign),
   colorSub=color_sub_im,
   caption="Group design matrix") %>%
   kableExtra::column_spec(column=2:(ncol(iDesign) + 1),
      border_left="1px solid #DDDDDD")
## Print contrast matrix
kable_coloring(
   data.frame(check.names=FALSE,
      iContrasts),
   colorSub=color_sub_im,
   caption="Statistical contrast matrix") %>%
   kableExtra::column_spec(column=2:(ncol(iContrasts) + 1),
      border_left="1px solid #DDDDDD")
```


#### Limma Contrasts {.tabset}

```{r, limma_contrasts, dependson=c("stats_design"), results='asis'}

ipeaks_names <- c(
   # list(`MGM 0`=rownames(fc_se)), # do not include all peaks by default
   iPeaks_list)
# do not include raw data by default
signal_names <- head(unique(c(bnormname, normname)), 1)
lba_to_block <- function(x){
   gsub("^(limma_batch_adjust)_(.+)",
      "\\2 (blocking by Run)",
      x)
}
names(signal_names) <- lba_to_block(signal_names)
signal_labels <- nameVector(names(signal_names), signal_names)

## Determine whether we can use Voom
use_voom <- TRUE;
if (all(colSums(design(sedesign), na.rm=TRUE) < 2)) {
   use_voom <- FALSE;
}


## sestats for each MGM threshold
fold_cutoff <- 1.5;
adjp_cutoff <- 0.05;
sestats_list <- lapply(nameVectorN(ipeaks_names), function(ipeaks_name) {
   ipeaks_use <- ipeaks_names[[ipeaks_name]];
   if (length(ipeaks_use) == 0) {
      return(NULL)
   }
   cat(paste0("\n\n##### ", ipeaks_name, " {.tabset}\n\n"));
   mgm_cutoff <- mgm_list[[ipeaks_name]];
   # iterate each signal so that blocking factor can be handled appropriately
   sestats_inner_list <- lapply(signal_names, function(isignal_name){
      isignal_label <- signal_labels[[signal_name]]
      blocking_values <- NULL;
      if (any(grepl("limma_batch_adjust", isignal_name))) {
         # apply blocking factor instead of using batch-adjusted data
         blocking_values <- nameVector(
            colData(fc_se_b[, iSamplesUse])$Run,
            iSamplesUse)
         # since we use blocking factor, do not use batch-adjusted signal
         isignal_name <- gsub("limma_batch_adjust_", "", isignal_name);
      }
      sestats <- tryCatch({
         jamses::se_contrast_stats(
            se=fc_se_b[ipeaks_use, iSamplesUse],
            handle_na="none",
            assay_names=isignal_name,
            p_cutoff=NULL,
            adjp_cutoff=adjp_cutoff,
            fold_cutoff=fold_cutoff,
            mgm_cutoff=mgm_cutoff,
            use_voom=use_voom,
            igenes=ipeaks_use,
            isamples=iSamplesUse,
            sedesign=sedesign,
            block=blocking_values,
            verbose=FALSE);
      }, error=function(e){
         jamba::printDebug("Error during se_contrast_stats():", comment=FALSE);
         print(e);
         NULL;
      });
      # print kable summary
      if (length(sestats) > 0) {
         # remove stats_df to reduce object size
         if ("stats_df" %in% names(sestats)) {
            sestats[["stats_df"]] <- NULL;
         }
         # trim rownames to reduce object size
         if ("stats_dfs" %in% names(sestats)) {
            for (k1 in names(sestats[["stats_dfs"]])) {
               for (k2 in names(sestats[["stats_dfs"]][[k1]])) {
                  # remove rownames to reduce object size
                  rownames(sestats[["stats_dfs"]][[k1]][[k2]]) <- NULL;
               }
            }
         }
         # summarize while using the isignal_label
         hitdf <- sestats_to_df(sestats)
         hitdf$Signal <- isignal_label;
         cat(paste0("\n\n###### ", isignal_label, "\n\n"));
         print(kable_coloring(
            hitdf,
            caption=paste0(
               "Statistical hits per contrast. ",
               "Peak Set: ", ipeaks_name,
               ". Signal: ", isignal_label),
            row.names=FALSE)) 
      }
      sestats
   })
   # return non-empty results
   sestats_inner_list[lengths(sestats_inner_list) > 0];
})
# remove empty entries
# - when no peaks meet threshold
# - some other error during processing
sestats_list <- rmNULL(sestats_list)

```


### Volcano Plots {.tabset}

Base filename: `volcano_plots`

```{r, volcano_plots, fig.height=12, fig.width=12, dependson=c("run_limma"), results='asis'}
options("warn"=-1);
# iterate each sestats
for (sestats_name in names(sestats_list)) {
   cat(paste0("\n\n#### ", sestats_name, " {.tabset}\n\n"));
   mgm_cutoff <- mgm_list[[sestats_name]];
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];
      cat(paste0("\n\n##### ", isignal_label, " {.tabset}\n\n"));
      for (icontrast_name in icontrast_names) {
         icomp_name <- jamses::contrast2comp(icontrast_name)
         cat(paste0("\n\n###### ", icomp_name, "\n\n"));
         # make volcano plot
         # iDF <- sestats_list[[sestats_name]]$stats_dfs[[iSig]][[iCon]];
         par("mar"=c(7, 7, 4, 3));
         submain <- paste0(sestats_name, ", ", isignal_label);
         idf <- sestats_list[[sestats_name]][[isignal_label]][["stats_dfs"]][[
               isignal_name]][[icontrast_name]];
         # print(kable_coloring(head(idf, 10)));
         vp1adjp <- volcano_plot(
            x=sestats_list[[sestats_name]][[isignal_label]][["stats_dfs"]][[
               isignal_name]][[icontrast_name]],
            verbose=FALSE,
            sig_cutoff=0.05,
            fold_cutoff=1.5,
            expr_cutoff=mgm_cutoff,
            main=icomp_name,
            submain=submain,
            sig_colname="adj.P.Val");
         par("mar"=c(5, 4, 4, 2) + 0.1);
      }
   }
}

```


### MGM Distribution {.tabset}

These plots represent the distribution of max group mean values (MGM)
among those peaks which are considered statistical hits using
the adjust P-value and fold change criteria.

Base filename: `mgm_hit_distribution`

```{r, mgm_hit_distribution, fig.height=10, fig.width=10, dependson=c("run_limma"), results='asis'}
# iterate each sestats
for (sestats_name in names(sestats_list)) {
   cat(paste0("\n\n#### ", sestats_name, " {.tabset}\n\n"));
   mgm_cutoff <- mgm_list[[sestats_name]];
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];
   
      cat(paste0("\n\n##### ", isignal_label, " {.tabset}\n\n"));
      for (icontrast_name in icontrast_names) {
         icomp_name <- jamses::contrast2comp(icontrast_name);
         idf <- sestats_list[[sestats_name]][[isignal_label]][["stats_dfs"]][[
            isignal_name]][[icontrast_name]];
         mgmcol <- vigrep("^mgm ", colnames(idf));
         # sigcol <- vigrep("^P.Value ", colnames(idf));
         foldcol <- vigrep("^fold ", colnames(idf));
         # filter by adjusted P-value
         sigcol <- vigrep("^adj.P.Val ", colnames(idf));
         sigadjp <- which(abs(idf[[foldcol]]) >= 1.5 &
            idf[[sigcol]] <= 0.05);
         if (length(sigadjp) > 2) {
            cat(paste0("\n\n###### ", icomp_name, "\n\n"));
            tryCatch({
               jamba::plotPolygonDensity(
                  main=paste0("Max Group Mean for ",
                     formatInt(length(sigadjp)),
                     " statistical hits\n",
                     "with adj.P.Val <= 0.05 and abs(fold) >= 1.5\n",
                     icomp_name
                     # gsub("^mgm ", "", mgmcol)
                     ),
                  xlab=mgmcol,
                  idf[sigadjp, mgmcol])
               abline(v=mgm_cutoff, col="black", lwd=2);
               abline(v=mgm_cutoff, col="gold", lty=2, lwd=2);
            }, error=function(e){
               jamba::printDebug("plot skipped, signal: ",
                  isignal_name,
                  ", contrast: ",
                  icontrast_name,
                  comment=FALSE,
                  ", error:");
               print(e);
            })
         }

      }
   }
}

```



### Venn diagrams {.tabset}

Base filename: `venn_diagrams`

```{r, venn_diagrams, fig.height=10, fig.width=10, results='asis'}
suppressMessages(require(venndir));
# iterate each sestats
for (sestats_name in names(sestats_list)) {
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];
      contrast_colors <- group2colors(icontrast_names);
   
      if (length(icontrast_names) > 1) {
         cat(paste0("\n\n#### ", sestats_name, " {.tabset}\n\n"));
      }
   
      set_lists <- jamses::hit_array_to_list(
         sestats_list[[sestats_name]][[isignal_label]]$hit_array,
         assay_names=isignal_name,
         cutoff_names=icutoff_names)
      icontrast_names <- names(set_lists)[lengths(set_lists) > 0];
      
      if (length(icontrast_names) > 1) {
         cat(paste0("\n\n##### ", isignal_label, " {.tabset}\n\n"));
         # For now: use all non-empty pairs of contrasts
         # TODO: determine appropriate Venn diagrams
         pair_l <- unlist(recursive=FALSE,
            apply(combn(icontrast_names, 2), 2, list));
         
         # iterate Venn pairs
         for (venn_contrasts in pair_l) {
            venn_comps <- jamses::contrast2comp(venn_contrasts);
            venn_label <- jamba::cPaste(venn_comps, sep=", ")
            cat(paste0("\n\n###### ", venn_label, "\n\n"));
            tryCatch({
               setlist <- set_lists[venn_contrasts];
               names(setlist) <- gsub(":",
                  "<br>\n",
                  jamses::contrast2comp(names(setlist)))
               venn_title <- paste0(
                  "Signal:", isignal_label,
                  "\nPeaks: ", sestats_name,
                  "\nThreshold: ", cPaste(icutoff_names))
               venndir::venndir(setlist,
                  overlap_type="each",
                  offset=c(0, -1),
                  show_segments=FALSE,
                  label_preset="main outside",
                  set_colors=contrast_colors[venn_contrasts],
                  main=venn_title);
            }, error=function(e){
               jamba::printDebug("venndir::venndir() ",
                  comment=FALSE,
                  "skipping plot for signal:", isignal_name,
                  ", icutoff_names:", icutoff_names,
                  ", contrasts:", venn_contrasts,
                  "\nError (below):\n");
               print(e);
            });
            cat("\n\n");
         }
      }
   }
}

```


### Annotate Peaks

The GTF file was used to define genomic regions. Each peak is annotated
with only one of the following categories, assigned in order.
Therefore a peak that overlaps a promoter and an exon is only
annotated "promoter".

* **promoters**: `r upstream_promoter` upstream, and `r downstream_promoter`
downstream each transcript TSS
* **TTS**: `r upstream_tts` upstream, and `r downstream_tts` downstream
the last stranded nucleotide position in each transcript.
* **exon**: each stranded exon per transcript
* **intron**: non-exonic region within a stranded transcript region
* **extragenic**: all other regions.


#### Define Genome Regions

This section converts the given GTF file to a set of genome regions.

The GTF input file: **`r GTF`**

`r jamba::printDebug(unlist(strsplit(GTF, "/")), collapse="/", sep="", comment=FALSE)`


A cached RData file is stored for future re-use, using this file extension:

> `.genome_regions.RData`


```{r, define_genome_regions}
genome_regions <- genomic_regions_from_gtf(gtf=GTF,
   upstream_promoter=upstream_promoter,
   downstream_promoter=downstream_promoter,
   upstream_tts=upstream_tts,
   downstream_tts=downstream_tts,
   detectedTx=detectedTx,
   detectedGenes=detectedGenes,
   save_rdata=TRUE,
   save_txdb=TRUE);
```


#### Annotate Genome Regions

```{r, peak_genes_fn}
## Annotate ATAC peaks
fc_gr <- as(rowData(fc_se), "GRanges");
values(fc_gr) <- renameColumn(values(fc_gr),
   from="Geneid",
   to="Gene");

# Call function that wraps the logic together
fc_gr <- annotate_gr_by_genome_region(
   gr=fc_gr,
   genome_regions=genome_regions,
   mask_regions=MASK,
   verbose=FALSE);

# now propagate annotations into the fc_se and fc_se_b objects
region_colnames <- provigrep(c(
   "gene_(name|id|feature)|feature_type|nearest_gene",
   "broad_peak|num_slices"),
   colnames(values(fc_gr)))
gr_match <- match(rownames(fc_se), values(fc_gr)$Gene);
if (!any(is.na(gr_match))) {
   rowData(fc_se)[,region_colnames] <- data.frame(check.names=FALSE,
      values(fc_gr)[gr_match, region_colnames, drop=FALSE])
}
gr_match <- match(rownames(fc_se_b), values(fc_gr)$Gene);
if (!any(is.na(gr_match))) {
   rowData(fc_se_b)[,region_colnames] <- data.frame(check.names=FALSE,
      values(fc_gr)[gr_match, region_colnames, drop=FALSE])
}

feature_type_colors <- colorjam::group2colors(unique(
   as.character(values(fc_se)$feature_type_winner)));
sample_color_list$feature_type_winner <- feature_type_colors;

# show the full set of colors
# showColors(sample_color_list)

```


#### Incidence Matrix by feature_type

The plot below shows the association of each peak with one or more
`feature_type` values.

Along the diagonal, it represents a count of overlaps for all
peaks over each `feature_type` value.

The cells off the diagonal indicate the number of peaks
that overlap two `feature_type` values. For example, `"Promoter"`
often also overlaps `"Exon"` particularly when the peak covers
the TSS itself. Similarly, `"Exon"` often overlaps `"Intron"`
partly because the average exon is only 100 bases wide, which
is usually smaller than most peaks.

```{r, peak_featuretype_im}
## For fun make an logical matrix of feature_type
ftcols <- c("Promoter", "TTS", "exon", "intron", "extragenic")
fc_typem <- as.data.frame(lapply(nameVector(ftcols), function(i){
   j <- grepl(i, values(fc_gr)$feature_type);
   names(j) <- values(fc_gr)$Gene;
   j;
}));
m1 <- as.matrix(fc_typem);
m1_counts <- (t(m1) %*% (m1));
m1_df <- as.data.frame(m1_counts);
for (i1 in colnames(m1_df)) {
   m1_df[,i1] <- formatInt(m1_df[,i1])
}
m1_values <- sort(unique(as.vector(m1_counts)));
names(m1_values) <- formatInt(m1_values)
color_sub_imct <- colorjam::vals2colorLevels(
   m1_values,
   lens=10);
kable_coloring(as.data.frame(m1_counts),
   colorSub=color_sub_imct,
   color_cells=TRUE,
   format.args=list(big.mark = ",",
      scientific=FALSE),
   row.names=TRUE) %>%
   kable_styling(bootstrap_options=c("striped","bordered")) %>%
   row_spec(0, background="#DDDDDD", color="#333333", bold=TRUE) %>%
   column_spec(1, background="#DDDDDD", color="#333333", bold=TRUE)

```


### Hits by Region

#### Pie Charts {.tabset}

A pie chart is created for each statistical comparison, showing
the number of peaks overlapping each genome region. A peak
is only assigned one genome region, with preference in order:
promoter, TTS, exon, intron, extragenic.

For comparison, the pie chart using all peaks is also shown.

Base filename: `peak_region_piecharts`

```{r peak_dimensions, results='asis'}
icontrast_names <- jamses::contrast_names(sedesign)
gg2_num <- length(icontrast_names)
gg2_width <- 10 + 7 * (gg2_num > 1);
gg2_height <- ceiling(gg2_num / 2) * 7;
```

```{r, peak_region_piecharts, fig.height=gg2_height, fig.width=gg2_width, results='asis'}

feature_type_winner <- values(fc_gr)$feature_type_winner;
names(feature_type_winner) <- values(fc_gr)$Gene;

# iterate
for (sestats_name in names(sestats_list)) {
   mgm_cutoff <- mgm_list[[sestats_name]];
   cat(paste0("\n\n##### ", sestats_name, " {.tabset}\n\n"));
   # isignal_names <- names(sestats_list[[sestats_name]]);
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];

      cat(paste0("\n\n###### ", isignal_label, "\n\n"));
      if (max(lengths(sestats_list[[sestats_name]][[isignal_label]]$hit_array)) > 0) {
         # iterate each set of pie charts
         gg2 <- gg_pie_by_feature_type(
            hit_array=sestats_list[[sestats_name]][[isignal_label]][["hit_array"]],
            iSig=isignal_name,
            colorSub=sample_color_list$feature_type_winner,
            add_peak_list=iPeaks_list[sestats_name],
            do_pie=TRUE,
            feature_type_winner=feature_type_winner);
         print(gg2);
         cat("\n");
      }
   }
}

```


#### Table Summary {.tabset}

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapsePie1" role="button" aria-expanded="false" aria-controls="collapsePie1">(Table summary...)</a></p>
<div class="collapse" id="collapsePie1"><div class="card card-body">

```{r, peak_region_table, fig.height=gg2_height, fig.width=gg2_width, results='asis'}

feature_type_winner <- values(fc_gr)$feature_type_winner;
names(feature_type_winner) <- values(fc_gr)$Gene;

# iterate
for (sestats_name in names(sestats_list)) {
   mgm_cutoff <- mgm_list[[sestats_name]];
   cat(paste0("\n\n##### ", sestats_name, " {.tabset}\n\n"));
   # isignal_names <- names(sestats_list[[sestats_name]]);
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];
      
      cat(paste0("\n\n###### ", isignal_label, "\n\n"));
      if (max(lengths(sestats_list[[sestats_name]][[isignal_label]][["hit_array"]])) > 0) {
         # iterate each set of pie charts
         jdf <- gg_pie_by_feature_type(
            hit_array=sestats_list[[sestats_name]][[isignal_label]][["hit_array"]],
            iSig=isignal_name,
            colorSub=sample_color_list$feature_type_winner,
            add_peak_list=iPeaks_list[sestats_name],
            do_pie=FALSE,
            feature_type_winner=feature_type_winner);
         print(
            kable_coloring(jdf[, 1:10],
               row.names=FALSE,
               colorSub=sample_color_list$feature_type_winner))
         cat("\n");
      }
   }
}

```


```{r, peak_pies_gg_all, fig.height=gg2_height, fig.width=gg2_width}
if (FALSE) {
if (length(gg2) > 0) {
   print(gg2);
   
   # print table format
   kable_coloring(jdf[, 1:10, drop=FALSE],
      row.names=FALSE,
      colorSub=sample_color_list$feature_type_winner,
      full_width=TRUE,
      caption="Data used for the pie charts above:") %>%
      kable_styling(bootstrap_options=c("striped","bordered")) %>%
      kableExtra::column_spec(column=seq_len(10),
         extra_css="white-space: nowrap;") %>%
      row_spec(0, background="#DDDDDD");

}
}
```

</div></div>


### Broad Peaks

Peaks are re-numbered so that contiguous peaks are numbered
in column `"broak_peak"` to indicate when one broad peak
has been sliced by `mergeSplitCountBed` into multiple pieces.
Each peak slice is labeled in `"broak_peak_number"` with
a suffix to indicate the slice number.

This process makes it feasible to count the number of contiguous enriched
regions, and to evaluate the statistical hits across different slices
in the same region.

For example, in one `"broad_peak"`, one slice may have enriched signal,
while another slice may be reduced. The total signal in the `"broad_peak"`
may remain constant even while there is significant change in the
detected chromatin signal.


```{r, peak_renumbering}
fc_gr_red <- GenomicRanges::reduce(fc_gr);
values(fc_gr_red)$broad_peak <- makeNames(
   rep("broad", length(fc_gr_red)),
   doPadInteger=TRUE,
   renameOnes=TRUE,
   suffix="_");
fc_gr <- annotateGRfromGR(fc_gr,
   fc_gr_red);
values(fc_gr)$broad_peak_number <- makeNames(
   values(fc_gr)$broad_peak,
   suffix="_",
   doPadInteger=TRUE,
   renameOnes=TRUE);
fc_gr_bp_ct <- table(values(fc_gr)$broad_peak);
values(fc_gr)$num_slices <- as.vector(fc_gr_bp_ct[values(fc_gr)$broad_peak]);
rowData(fc_se)$broad_peak <- values(fc_gr)$broad_peak;
rowData(fc_se)$broad_peak_number <- values(fc_gr)$broad_peak_number;
rowData(fc_se)$broad_peak_slices <- fc_gr_bp_ct[values(fc_gr)$broad_peak];
```


#### Statistical Hits {.tabset}

Statistical hits are tabulated by unique `"broad_peak"` values.

There are two potential outcomes of this process:

1. **Significant broad_peak hits are nearly identical to individual peak hits.**

   * This result suggests most peaks are not sliced, or that
   significantly changing peaks are not regions broader than the slice width.

2. **Significant broad_peak hits are much smaller than individual peak hits.**

   * This result suggests most peaks are sliced, or that
   significantly changing peaks tend to be regions broader than
   the slice width.


```{r, broad_hit_counts, results='asis'}

broad_hit_counts <- list();
for (sestats_name in names(sestats_list)) {
   mgm_cutoff <- mgm_list[[sestats_name]];
   cat(paste0("\n\n##### ", sestats_name, " {.tabset}\n\n"));
   # isignal_names <- names(sestats_list[[sestats_name]]);
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];

      if (max(lengths(sestats_list[[sestats_name]][[isignal_label]][["hit_array"]])) > 0) {
         cat(paste0("\n\n###### ", isignal_label, "\n\n"));
         # iterate cutoffs
         for (icutoff_name in icutoff_names) {
            # iterate contrasts
            idf <- rbindList(
               lapply(nameVector(icontrast_names), function(icontrast_name){
                  k <- sestats_list[[sestats_name]][[isignal_name]][["hit_array"]][
                     icutoff_name, icontrast_name, isignal_name][[1]];
                  k_bp <- values(fc_gr[names(k)])$broad_peak;
                  icomp_name <- jamses::contrast2comp(icontrast_name)
                  data.frame(check.names=FALSE,
                     `contrast name`=icontrast_name,
                     `comp name`=icomp_name,
                     `total peak hits`=length(k_bp),
                     `unique broad peak hits`=length(unique(k_bp)),
                     `signal label`=isignal_label,
                     `signal name`=isignal_name,
                     `cutoff`=icutoff_name
                  );
               }));
            broad_hit_counts[[sestats_name]][[isignal_label]][[
               icutoff_name]] <- idf;
            print(kable_coloring(idf));
            cat("\n\n");
         }
         # jamba::printDebug("broad_hit_counts:");
         # broad_hit_counts;
      }
   }
}
# data is stored in a nested list broad_hit_counts
# - named by sestats_names, signal_labels, signal_name, cutoff_names

# Consider TODO:
# - Venn diagram comparing hits across sestats_names, signal_labels,
#    - for each of the same contrast_names
#    - for each of the signal_labels

```

### Stats Heatmaps {.tabset}

```{r stat_heatmaps, fig.height=14, fig.width=4 + sqrt(length(iSamplesUse)) * 1.5, results='asis'}

# centering and controls
centerby_list <- list(
   `centered by first group`=NULL,
   `global-centered`=NULL,
   `centered by Run`="Run")
# control samples in the first group
iControl <- head(levels(iGroup), 1);
iControlSamples <- iSamplesUse[
   colData(fc_se[, iSamplesUse])[[group_colname]] %in% iControl]
centering_controls <- list(
   `centered by first group`=iControlSamples,
   `global-centered`=iSamplesUse,
   `centered by Run`=iSamplesUse)
centering_control_labels <- list(
   `centered by first group`=paste0("versus ",
      as.character(iControl), ""),
   `global-centered`="",
   `centered by Run`="")

# iterate each MGM threshold used in sestats_list
# - then at each signal_label
for (sestats_name in names(sestats_list)) {
   cat(paste0("\n\n#### ", sestats_name, " {.tabset}\n\n"));
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[isignal_label]]$hit_array);
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];

      # Only create heatmap when there are statistical hits.
      # - consider hiding contrasts that have no hits from the heatmap
      #   sidebar, as created by argument sestats. We can do that by
      #   supplying a subset of contrast_names with length > 0.
      # - Consider subsetting contrasts shown by the factor being compared?
      #   This option is more interesting with multi-factor designs.
      hit_lengths <- lengths(sestats_list[[sestats_name]][[isignal_label]][[
         "hit_array"]][, , isignal_name, drop=FALSE])
      if (max(hit_lengths) > 0) {
         #
         cat(paste0("\n\n##### Signal: ", isignal_label, "  {.tabset}\n\n"));
         icontrast_names <- names(hit_lengths)[hit_lengths > 0];
         for (centering_name in names(centerby_list)) {
            cat(paste0("\n\n###### ", centering_name, "\n\n"));
            # determine centering
            icenterby_colnames <- centerby_list[[centering_name]];
            iControlSamples <- centering_controls[[centering_name]];
            # optionally hide dendrogram when too many rows
            show_row_dend <- TRUE;
            temp_hit_im <- venndir::list2im_value(
               sestats_list[[sestats_name]][[isignal_label]]$hit_array[, ,
                  isignal_name]);
            cluster_rows <- TRUE;
            row_km <- 1;
            row_split <- "feature_type_winner";
            if (nrow(temp_hit_im) > 5000) {
               show_row_dend <- FALSE
               cluster_rows <- FALSE;
               row_km <- 12;
               row_split <- NULL;
            }
            # generate heatmap
            hm_sestats <- heatmap_se(
               fc_se_b[, iSamplesUse],
               sestats=sestats_list[[sestats_name]][[isignal_label]],
               assay_name=isignal_name,
               correlation=FALSE,
               top_colnames=c("Run", "Group"),
               centerby_colnames=icenterby_colnames,
               controlSamples=iControlSamples,
               control_label=centering_control_labels[[centering_name]],
               contrast_names=icontrast_names,
               data_type="signal",
               row_type="significant peaks",
               color_max=3,
               sample_color_list=sample_color_list,
               row_split=row_split,
               row_km=row_km,
               cluster_rows=cluster_rows,
               show_row_dend=show_row_dend,
               show_row_names=FALSE,
               column_cex=0.4,
               column_split=c("Group"),
               column_title_rot=90)
            ComplexHeatmap::draw(hm_sestats,
               column_title=attr(hm_sestats, "hm_title"),
               merge_legends=TRUE)
         }
      }
   }
}

```


### ATAC Mode

**`r if (ATAC) {("ATAC-mode is enabled.")} else {("ATAC-mode is disabled.")}`**

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseAtac1" role="button" aria-expanded="false" aria-controls="collapseAtac1">(Expand...)</a></p>
<div class="collapse" id="collapseAtac1"><div class="card card-body">

This workflow is enabled only when environment variable `ATAC=1`,
and when stats have been performed, and when the max group mean
filter has been applied.

This workflow defines a subset of genes with "active promoters" if
any ATAC-seq peaks are found in the `"Promoter"` region of the gene.
For this purpose, we require the peak to meet the MGM threshold
for detection above background.

The purpose of this step is to reduce the total genes defined in the GTF
to the subset of genes that were observed to have "active promoter",
so the subset of GTF genes can be used to annotate peaks.
The result is improved accuracy in assigning genes to peaks,
by removing GTF entries that are not associated with active
transcription in the given experiment.

In practice, most actively transcribed genes, or genes that are
poised to become actively transcribed, have open chromatin
seen by ATAC-seq across most or all conditions of an experiment.

Peaks are annotated to the nearese MGM-filtered gene TSS, as
defined by the promoter regions, along with the distance to that region.

</div></div>

```{r, peak_promoters}
mgmpg <- NULL;
mgmpd <- NULL;
if (ATAC) {
   
   ## iPeaksMgm that are promoters
   ## - call these active promoters
   ##
   ## annotate all peaks by nearest active promoter
   ## - promoter,tts,intron,exon -- gene_name intersected
   ## - all cases, nearest mgm4 promoter, and its gene_name
   ## - distance to nearest mgm4 promoter
   
   # subset to promoters
   promoters_gr <- subset(genome_regions,
      grepl("promoter", 
         ignore.case=TRUE, 
         feature_type));

   upstream <- upstream_promoter;
   downstream <- downstream_promoter;
   
   # Iterate each MGM threshold
   for (ipeaks_name in names(iPeaks_list)) {
      tryCatch({
         ipeaks_use <- iPeaks_list[[ipeaks_name]];
         mgm_cutoff <- mgm_list[[ipeaks_name]];
         mgm_promoter_gr <- IRanges::subsetByOverlaps(promoters_gr,
            fc_gr[ipeaks_use],
            ignore.strand=TRUE);
         mgm_tss_gr <- promoter_to_tss(mgm_promoter_gr[,c("gene_name", "gene_id")],
            upstream=upstream,
            downstream=downstream);
         
         mgmpg <- paste0("mgm", mgm_cutoff, "_annoTSS_gene_name");
         mgmpgid <- paste0("mgm", mgm_cutoff, "_annoTSS_gene_id");
         mgmpd <- paste0("mgm", mgm_cutoff, "_annoTSS_distance");
         
         # annotate every peak to nearest TSS with ATAC peak above mgm_cutoff
         # nearest mgm_annoTSS distance
         fc_gr_dist <- GenomicRanges::distanceToNearest(fc_gr,
            ignore.strand=TRUE,
            select="all",
            mgm_tss_gr);
         
         # Add comma-delimited values per column as needed
         genecol1 <- head(provigrep(c("^Geneid", "^Gene"),
            colnames(values(fc_gr))), 1);
         fc_gr_dist_df <- renameColumn(
            data.frame(
               as.data.frame(values(
                  fc_gr[queryHits(fc_gr_dist)])[, genecol1, drop=FALSE]),
               gene_distance=values(fc_gr_dist)$distance,
               as.data.frame(values(
                  mgm_tss_gr[subjectHits(fc_gr_dist)])[,
                     c("gene_name", "gene_id"), drop=FALSE])
            ),
            from=c("gene_name", "gene_id", "gene_distance"),
            to=c(mgmpg, mgmpgid, mgmpd)
         )
         for (i in c(mgmpg, mgmpgid, mgmpd)) {
            if (mgmpg %in% colnames(fc_gr_dist_df)) {
               fc_gr_dist_df1 <- unique(fc_gr_dist_df[,
                  c(genecol1, i), drop=FALSE]);
               if (igrepHas("_distance", i)) {
                  fc_gr_dist_df1vals <- nameVector(fc_gr_dist_df1[,
                     c(i, genecol1), drop=FALSE]);
                  values(fc_gr)[, i] <- c(NA, 0)[1];
               } else {
                  fc_gr_dist_df1vals <- cPaste(split(fc_gr_dist_df1[[i]],
                     fc_gr_dist_df1[[genecol1]]))
                  values(fc_gr)[, i] <- head(c(NA, ""), 1);
               }
               imatch <- match(names(fc_gr_dist_df1vals),
                  values(fc_gr)[[genecol1]]);
               if (length(imatch) > 0) {
                  if (igrepHas("_distance", i)) {
                     values(fc_gr)[imatch, i] <- fc_gr_dist_df1vals;
                  } else {
                     values(fc_gr)[imatch, i] <- as.character(
                        fc_gr_dist_df1vals);
                  }
               }
            }
         }
      }, error=function(e){
         printDebug("Error during ATAC peak TSS annotation:",
            fgText="red");
         print(e);
      })
   }
} else {
   ATAC <- FALSE;
}
```

### Annotate Gene Name

This section uses each `_gene_id` column to
retrieve the Entrez gene description, using the R package
`"genejam"` if available. The resulting column has suffix
`_gene_description`.

```{r, add_gene_names, results='asis'}
if (suppressPackageStartupMessages(require(genejam))) {
   if (igrepHas("gene_id", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_id", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genenamecol <- gsub("gene_id", "gene_name", genecol);
         genedescol <- gsub("gene_id", "gene_description", genecol);
         # jamba::printDebug(c("Converting from '", genecol,
         #   "' to '", genedescol, "'"), sep="",
         # comment=FALSE);
         genedf0 <- genejam::freshenGenes(
            data.frame(intermediate=values(fc_gr)[[genecol]],
               stringsAsFactors=FALSE),
            try_list=NULL,
            sep=";",
            intermediate="intermediate",
            final=c("SYMBOL", "GENENAME"));
         genedf3 <- trim_multigenedesc(genedf0,
            max_genes=4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            # jamba::printDebug(
            #    "   No gene descriptions were found for the input data.",
            #    comment=FALSE,
            #    fgText="red2");
         }
      }
   } else if (igrepHas("gene_name", colnames(values(fc_gr)))) {
      genecols <- head(rev(vigrep("gene_name", colnames(values(fc_gr)))), 1);
      for (genecol in genecols) {
         genedescol <- gsub("gene_name", "gene_description", genecol);
         # jamba::printDebug(c("Converting from '", genecol,
         #    "' to '", genedescol, "'"), sep="",
         #    comment=FALSE);
         genedf2 <- genejam::freshenGenes(
            data.frame(gene_id=values(fc_gr)[[genecol]],
               stringsAsFactors=FALSE),
            try_list=c("SYMBOL2EG", "ALIAS2EG"),
            sep=";",
            intermediate="intermediate",
            final=c("SYMBOL", "GENENAME"))
         genedf3 <- trim_multigenedesc(
            genedf2[,c("intermediate", "SYMBOL", "GENENAME"), drop=FALSE],
            max_genes=4);
         if (any(nchar(genedf3$GENENAME) > 0)) {
            values(fc_gr)[[genedescol]] <- genedf3$GENENAME;
         } else {
            # jamba::printDebug(
            #    "   No gene descriptions were found for the input data.",
            #    comment=FALSE,
            #    fgText="red2");
         }
      }
   }
}
```


### Summary Table

```{r summary_table}
idf <- head(as.data.frame(fc_gr), 20);
kable_coloring(
   idf,
   row.names=FALSE,
   colorSub=sample_color_list)

```


### Export stats files

Statistical results are saved as text files and BED files.

> **`r OUTDIR`**


```{r, export_stats, results='asis'}
#
# iterate each sestats
for (sestats_name in names(sestats_list)) {
   sestats_label <- tolower(gsub("[ ]*>=[ ]*", "", sestats_name));
   mgm_cutoff <- mgm_list[[sestats_name]];
   isignal_labels <- names(sestats_list[[sestats_name]]);
   for (isignal_label in isignal_labels) {
      sestats_dimnames <- dimnames(sestats_list[[sestats_name]][[
         isignal_label]]$hit_array);
      isignal_file_label <- gsub("^_|_$", "",
         gsub("[ ()]+", "_",
            isignal_label))
      isignal_name <- sestats_dimnames[[3]];
      icutoff_names <- sestats_dimnames[[1]];
      icontrast_names <- sestats_dimnames[[2]];

      # create overall data.frame
      stats_df <- tryCatch({
         jamba::mergeAllXY(
            sestats_list[[sestats_name]][[
               isignal_label]]$stats_dfs[[isignal_name]])
      }, error=function(e){
         printDebug(comment=FALSE,
            "Error during merge of stat data.frame objects",
            sestats_name,
            isignal_label,
            isignal_name);
         print(e);
         NULL
      })
      if (length(stats_df) > 0) {
         ## Add nearest promoter df
         pro_cols1 <- provigrep(c("^gene_id",
            "^gene_name",
            "feature_type($|_winner)",
            "_gene",
            "_distance",
            "broad_peak|num_slices",
            "^mask$"),
            colnames(values(fc_gr)));
         pro_cols <- intersect(
            colnames(values(fc_gr)),
            pro_cols1
         );
         if (length(pro_cols) > 0) {
            idf_match <- match(stats_df[, 1], values(fc_gr)$Gene);
            if (any(!is.na(idf_match))) {
               promoter_df <- data.frame(check.names=FALSE,
                  stringsAsFactors=FALSE,
                  values(fc_gr[idf_match, pro_cols]));
               stats_df[, pro_cols] <- promoter_df;
               if ("mask" %in% pro_cols) {
                  # convert TRUE/FALSE to 1/0
                  stats_df[, "mask"] <- as.numeric(stats_df[, "mask"]);
               }
            }
         }
         ## Add genome coordinate annotations
         coord_cols <- c("Chr", "Start", "End", "Length");
         idf_match <- match(stats_df[,1], rowData(fc_se)$Geneid);
         if (any(!is.na(idf_match))) {
            stats_df <- data.frame(check.names=FALSE,
               stringsAsFactors=FALSE,
               stats_df[, 1, drop=FALSE],
               data.frame(check.names=FALSE,
                  stringsAsFactors=FALSE,
                  rowData(fc_se[idf_match, ])[, coord_cols, drop=FALSE]),
               stats_df[, -1, drop=FALSE]);
         }
         # define output filename
         txt_filename <- paste0("slicejam_stats.",
            sestats_label,
            ifelse(ATAC, "_atac1", ""),
            ".", isignal_file_label, ".txt");
         txt_file <- file.path(OUTDIR,
            txt_filename)
         # save to file
         data.table::fwrite(
            x=stats_df,
            file=txt_file,
            sep="\t",
            quote=FALSE,
            col.names=TRUE);
         kable_coloring(
            data.frame(check.names=FALSE,
               `Rows saved`=formatInt(nrow(stats_df)),
               `Output file`=txt_file),
            row.names=FALSE,
            caption="Filename saved.")
         #
         # Save all peaks in BED format
         # TODO: also export "all" peaks tested
         statsbed <- tryCatch({
            statsdf2bed(stats_df,
               MGM=mgm_cutoff,
               type="hits")
         }, error=function(e){
            NULL
         })
         if (length(statsbed) > 0 && nrow(statsbed) > 0) {
            bed_filename <- paste0("slicejam_stats.",
               sestats_label,
               ifelse(ATAC, "_atac1", ""),
               ".", isignal_file_label, ".bed");
            bed_file <- file.path(OUTDIR,
               bed_filename)
            trackname <- paste0(
               format(Sys.time(), "%Y%m%d%H%M"),
               paste0("mgm", mgm_cutoff));
            trackline <- paste0('track name=',
               trackname,
               ' description="',
               trackname,
               ' peaks"');
            writeLines(text=trackline,
               con=bed_file);
            data.table::fwrite(
               x=statsbed,
               file=bed_file,
               append=TRUE,
               sep="\t",
               quote=FALSE,
               col.names=FALSE);
         }
         #

      }
   }
}

```


## Export RData for follow-up analyses

**`r if (SAVE_RDATA) {("RData was saved.")} else {("RData was not saved.")}`**


```{r, export_rdata, dependson=c("check_reps", "run_limma", "run_limma_mgm")}
if (SAVE_RDATA) {
   rdata_ext <- ".RData";
   fc_dat_file <- file.path(OUTDIR,
      "slicejam_analysis.RData")

   if (!igrepHas("RData", fc_dat_file)) {
      fc_dat_file <- paste0(fc_dat_file, ".RData");
   }
   save.image(fc_dat_file);
   (kable_coloring(
      data.frame(check.names=FALSE,
         `RData file saved:`=fc_dat_file),
      row.names=FALSE));
}
```

## R Session Info

<p><a class="btn btn-primary" data-toggle="collapse" href="#collapseSessionInfo1" role="button" aria-expanded="false" aria-controls="collapseSessionInfo1">(R Session Information)</a></p>
<div class="collapse" id="collapseSessionInfo1"><div class="card card-body">

```{r r_session_info, cache=FALSE}
sessioninfo::session_info()
```

</div></div>
